\chapter{ICNN, Brenier Maps, and Monge-Gap Direct Map Learning}
\label{ch:bf-neural-ot-direct-map}

The previous chapter studied the narrowest neural extension of the entropic OT
teacher: retain the teacher and learn how to start its solver better.  The present
chapter changes the question.  Instead of accelerating a declared solver, it asks
how one can learn the transport map itself.

This is a more ambitious route.  The potential gain is obvious: if a good map is
learned directly, one may bypass repeated iterative solves at deployment time.
The cost is equally important: the learned object is no longer an initialization
for a retained teacher.  It is the transport map itself, so every modeling and
training choice now acts directly on the output cloud.

The chapter is deliberately slower than the earlier neural-OT material.  Its job
is not to survey every map-learning paper quickly.  Its job is to teach the
reader what a direct map-learning route is trying to learn, why Brenier/ICNN and
Monge-gap methods are mathematically natural, and where the new approximation
burden enters.

\section{Reader orientation: what changes in this chapter?}
\label{sec:bf-neural-ot-direct-orientation}

The best way to read this chapter is to compare it with the retained-teacher
warm-start chapter.
\begin{itemize}
  \item In the warm-start chapter, the teacher still decides what the final
  transport plan and output cloud are; the learner only helps the teacher solve
  its own problem faster.
  \item In the present chapter, the learner itself proposes the transport map.
  There may be no corrective Sinkhorn phase at all.
\end{itemize}
So the central question is no longer “did we accelerate the same teacher?”  It is
“what map object have we learned, and under what assumptions should we trust it?”

\paragraph{Plain-language takeaway.}
If the previous chapter was about helping an expensive trusted calculation, this
chapter is about replacing that calculation by a learned map.  That is why the
mathematical structure and the burden of proof both change.

\section{Object inventory for direct map learning}
\label{sec:bf-neural-ot-direct-object}

A direct map-learning method tries to produce a deterministic transport map
\begin{equation}
\label{eq:bf-neural-ot-direct-map-object}
  T:
  \mathbb R^{d_x}	o\mathbb R^{d_x}.
\end{equation}
The source measure \(\mu\) and target measure \(\nu\) are related by the
pushforward relation
\begin{equation}
\label{eq:bf-neural-ot-direct-pushforward}
  T_\#\mu = \nu,
\end{equation}
meaning that if one samples \(X\sim\mu\) and then computes \(T(X)\), the law of
that transformed sample is \(\nu\).

The main local objects are:
\begin{center}
\small
\begin{tabular}{p{0.22\textwidth} p{0.65\textwidth}}
\toprule
Object & Meaning in this chapter \\
\midrule
$\mu,\nu$ & source and target probability measures \\
$T$ & direct transport map sending source points to target points \\
$\psi$ & scalar potential whose gradient may define the transport map \\
$\psi^*$ & convex conjugate of the potential when that representation is used \\
$T_\#\mu$ & pushforward law obtained by applying $T$ to source samples \\
$\mathcal M_\rho^c(T)$ & Monge-gap regularizer measuring how OT-like the map is for cost $c$ and reference measure $\rho$ \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Reader checkpoint.}
The important change relative to the warm-start chapter is immediate: the learned
object is no longer an internal solver coordinate.  It is the transport map
itself, or a potential from which that map is derived.

\section{Running direct-map cell}
\label{sec:bf-neural-ot-direct-running-cell}

A tiny one-dimensional cell is enough to carry the main ideas.  Let the source be
the two-point cloud
\[
  X_{\mathrm{src}}=\{-1,1\}
  \quad\text{with equal weights }\frac{1}{2},
\]
and let the target be
\[
  X_{\mathrm{tgt}}=\{0,2\}
  \quad\text{with equal weights }\frac{1}{2}.
\]
The obvious translation map is
\begin{equation}
\label{eq:bf-neural-ot-direct-shift-map}
  T^\star(x)=x+1.
\end{equation}
Applying \(T^\star\) to the source cloud sends \(-1\mapsto 0\) and
\(1\mapsto 2\), so the pushforward equals the target cloud exactly.

For quadratic cost,
\begin{equation}
\label{eq:bf-neural-ot-direct-shift-cost}
  \frac{1}{2}\sum_{x\in X_{\mathrm{src}}}|x-T^\star(x)|^2
  =
  \frac{1}{2}(1^2+1^2)=1.
\end{equation}
This simple cell already shows the difference between a direct map and a retained
teacher solve.  The output cloud is obtained directly by evaluating \(T\) on the
source particles; there is no intermediate coupling matrix or balancing loop in
this picture.

\section{Brenier map viewpoint}
\label{sec:bf-neural-ot-direct-brenier}

In the quadratic-cost Euclidean setting, a classical route is to represent the map
as the gradient of a convex potential:
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier}
  T^\star(x)=\nabla\psi^\star(x),
  \qquad
  \psi^\star\text{ convex}.
\end{equation}
This statement should be read with its theorem-level assumptions attached.  The
classical Brenier characterization is a continuous-measure result for quadratic
cost in Euclidean space under regularity conditions such as finite second
moments and absolute continuity of the source measure with respect to Lebesgue
measure.  In the finite empirical-cloud setting used throughout BayesFilter, a
Monge map may fail to exist or fail to be unique in that theorem-strength sense.
So the present chapter uses the Brenier viewpoint primarily as structured
motivation for a learned map family, not as a blanket theorem that every finite
cloud transport problem in the particle-filter setting is exactly represented by a
gradient of a convex potential.

In one dimension, this means the derivative of the potential defines the map.
For the translation cell above, one convenient choice is
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier-example}
  \psi^\star(x)=\frac{1}{2}x^2+x.
\end{equation}
Then
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier-example-grad}
  \nabla\psi^\star(x)=x+1=T^\star(x).
\end{equation}
The second derivative of \(\psi^\star\) is \(1\), so the potential is convex.
This tiny calculation is the cleanest way to see what the Brenier/ICNN family is
trying to exploit: instead of learning the map directly, it learns a structured
potential whose gradient gives the map.

\paragraph{Reader checkpoint.}
The direct-map route is now more concrete.  Instead of predicting the teacher’s
internal solver coordinates, it tries to learn a map \(T\), often through a
potential \(\psi\) that imposes structure on what that map can be.

\section{ICNN parameterization and training contract}
\label{sec:bf-neural-ot-direct-icnn}

Input-convex neural networks (ICNNs) provide a parametric family of scalar
functions that remain convex in designated inputs.  The implementation idea is:
\begin{enumerate}[label=(\roman*)]
  \item parameterize the potential \(\psi\) by an ICNN \(\psi_\theta\),
  \item obtain the transport map by differentiating the network,
  \(
    T_\theta(x)=\nabla\psi_\theta(x),
  \)
  \item train \(\psi_\theta\) so that the induced map sends source mass toward the
  target according to the chosen OT objective.
\end{enumerate}

A useful objective to keep in view is the convex-potential dual form
\begin{equation}
\label{eq:bf-neural-ot-direct-dual-objective}
  J(\psi)
  =
  \mathbb E_{X\sim\mu}[\psi(X)]
  +
  \mathbb E_{Y\sim\nu}[\psi^*(Y)],
\end{equation}
where \(\psi^*\) is the convex conjugate
\begin{equation}
\label{eq:bf-neural-ot-direct-conjugate}
  \psi^*(y)=\sup_x\{x^\top y-\psi(x)\}.
\end{equation}
This is best read as continuous-measure theorem language or as motivation for a
structured parametric family, not as a claim that an arbitrary finite-cloud
training loop automatically inherits the full optimal-transport theorem.  In a
finite empirical implementation, one still has to choose how \(\mu\) and \(\nu\)
are represented, how gradients of the potential are evaluated on particle clouds,
and how any conjugate or dual approximation is actually computed numerically.
The attraction of this route is that the map is not a black-box regressor.  It is
constrained to come from a convex potential, which is exactly the sort of
structure the quadratic-cost theory suggests under its stated assumptions.

The cost is also immediate from the equations.  One must evaluate gradients of the
potential and, depending on the chosen training formulation, may need to evaluate
or approximate the convex conjugate.  So the route trades a retained transport
solver for a more structured but more delicate map-learning objective.

\begin{algorithm}[ht]
\caption{\texttt{train\_icnn\_brenier\_map}}
\begin{algorithmic}[1]
\Require source-cloud sampler for \(\mu\), target-cloud sampler for \(\nu\), ICNN potential family \(\psi_\theta\), declared quadratic-cost/Brenier setting, convexity-enforcement policy, optimizer, learning-rate schedule, batch size, stopping rule, and held-out validation clouds
\Ensure trained potential \(\psi_\theta\), induced map \(T_\theta(x)=\nabla\psi_\theta(x)\), and validation diagnostics for pushforward mismatch and scalar drift
\State Initialize ICNN parameters \(\theta\) and optimizer state
\While{the stopping rule has not been met}
  \State Draw minibatches \(x_b\sim\mu\) and \(y_b\sim\nu\)
  \State Evaluate the potential values \(\psi_\theta(x_b)\)
  \State Evaluate the gradient-defined transport map \(T_\theta(x_b)=\nabla\psi_\theta(x_b)\)
  \State Compute the declared dual or potential-based training objective, including any required convex-conjugate or surrogate terms
  \State Add any declared regularization enforcing smoothness, stability, or numerical tractability
  \State Backpropagate the total loss and update \(\theta\)
  \State Re-enforce convexity according to the declared ICNN policy if it is not built directly into the parameterization
\EndWhile
\State On held-out validation clouds, compute mapped particles \(T_\theta(x_i)\)
\State Evaluate endpoint pushforward mismatch, transport-cost proxy, and downstream scalar sensitivity on the held-out clouds
\State Return \(\psi_\theta\), the induced map \(T_\theta\), and the declared diagnostics
\end{algorithmic}
\end{algorithm}

\paragraph{Approximation enters here.}
The learned object is the transport map itself.  Errors in the learned potential or
its gradient directly change the transported output cloud.

\section{Monge-gap regularization}
\label{sec:bf-neural-ot-direct-monge-gap}

A more flexible direct-map family regularizes a candidate map by how far it
behaves from an OT map rather than forcing a particular architecture from the
start.  For cost \(c\) and reference measure \(\rho\), the Monge gap is
\begin{equation}
\label{eq:bf-neural-ot-direct-monge-gap}
  \mathcal M_\rho^c(T)
  =
  \int c(x,T(x))\,d\rho(x)-W_c(\rho,T_\#\rho).
\end{equation}
As a mathematical definition, this is exact.  But in a practical training loop,
the second term is typically not available in closed form because the pushforward
measure \(T_\#\rho\) itself depends on the current learned map.  So any finite-
sample training objective that uses an estimated OT reference term should be read
as a surrogate or approximate Monge-gap objective unless the estimator and its
properties are stated explicitly.  Operationally, the chapter’s point is:
fit a useful map, but penalize it when it behaves unlike an OT map for the chosen
cost.

A tiny example makes the gap visible.  Keep the source cloud
\(\{-1,1\}\) and target cloud \(\{0,2\}\).  Consider two maps:
\begin{align}
\label{eq:bf-neural-ot-direct-good-bad-maps}
  T_{\mathrm{good}}(-1)&=0,
  &T_{\mathrm{good}}(1)&=2,\\
  T_{\mathrm{bad}}(-1)&=2,
  &T_{\mathrm{bad}}(1)&=0.
\end{align}
Both maps have the same pushforward target cloud \(\{0,2\}\), but they incur
very different transport costs under the quadratic metric.  The good map has cost
\[
  \frac{1}{2}(1^2+1^2)=1,
\]
whereas the bad map has cost
\[
  \frac{1}{2}(3^2+1^2)=5.
\]
So the bad map reaches the same endpoint law while moving mass much less
efficiently.  The Monge-gap idea is precisely to penalize such behavior.

\begin{algorithm}[ht]
\caption{\texttt{train\_monge\_gap\_regularized\_direct\_map}}
\begin{algorithmic}[1]
\Require source-cloud sampler for \(\mu\), target-cloud sampler for \(\nu\), reference-measure sampler for \(\rho\), parametric map family \(T_\theta\), cost function \(c\), endpoint-fit objective, Monge-gap weight \(\lambda_{\mathrm{MG}}\), optimizer, stopping rule, and held-out validation suite
\Ensure trained direct map \(T_\theta\), Monge-gap diagnostic history, endpoint-fit diagnostics, and downstream scalar sensitivity report
\State Initialize map parameters \(\theta\) and optimizer state
\While{the stopping rule has not been met}
  \State Draw minibatches from the source, target, and reference distributions
  \State Compute mapped source particles \(z_i=T_\theta(x_i)\)
  \State Evaluate the declared endpoint-fit loss between the mapped source batch and the target batch
  \State Compute the empirical transport-cost term on the reference samples from \(\rho\)
  \State Estimate the OT reference term needed for the empirical Monge-gap penalty; this estimate defines the practical training surrogate unless a stronger equivalence result is supplied
  \State Form the total loss as endpoint-fit term plus \(\lambda_{\mathrm{MG}}\) times the empirical Monge-gap penalty, together with any declared regularization
  \State Backpropagate the total loss and update \(\theta\)
\EndWhile
\State On held-out validation clouds, evaluate endpoint mismatch, Monge-gap magnitude, transport-cost proxy, and downstream scalar drift
\State Return the trained map \(T_\theta\) and the declared diagnostics
\end{algorithmic}
\end{algorithm}

\paragraph{Reader checkpoint.}
The Monge-gap route still learns a direct map, but it relaxes the strong ICNN-only
structural commitment.  The gain is flexibility.  The cost is that the learned map
is now judged relative to a regularized OT-like behavior rather than by a retained
teacher solver.

\section{Structured direct-map variants}
\label{sec:bf-neural-ot-direct-structured}

The same direct-map logic can be combined with additional structural priors, such
as sparse displacement or factorized transport structure.  These are worth naming
because they may be useful when the filtering update itself has a scientifically
justified sparse or low-active-subspace geometry.  They should not be read as free
improvements in all settings.

\section{Why this chapter is not the default immediate route}
\label{sec:bf-neural-ot-direct-why-not-default}

The direct-map families are mathematically serious and potentially powerful, but
there are two reasons they are not the monograph’s default first implementation
route.
\begin{enumerate}[label=(\roman*)]
  \item They learn the final map object itself rather than only accelerating a
  declared teacher solve.
  \item Their strongest mathematical support depends on more specific geometric
  assumptions than the retained-teacher acceleration route requires.
\end{enumerate}

\section{Verification and non-claims}
\label{sec:bf-neural-ot-direct-verification}

A minimal implementation-facing reading of this chapter should distinguish:
\begin{itemize}
  \item what map object is being fit,
  \item what assumptions justify that map family,
  \item what direct-map residuals are measured,
  \item what stronger filtering or HMC claims are still not justified.
\end{itemize}

This chapter does \emph{not} claim:
\begin{itemize}
  \item that direct learned maps are interchangeable with retained-teacher EOT
  acceleration,
  \item that a map-level fit automatically preserves the same filtering scalar,
  \item that broader direct-map families are already validated for BayesFilter.
\end{itemize}
Its role is to slow down the direct-map family itself so the reader can see that
it is solving a different problem from the previous chapter.

\FloatBarrier
