\chapter{Kalman Hessian and Observed Information}
\label{ch:bf-kalman-hessian}

The Hessian layer is necessary for observed-information diagnostics, local
identification, and mass-matrix construction, but it is also the easiest place
to overstate the state of the mathematics.  BayesFilter therefore records the
Hessian as a source-backed contract and validation target, not as a finished
proof for every backend.

For second derivatives, write
\[
  \ddot z^{(ij)}
  =
  \frac{\partial^2 z}{\partial\theta_i\partial\theta_j}.
\]
The MacroFinance note expands the second-order prediction, innovation, and gain
recursions by product rule.  In solve form, the source introduces
\begin{equation}
\label{eq:bf-solve-hessian-w-dot}
  \dot w_t^{(j)}
  =
  \mathcal{S}_{S_t}\left(\dot v_t^{(j)}-\dot S_t^{(j)}w_t\right),
\end{equation}
where $\mathcal{S}_{S_t}(b)$ denotes solving $S_tx=b$.  It then decomposes the
per-time Hessian contribution into log-determinant, innovation-linear, and
quadratic terms:
\begin{equation}
\label{eq:bf-solve-hessian-summary}
  \frac{\partial^2\ell_t}{\partial\theta_i\partial\theta_j}
  =
  -\frac{1}{2}
  \left[
    T_{ij}^{\log\det}+T_{ij}^{v}-T_{ij}^{S}
  \right].
\end{equation}

\section{Second-Order Prediction}
\label{sec:bf-hessian-prediction}

The second-order prediction step is the product-rule derivative of
\eqref{eq:bf-score-predicted-mean-first}--\eqref{eq:bf-score-predicted-cov-first}.
For the predicted mean,
\begin{align}
\label{eq:bf-hessian-predicted-mean}
  \ddot m_{t|t-1}^{(ij)}
  &=
  \ddot c_t^{(ij)}
  +\ddot F_t^{(ij)}m_{t-1|t-1}
  +\dot F_t^{(i)}\dot m_{t-1|t-1}^{(j)}
  +\dot F_t^{(j)}\dot m_{t-1|t-1}^{(i)}
  +F_t\ddot m_{t-1|t-1}^{(ij)} .
\end{align}
For the predicted covariance,
\begin{align}
\label{eq:bf-hessian-predicted-cov}
  \ddot P_{t|t-1}^{(ij)}
  &=
  \ddot F_t^{(ij)}P_{t-1|t-1}F_t^\top
  +\dot F_t^{(i)}\dot P_{t-1|t-1}^{(j)}F_t^\top
  +\dot F_t^{(i)}P_{t-1|t-1}\dot F_t^{(j)\top} \nonumber\\
  &\quad
  +\dot F_t^{(j)}\dot P_{t-1|t-1}^{(i)}F_t^\top
  +F_t\ddot P_{t-1|t-1}^{(ij)}F_t^\top
  +F_t\dot P_{t-1|t-1}^{(i)}\dot F_t^{(j)\top} \nonumber\\
  &\quad
  +\dot F_t^{(j)}P_{t-1|t-1}\dot F_t^{(i)\top}
  +F_t\dot P_{t-1|t-1}^{(j)}\dot F_t^{(i)\top}
  +F_tP_{t-1|t-1}\ddot F_t^{(ij)\top}
  +\ddot Q_t^{(ij)} .
\end{align}
These equations are the second-order source recursions in the MacroFinance
analytic derivative note.

\section{Second-Order Innovation Objects}
\label{sec:bf-hessian-innovation}

The innovation derivatives are
\begin{align}
\label{eq:bf-hessian-innovation}
  \ddot v_t^{(ij)}
  &=
  -\ddot a_t^{(ij)}
  -\ddot H_t^{(ij)}m_{t|t-1}
  -\dot H_t^{(i)}\dot m_{t|t-1}^{(j)}
  -\dot H_t^{(j)}\dot m_{t|t-1}^{(i)}
  -H_t\ddot m_{t|t-1}^{(ij)},\\
\label{eq:bf-hessian-innovation-cov}
  \ddot S_t^{(ij)}
  &=
  \ddot H_t^{(ij)}P_{t|t-1}H_t^\top
  +\dot H_t^{(i)}\dot P_{t|t-1}^{(j)}H_t^\top
  +\dot H_t^{(i)}P_{t|t-1}\dot H_t^{(j)\top} \nonumber\\
  &\quad
  +\dot H_t^{(j)}\dot P_{t|t-1}^{(i)}H_t^\top
  +\dot H_t^{(j)}P_{t|t-1}\dot H_t^{(i)\top}
  +H_t\ddot P_{t|t-1}^{(ij)}H_t^\top \nonumber\\
  &\quad
  +H_t\dot P_{t|t-1}^{(i)}\dot H_t^{(j)\top}
  +H_t\dot P_{t|t-1}^{(j)}\dot H_t^{(i)\top}
  +H_tP_{t|t-1}\ddot H_t^{(ij)\top}
  +\ddot R_t^{(ij)}.
\end{align}
Together with the first-order innovation objects from
Chapter~\ref{ch:bf-kalman-score}, these are the only model-specific inputs
needed by the solve-form likelihood Hessian.

\section{Solve-Form Hessian}
\label{sec:bf-solve-form-hessian}

The compact statement above is useful only if the three terms are kept explicit.
Let $S_tw_t=v_t$ and define $\dot w_t^{(j)}$ by
\eqref{eq:bf-solve-hessian-w-dot}.  The second derivative of the solved vector is
\begin{equation}
\label{eq:bf-solve-hessian-w-ddot}
  \ddot w_t^{(ij)}
  =
  \mathcal{S}_{S_t}\left(
    \ddot v_t^{(ij)}
    -\ddot S_t^{(ij)}w_t
    -\dot S_t^{(i)}\dot w_t^{(j)}
    -\dot S_t^{(j)}\dot w_t^{(i)}
  \right).
\end{equation}
The log-determinant term is
\begin{equation}
\label{eq:bf-solve-hessian-logdet}
  T_{ij}^{\log\det}
  =
  \tr\!\left(
    S_t^{-1}\ddot S_t^{(ij)}
    -S_t^{-1}\dot S_t^{(j)}S_t^{-1}\dot S_t^{(i)}
  \right).
\end{equation}
The innovation-linear term is
\begin{equation}
\label{eq:bf-solve-hessian-linear}
  T_{ij}^{v}
  =
  2\ddot v_t^{(ij)\top}w_t
  +2\dot v_t^{(i)\top}\dot w_t^{(j)}.
\end{equation}
The quadratic term is
\begin{equation}
\label{eq:bf-solve-hessian-quadratic}
  T_{ij}^{S}
  =
  2\dot w_t^{(j)\top}\dot S_t^{(i)}w_t
  +w_t^\top\ddot S_t^{(ij)}w_t .
\end{equation}
Substituting \eqref{eq:bf-solve-hessian-logdet}--\eqref{eq:bf-solve-hessian-quadratic}
into \eqref{eq:bf-solve-hessian-summary} gives the observed Hessian contribution.
The formula requires only the innovation objects
\[
  v_t,\quad S_t,\quad
  \dot v_t^{(i)},\quad \dot S_t^{(i)},\quad
  \ddot v_t^{(ij)},\quad \ddot S_t^{(ij)}.
\]
It is therefore backend-neutral.  A linear Kalman filter, a square-root Kalman
filter, or a sigma-point filter may all use the same likelihood Hessian layer
once their own recursions supply these six objects.

\section{Second-Order Gain and Update}
\label{sec:bf-hessian-update}

The covariance-form backend also propagates the Hessian of the filtered state
objects.  Define
\begin{align}
\label{eq:bf-hessian-precision-first}
  \dot S_t^{-1(i)}
  &=
  -S_t^{-1}\dot S_t^{(i)}S_t^{-1},\\
\label{eq:bf-hessian-precision-second}
  \ddot S_t^{-1(ij)}
  &=
  S_t^{-1}\dot S_t^{(j)}S_t^{-1}\dot S_t^{(i)}S_t^{-1}
  +S_t^{-1}\dot S_t^{(i)}S_t^{-1}\dot S_t^{(j)}S_t^{-1}
  -S_t^{-1}\ddot S_t^{(ij)}S_t^{-1}.
\end{align}
Then
\begin{align}
\label{eq:bf-hessian-gain-first}
  \dot K_t^{(i)}
  &=
  \dot P_{t|t-1}^{(i)}H_t^\top S_t^{-1}
  +P_{t|t-1}\dot H_t^{(i)\top}S_t^{-1}
  +P_{t|t-1}H_t^\top\dot S_t^{-1(i)},\\
\label{eq:bf-hessian-gain-second}
  \ddot K_t^{(ij)}
  &=
  \ddot P_{t|t-1}^{(ij)}H_t^\top S_t^{-1}
  +\dot P_{t|t-1}^{(i)}\dot H_t^{(j)\top}S_t^{-1}
  +\dot P_{t|t-1}^{(i)}H_t^\top\dot S_t^{-1(j)}
  \nonumber\\
  &\quad
  +\dot P_{t|t-1}^{(j)}\dot H_t^{(i)\top}S_t^{-1}
  +P_{t|t-1}\ddot H_t^{(ij)\top}S_t^{-1}
  +P_{t|t-1}\dot H_t^{(i)\top}\dot S_t^{-1(j)}
  \nonumber\\
  &\quad
  +\dot P_{t|t-1}^{(j)}H_t^\top\dot S_t^{-1(i)}
  +P_{t|t-1}\dot H_t^{(j)\top}\dot S_t^{-1(i)}
  +P_{t|t-1}H_t^\top\ddot S_t^{-1(ij)}.
\end{align}
The filtered mean Hessian is
\begin{align}
\label{eq:bf-hessian-filtered-mean}
  \ddot m_{t|t}^{(ij)}
  &=
  \ddot m_{t|t-1}^{(ij)}
  +\ddot K_t^{(ij)}v_t
  +\dot K_t^{(i)}\dot v_t^{(j)}
  +\dot K_t^{(j)}\dot v_t^{(i)}
  +K_t\ddot v_t^{(ij)}.
\end{align}
For the simplified covariance update,
\begin{align}
\label{eq:bf-hessian-filtered-cov}
  \ddot P_{t|t}^{(ij)}
  &=
  -\ddot K_t^{(ij)}H_tP_{t|t-1}
  -\dot K_t^{(i)}\dot H_t^{(j)}P_{t|t-1}
  -\dot K_t^{(i)}H_t\dot P_{t|t-1}^{(j)}
  \nonumber\\
  &\quad
  -\dot K_t^{(j)}\dot H_t^{(i)}P_{t|t-1}
  -K_t\ddot H_t^{(ij)}P_{t|t-1}
  -K_t\dot H_t^{(i)}\dot P_{t|t-1}^{(j)}
  \nonumber\\
  &\quad
  -\dot K_t^{(j)}H_t\dot P_{t|t-1}^{(i)}
  -K_t\dot H_t^{(j)}\dot P_{t|t-1}^{(i)}
  +(I-K_tH_t)\ddot P_{t|t-1}^{(ij)}.
\end{align}
The gain and covariance formulas are useful as a covariance-form audit oracle.
Solve-form or square-root implementations may avoid explicit
\eqref{eq:bf-hessian-precision-first}--\eqref{eq:bf-hessian-precision-second}
internally, but they must agree with these recursions on controlled
well-conditioned systems.

\section{Information Objects and Sign Convention}
\label{sec:bf-hessian-information}

This chapter computes Hessians of the log likelihood contribution
$\ell_t(\theta)$.  The observed information contribution is
\[
  \mathcal{J}_t(\theta)=-\nabla_\theta^2\ell_t(\theta),
\]
and posterior precision at a mode additionally includes the negative Hessian of
the log prior.  BayesFilter should not call a matrix a mass matrix, local
covariance, or identification object unless the sign convention and prior
contribution have been stated explicitly.

\section{Hessian-Ready Backend Contract}
\label{sec:bf-hessian-ready-contract}

A backend is Hessian-ready when it can return, for every time step and active
parameter pair,
\[
  v_t,\quad S_t,\quad
  \dot v_t^{(i)},\quad \dot S_t^{(i)},\quad
  \ddot v_t^{(ij)},\quad \ddot S_t^{(ij)}
\]
for the same likelihood value it evaluates.  To propagate the scan, it must also
return either the covariance-form update derivatives
\eqref{eq:bf-hessian-filtered-mean}--\eqref{eq:bf-hessian-filtered-cov} or an
equivalent factor/backend-specific derivative update with reconstruction
diagnostics.  This is the bridge used later by the SVD sigma-point chapter: the
nonlinear backend supplies approximate innovation derivatives, and the
solve-form Hessian layer remains the same.

\section{Evidence Status}
\label{sec:bf-hessian-evidence-status}

The source derivation is detailed, but BayesFilter should keep three evidence
levels distinct:
\begin{itemize}
  \item formula provenance from the MacroFinance analytic derivative note;
  \item structural code evidence from MathDevMCP AST audits;
  \item numerical parity evidence from tests against autodiff and finite
    differences.
\end{itemize}

The current MathDevMCP audit of the MacroFinance solve differentiated Kalman
backend found the Kalman operation graph and solve structure but reported a
mismatch for a strict \texttt{logdet}/\texttt{trace} operation query and missing
explicit shape and covariance guards.  This should be treated as a
production-readiness prompt, not as a proof failure.  The generic LGSSM
autodiff-validation tests are the stronger mathematical check for the current
implementation because they compare likelihood, gradient, and Hessian against
solve backends, TensorFlow autodiff, and finite differences on controlled
small-dimensional models.
