\chapter{Mass Matrices and Local Geometry}
\label{ch:bf-mass-matrices}

Mass matrices translate derivative information into sampler geometry.  In a
linear Gaussian validation problem, a local posterior Hessian or observed
information matrix can be a useful preconditioner.  In a nonlinear or weakly
identified model, the same matrix can be misleading unless it is regularized and
checked against diagnostics.

\section{Sources of Curvature}
\label{sec:bf-curvature-sources}

BayesFilter recognizes several curvature sources:
\begin{itemize}
  \item analytic likelihood Hessian plus prior curvature;
  \item observed-information or expected-information approximations;
  \item finite-difference Hessian diagnostics at a MAP point;
  \item empirical covariance from warmup draws;
  \item diagonal fallbacks when dense curvature is unstable.
\end{itemize}
The source must be recorded because each source has a different failure mode.
MacroFinance mass-matrix tests check eigenvalue regularization, dense versus
diagonal choices, whitening finiteness, and covariance construction from
Hessian helpers.  These are geometry utilities; they are not identification
proofs.

\section{Regularization Policy}
\label{sec:bf-mass-regularization}

A mass matrix used by HMC must be symmetric positive definite.  If a curvature
estimate is indefinite or ill-conditioned, BayesFilter should regularize it,
record the regularization, and expose the eigenvalue diagnostics.  Silently
using an indefinite curvature matrix as if it were a covariance is a target
geometry error.

\section{Geometry-Scaled Tuning Budgets}
\label{sec:bf-geometry-scaled-tuning-budgets}

Mass-matrix diagnostics also determine how much evidence a tuning stage needs
before BayesFilter makes a repair decision.  A fixed number of samples is not a
property of the model; it is a policy choice.  BayesFilter therefore records a
public-safe geometry summary before serious HMC tuning:

\begin{itemize}
  \item parameter dimension \(d\);
  \item covariance condition number, when available;
  \item effective dimension, computed from the covariance eigenvalue spectrum;
  \item anisotropy ratio \(d/d_{\mathrm{eff}}\);
  \item regularization pressure, including eigenvalue clipping, non-positive
    raw eigenvalues, and diagonal fallback use.
\end{itemize}

The serious tuning budget uses this summary only to decide how much local
tuning work to allocate.  In code, the central policy is
\texttt{HMCGeometryScaledBudgetTimingPolicy}.  Its first attempt budget is
\[
  B_0 =
  \operatorname{clamp}\left(
    \left\lceil c_d\, d\, m_{\mathrm{geom}}\right\rceil,
    B_{\min}, B_{\max,0}
  \right),
\]
where \(m_{\mathrm{geom}}\) is a clipped multiplier built from condition,
anisotropy, and regularization pressure.  Later attempts grow as
\[
  B_k = \min(B_{\max}, B_0 2^k).
\]
Bootstrap screens use a smaller square-root dimension rule,
\[
  N_{\mathrm{boot}} =
  \operatorname{clamp}\left(
    \left\lceil c_{\sqrt d}\sqrt d\,m_{\mathrm{geom}}\right\rceil,
    N_{\min}, N_{\max}
  \right),
\]
with burn-in chosen as a fixed fraction of the screen length.

These formulas are not posterior convergence criteria.  They are a disciplined
replacement for arbitrary constants: larger, more anisotropic, or more heavily
regularized targets receive more local tuning evidence before BayesFilter
decides whether a step-size, trajectory, or mass repair is making progress.

\section{Timing And Progress}
\label{sec:bf-hmc-timing-and-progress}

Wall-clock limits are treated differently from statistical tuning budgets.  A
stage may run for a long time because the model is large but still making
meaningful progress.  BayesFilter therefore separates:

\begin{itemize}
  \item progress evidence, such as stage transitions, updated mass summaries,
    changed step-size hashes, changed trajectory hashes, repair decisions, and
    fresh verification summaries;
  \item emergency caps, which are machine-protection limits rather than claims
    that the sampler has failed scientifically.
\end{itemize}

The staged timeout policy records emergency stage caps, a global cap, and a
reserve.  Those numbers should be read as safety rails.  A scientific result
note must not interpret an emergency cap as evidence against the posterior
unless the predeclared continuation veto says so.
