\chapter{Entropic OT, Sinkhorn, and Barycentric Projection}
\label{ch:bf-dpf-entropic-ot-sinkhorn}

The previous chapter fixed the deterministic OT equal-weighting baseline.  This
chapter explains the main differentiable computational route used in the OT lane:
entropic regularization, finite Sinkhorn iteration, barycentric projection, and
the distinction between the mathematical teacher object and the numerical object
actually differentiated in code.

This chapter has one dominant goal: a reader should be able to understand what
object finite Sinkhorn computes, how barycentric output is produced from it, where
approximation enters, and what diagnostics must be checked before the result is
used in a downstream filtering scalar.

\section{What changes when entropy is added}
\label{sec:bf-eot-what-changes}

The previous chapter fixed the unregularized deterministic baseline by choosing a
feasible coupling with minimal declared transport cost,
\begin{equation}
\label{eq:bf-eot-deterministic-baseline-repeat}
  \Pi_0^\star
  \in
  \arg\min_{\Pi\in\mathcal U(w,u)}
  \langle C,\Pi\rangle.
\end{equation}
Entropic OT changes that teacher object before any numerical approximation enters.
For a regularization level \(\varepsilon>0\), it replaces the linear program by
an objective that also penalizes overly concentrated couplings through the entropy
term.  In plain language, the plan is now rewarded for spreading mass more smoothly
instead of concentrating all mass as sharply as possible.

This change matters mathematically as well as numerically.  The transport object is
no longer the unregularized deterministic optimizer from
\eqref{eq:bf-eot-deterministic-baseline-repeat}.  It is a different teacher problem
whose solution is smoother and whose first-order conditions can be rewritten as a
matrix-scaling problem.  That rewrite is what later makes Sinkhorn iteration
possible.  Thus the chapter's logic is:
\begin{enumerate}[label=(\roman*)]
  \item change the deterministic teacher by adding entropy,
  \item derive the factorized form of the new optimizer,
  \item derive the balancing equations for its scaling vectors,
  \item and only then discuss the finite numerical route used in code.
\end{enumerate}

The gain is numerical smoothness and a tractable scaling-based solver.  The cost is
that the transport object is no longer the unregularized deterministic OT solution.
This is the chapter's first approximation layer, and later finite iteration will add
another one on top of it.

\section{Running three-particle cell with entropic smoothing}
\label{sec:bf-eot-running-cell}

Keep the same running cell as before:
\[
  x_1=-2,
  \qquad x_2=0,
  \qquad x_3=3,
  \qquad
  w=
  \left(
    \frac{1}{2},
    \frac{1}{3},
    \frac{1}{6}
  \right),
  \qquad
  u=
  \left(
    \frac{1}{3},
    \frac{1}{3},
    \frac{1}{3}
  \right),
\]
with quadratic cost matrix
\begin{equation}
\label{eq:bf-eot-running-cost}
  C
  =
  \begin{pmatrix}
    0 & 4 & 25 \\
    4 & 0 & 9 \\
    25 & 9 & 0
  \end{pmatrix}.
\end{equation}
For \(\varepsilon=1\), the Gibbs kernel is
\begin{equation}
\label{eq:bf-eot-running-kernel}
  K_1=
  \begin{pmatrix}
    1 & e^{-4} & e^{-25} \\
    e^{-4} & 1 & e^{-9} \\
    e^{-25} & e^{-9} & 1
  \end{pmatrix}.
\end{equation}
The entries are heavily concentrated near the diagonal, but not exactly zero away
from it.  That is a simple visual sign of what entropy has changed: the plan is no
longer forced into the sharpest possible assignment.

\section{Entropic OT teacher object}
\label{sec:bf-eot-teacher-object}

Let \(\mathcal U(w,u)\) be the feasible coupling set from the previous chapter.
The entropically regularized OT problem is \citep{cuturi2013sinkhorn,peyre2019computational}
\begin{equation}
\label{eq:bf-eot-primal}
  \Pi_\varepsilon^\star
  =
  \arg\min_{\Pi\in\mathcal U(w,u)}
  \left\{
    \langle \Pi,C\rangle
    + \varepsilon\sum_{i,j}\Pi_{ij}(\log\Pi_{ij}-1)
  \right\}.
\end{equation}
This chapter uses the same entropy convention as the earlier OT material, so the
minimizer is the exact teacher for the chosen regularized object, not for the
unregularized linear program and not for categorical resampling.

\paragraph{Reader checkpoint.}
At this stage, the chapter has changed the mathematical object once: from the
unregularized deterministic OT coupling to the entropically smoothed OT coupling.
Nothing numerical has happened yet.  The next step still stays at the exact-object
level: first identify the factorized teacher form, and only after that turn to the
finite executable route.

\subsection{Exact teacher versus exact engineering route}
\label{sec:bf-eot-teacher-versus-engineering}

The scalable-OT survey suggests a useful extra distinction before moving on to
finite Sinkhorn iterations.  There are two different senses in which a route may
still be called ``exact'' for the entropic OT object.

First, there is the mathematical teacher object
\(\Pi_\varepsilon^\star\) from \eqref{eq:bf-eot-primal}.  That is the exact
minimizer of the entropically regularized problem.  Second, there is an
engineering route that preserves that same entropic object while changing how the
matrix-scaling updates are executed in memory or on hardware.  Exact online,
streaming, and GPU Sinkhorn methods belong to this second category: they do not
change the entropic teacher problem, but they avoid materializing the full cost,
kernel, or coupling whenever possible \citep{feydy2019geomloss,charlier2021keops,ye2026flashsinkhorn,xiao2026fastsinkhorn}.

This distinction matters for the BayesFilter transport lane.  A method that
recomputes the same entropic scaling updates through blocked reductions or
streamed kernel application is still on the semantics-preserving reference lane.
By contrast, a method that replaces the kernel, truncates the coupling family, or
changes the transport objective has moved to a different approximation class even
if it is motivated by scalability.  The reader should therefore keep separate:
\begin{enumerate}[label=(\roman*)]
  \item the exact entropic teacher object,
  \item the finite numerical object produced by a declared Sinkhorn execution
    path,
  \item the engineering strategy used to realize that numerical path.
\end{enumerate}
The next sections fix the second object explicitly and later staged chapters will
discuss scalability routes that either preserve or alter it.

\subsection{The current BayesFilter bottleneck}
\label{sec:bf-eot-current-bottleneck}

The need for scalable OT in BayesFilter is not only a generic literature concern.
The current TensorFlow route already distinguishes dense and streamed execution.
Dense mode materializes the effective transport matrix and applies it to the
particle cloud directly.  Streamed mode avoids storing that full matrix, but it
still loops over row and column blocks and therefore still computes the same
underlying pairwise transport interactions.

For equal-weight barycentric output, the streamed route still forms costs of the
schematic form
\begin{equation}
\label{eq:bf-eot-current-cost}
  C_{ij}
  =
  \frac{1}{2}\|\tilde x_i-\tilde x_j\|_2^2,
\end{equation}
then log-transport weights of the schematic form
\begin{equation}
\label{eq:bf-eot-current-log-transport}
  \log T_{ij}
  =
  \frac{f_i+g_j-C_{ij}}{\varepsilon}
  - \ell_j
  + \log N
  + \log w_j,
\end{equation}
and finally blockwise transport application by accumulation,
\begin{equation}
\label{eq:bf-eot-current-streamed-application}
  z_i = \sum_j T_{ij}x_j.
\end{equation}
The gain is real: memory pressure is reduced because the full transport matrix need
not be kept in RAM or device memory.  But the all-pairs arithmetic burden remains.
Streaming therefore solves part of the storage problem without yet solving the full
large-\(N\), large-\(d_x\) interaction problem.  This is exactly why the derivation
below matters: once the teacher object has been rewritten as a scaling problem, one
can see clearly which later routes merely change execution strategy and which ones
change the transport object itself.

\section{Why the optimizer factorizes}
\label{sec:bf-eot-factorization}

The next step is to derive the multiplicative structure of the entropic optimizer
rather than only state it.  Introduce Lagrange multipliers \(\alpha\in\mathbb R^N\)
for the row constraints and \(\beta\in\mathbb R^N\) for the column constraints.
With the entropy convention from \eqref{eq:bf-eot-primal}, the Lagrangian is
\begin{align}
\label{eq:bf-eot-lagrangian}
  \mathcal L(\Pi,\alpha,\beta)
  &=
  \sum_{i,j}
  \left[
    C_{ij}\Pi_{ij}
    + \varepsilon \Pi_{ij}(\log\Pi_{ij}-1)
  \right]
  + \sum_i \alpha_i\left(\sum_j\Pi_{ij}-w_i\right) \\
  &\qquad
  + \sum_j \beta_j\left(\sum_i\Pi_{ij}-u_j\right).
\end{align}
Stationarity with respect to each entry \(\Pi_{ij}\) gives
\begin{equation}
\label{eq:bf-eot-stationarity}
  0
  =
  \frac{\partial \mathcal L}{\partial \Pi_{ij}}
  =
  C_{ij}
  + \varepsilon \log \Pi_{ij}
  + \alpha_i + \beta_j,
\end{equation}
so
\begin{equation}
\label{eq:bf-eot-log-pi}
  \log\Pi_{ij}
  =
  -\frac{C_{ij}}{\varepsilon}
  -\frac{\alpha_i}{\varepsilon}
  -\frac{\beta_j}{\varepsilon}.
\end{equation}
Exponentiating entrywise yields
\begin{equation}
\label{eq:bf-eot-entrywise-factorization}
  \Pi_{ij}
  =
  \exp\!\left(-\frac{\alpha_i}{\varepsilon}\right)
  \exp\!\left(-\frac{C_{ij}}{\varepsilon}\right)
  \exp\!\left(-\frac{\beta_j}{\varepsilon}\right).
\end{equation}
Define the positive scalings
\begin{equation}
\label{eq:bf-eot-scaling-definitions}
  a_i = \exp(-\alpha_i/\varepsilon),
  \qquad
  b_j = \exp(-\beta_j/\varepsilon),
  \qquad
  (K_\varepsilon)_{ij}=\exp(-C_{ij}/\varepsilon).
\end{equation}
Then the optimizer must have the Gibbs-scaled form
\begin{equation}
\label{eq:bf-eot-factorization}
  \Pi_\varepsilon^\star
  =
  \diag(a)K_\varepsilon\diag(b).
\end{equation}
This is the key structural step.  The entropic teacher problem has been converted
from a constrained optimization over all couplings into the search for two positive
scaling vectors that make a known kernel satisfy the marginal constraints.

\subsection{Why the balancing equations are Sinkhorn updates}
\label{sec:bf-eot-balancing-equations}

Once \eqref{eq:bf-eot-factorization} is available, the marginal constraints from
\(\mathcal U(w,u)\) become explicit equations for the scaling vectors.  The row
constraints require
\begin{equation}
\label{eq:bf-eot-row-balance-derived}
  \Pi_\varepsilon^\star\mathbf 1
  =
  \diag(a)K_\varepsilon\diag(b)\mathbf 1
  =
  a \odot (K_\varepsilon b)
  = w,
\end{equation}
where \(\odot\) denotes entrywise multiplication.  Therefore
\begin{equation}
\label{eq:bf-eot-a-balance}
  a = w \oslash (K_\varepsilon b).
\end{equation}
Likewise the column constraints require
\begin{equation}
\label{eq:bf-eot-col-balance-derived}
  (\Pi_\varepsilon^\star)^\top\mathbf 1
  =
  \diag(b)K_\varepsilon^\top\diag(a)\mathbf 1
  =
  b \odot (K_\varepsilon^\top a)
  = u,
\end{equation}
so
\begin{equation}
\label{eq:bf-eot-b-balance}
  b = u \oslash (K_\varepsilon^\top a).
\end{equation}
These are exactly the balancing equations behind Sinkhorn iteration.  The algorithm
is not a mysterious OT oracle added from outside the derivation.  It simply
alternates the two marginal repairs implied by the factorized optimizer.

\paragraph{Reader checkpoint.}
The chapter has now introduced three different objects that should not be
confused:
\begin{enumerate}[label=(\roman*)]
  \item the exact entropic teacher problem \eqref{eq:bf-eot-primal},
  \item the factorized teacher optimizer \eqref{eq:bf-eot-factorization},
  \item and the balancing equations \eqref{eq:bf-eot-a-balance}--\eqref{eq:bf-eot-b-balance}
  that a finite numerical routine will try to satisfy.
\end{enumerate}
Finite Sinkhorn iteration begins only after these exact relations have been fixed.

\section{Algorithm: finite Sinkhorn equal-weighting}
\label{sec:bf-eot-algorithm}

\begin{algorithm}[ht]
\caption{\texttt{sinkhorn\_equalweight\_step}}
\begin{algorithmic}[1]
\Require weighted particle cloud \(\{x_i,w_i\}_{i=1}^N\), equal target marginal \(u\), cost matrix \(C\), regularization \(\varepsilon>0\), initialization \((a^{(0)},b^{(0)})\), iteration budget \(K\), and stabilization policy
\Ensure finite coupling \(\Pi_{\varepsilon,K}\), barycentric equal-weight cloud \(\widetilde X\), row/column residuals, and declared numerical metadata
\State Form the Gibbs kernel \((K_\varepsilon)_{ij}=\exp(-C_{ij}/\varepsilon)\)
\For{\(k=0,\ldots,K-1\)}
  \State Update \(a^{(k+1)} = w \oslash (K_\varepsilon b^{(k)})\)
  \State Update \(b^{(k+1)} = u \oslash (K_\varepsilon' a^{(k+1)})\)
\EndFor
\State Form the finite coupling \(\Pi_{\varepsilon,K}=\diag(a^{(K)})K_\varepsilon\diag(b^{(K)})\)
\State Compute residuals
\[
  r_{\mathrm{row}}=\|\Pi_{\varepsilon,K}\mathbf 1-w\|_1,
  \qquad
  r_{\mathrm{col}}=\|\Pi_{\varepsilon,K}'\mathbf 1-u\|_1
\]
\For{each output slot \(j=1,\ldots,N\)}
  \State Compute \(\tilde x^{(j)} = \frac{1}{u_j}\sum_{i=1}^N \Pi_{ij}x_i\)
\EndFor
\State Return the coupling, equal-weight cloud, residuals, \(\varepsilon\), \(K\), initialization policy, and stabilization metadata
\end{algorithmic}
\end{algorithm}

The algorithm returns a \emph{finite} numerical object, not the exact teacher
optimizer from \eqref{eq:bf-eot-primal}.  After \(K\) balancing passes, the
executed coupling is
\begin{equation}
\label{eq:bf-eot-finite-coupling-repeat}
  \Pi_{\varepsilon,K}
  =
  \diag(a^{(K)})K_\varepsilon\diag(b^{(K)}),
\end{equation}
with residuals that measure how closely the declared row and column constraints
have been met.  This distinction is the second approximation layer of the chapter:
first the deterministic teacher was replaced by an entropic teacher, and then that
entropic teacher was approximated by a finite numerical route.  The next custom-
gradient chapter will differentiate this declared finite route, not an abstract
teacher floating above the executed computation.

\paragraph{Reader checkpoint.}
It is useful to keep the chain of objects explicit:
\begin{enumerate}[label=(\roman*)]
  \item deterministic OT teacher \(\Pi_0^\star\),
  \item exact entropic teacher \(\Pi_\varepsilon^\star\),
  \item finite executed coupling \(\Pi_{\varepsilon,K}\),
  \item barycentric equal-weight cloud built from \(\Pi_{\varepsilon,K}\).
\end{enumerate}
Later differentiation claims must attach to the actually executed object in this
chain, not jump back silently to one of the earlier teachers.

\subsection{Exact online and streaming Sinkhorn as the reference lane}
\label{sec:bf-eot-reference-lane}

The scalable-OT survey also clarifies how BayesFilter should talk about scaling
without silently changing the transport object.  A finite Sinkhorn route can be
made substantially more practical by changing the execution strategy rather than
the underlying entropic teacher problem.  Streaming, online, or GPU-specialized
implementations still belong to the reference lane when they preserve the same
entropic scaling equations and the same barycentric output rule while avoiding
full materialization of the cost or kernel matrix \citep{feydy2019geomloss,charlier2021keops,ye2026flashsinkhorn,xiao2026fastsinkhorn}.

In that reference-lane setting, the implementation may compute the needed
row-wise or column-wise reductions in blocks, keep only partial transport state
in memory, or fuse barycentric application with solver passes.  What must remain
fixed is the declared numerical object: the finite coupling
\(\Pi_{\varepsilon,K}\), its residual checks, and the barycentric output derived
from that same finite coupling.  Thus a streamed or GPU-tiled Sinkhorn routine is
not automatically a new approximation class.  It becomes a new approximation
class only when the scaling equations, kernel, feasible family, or output object
itself are changed.

This exact-reference reading is important because it separates engineering success
from scientific overclaim.  Exact-online or GPU-specialized Sinkhorn may reduce
memory traffic, improve device utilization, and make larger particle clouds
practical while still computing the same finite entropic object.  But these routes
normally leave the dominant all-pairs interaction structure in place: every full
row or column update still depends on interactions with the whole opposing cloud.
So they are the right first scaling lane when the goal is to preserve semantics,
but they are not by themselves the final answer to truly large-\(N\), large-\(d_x\)
transport.

This chapter therefore treats exact-online and streaming implementations as
engineering variants of the same entropic reference lane.  Later staged chapters
on retained-teacher approximations will take up the different case where the
transport object itself is compressed, factorized, or otherwise altered for
scalability.

\subsection{Sparse, screened, stabilized, and accelerated Sinkhorn variants}
\label{sec:bf-eot-solver-variant-boundary}

The scalable-OT survey also suggests a useful boundary inside the broader
Sinkhorn family.  Some variants still belong naturally to the exact-reference
lane because they mainly stabilize or schedule the same scaling equations more
carefully.  Others move toward approximation because they restrict which entries,
rows, or columns are treated as active, or because they replace dense kernel use
by a screened or sparse surrogate \citep{schmitzer2016sparse,alaya2019screenkhorn,schmitzer2019stabilized,altschuler2017nearlinear,dvurechensky2018computational,abid2018greedy,lin2019efficient,xu2026accelerating,genevay2016stochastic}.

At the most conservative end, stabilization policies remain part of the same
finite-Sinkhorn object.  Log-domain updates, shifting, epsilon scheduling, or
other numerical defenses are execution details attached to the same declared
coupling \(\Pi_{\varepsilon,K}\) and should therefore be judged first by the
usual marginal residual and barycentric parity checks.

Sparse or screened routes require a sharper audit.  If they only prune obviously
negligible work while preserving the same effective scaling equations and output
object to numerical tolerance, they may still be read as near-reference
engineering variants.  But once the active support, feasible family, or effective
kernel semantics are altered in a way that changes the finite transport object,
the method is no longer only a better implementation of the same reference lane.
It has crossed into retained-teacher or structure-changing territory and must be
judged accordingly.

A similar caution applies to accelerated, greedy, or stochastic update schedules.
These methods may reduce iteration count, repair the most violated row or column
first, or otherwise improve convergence behavior of the balancing routine.  That
is valuable, but it does not by itself remove dense-kernel access.  If each major
update still needs interactions against the whole opposing cloud, the dominant
large-\(N\), large-\(d_x\) arithmetic bottleneck remains.

This is why the present chapter keeps these variants conceptually downstream from
the finite teacher object rather than promoting them to new transport doctrines.
They refine how the reference route is executed.  They do not automatically change
what object the filter is claiming to compute.

\section{Worked balancing pass on the running cell}
\label{sec:bf-eot-worked-balance}

Take the simplest initialization \(b^{(0)}=(1,1,1)'\).  Then the first
row-balancing step computes
\begin{equation}
\label{eq:bf-eot-first-row-balance}
  a^{(1)} = w \oslash (K_1 b^{(0)}).
\end{equation}
For the running cell,
\[
  K_1 b^{(0)}
  \approx
  (1.0183,\,1.0184,\,1.0001),
\]
so
\[
  a^{(1)}
  \approx
  (0.491,\,0.327,\,0.167).
\]
The next column-balancing step then uses these row-repaired values to push the
column sums toward \((1/3,1/3,1/3)\).  The exact arithmetic is less important
than the interpretation: finite Sinkhorn repeatedly repairs whichever marginal is
currently wrong.

\paragraph{Plain-language takeaway.}
Sinkhorn is not a mysterious OT oracle.  It is an alternating balancing procedure
for a smoothed transport matrix.  That is why later learned warm-start methods try
to predict better initial balancing coordinates rather than replacing the solver
outright.

\section{Barycentric projection as the output map}
\label{sec:bf-eot-barycentric}

The solver output is still only a coupling matrix.  The equal-weight particle cloud
that later code consumes is obtained by barycentric projection:
\begin{equation}
\label{eq:bf-eot-barycentric}
  \tilde x^{(j)}
  =
  \frac{1}{u_j}\sum_{i=1}^N \Pi_{ij}x_i.
\end{equation}
In the equal-weight case, this is
\begin{equation}
\label{eq:bf-eot-barycentric-equal}
  \tilde x^{(j)}=N\sum_{i=1}^N \Pi_{ij}x_i.
\end{equation}
This step is where the numerical coupling becomes the actual equal-weight output
cloud.  It is therefore the place where an implementation must be especially clear
about shapes, normalization, and any auxiliary state that is or is not carried
alongside the particle locations.

\begin{proposition}[Barycentric equal-weighting contracts covariance]
\label{prop:bf-eot-barycentric-covariance-contraction}
Let \(X\) be a random variable that takes values \(x_1,\ldots,x_N\in\mathbb R^d\)
with probabilities \(w_1,\ldots,w_N\).  Let \(J\) be an equal-weight output index,
\(\mathbb P(J=j)=1/N\), and suppose a coupling between \(X\) and \(J\) has the
same source and target marginals as the OT coupling used in
\eqref{eq:bf-eot-barycentric-equal}.  The barycentric output associated with
index \(J\) is
\[
  \widetilde X = \mathbb E[X\mid J].
\]
Then
\begin{equation}
\label{eq:bf-eot-barycentric-mean-preservation}
  \mathbb E[\widetilde X]=\mathbb E[X],
\end{equation}
but
\begin{equation}
\label{eq:bf-eot-barycentric-covariance-contraction}
  \operatorname{Cov}(\widetilde X)
  =
  \operatorname{Cov}(X)
  -
  \mathbb E\!\left[\operatorname{Cov}(X\mid J)\right]
  \preceq
  \operatorname{Cov}(X).
\end{equation}
Equality holds only when the conditional covariance
\(\operatorname{Cov}(X\mid J)\) is zero almost surely.
\end{proposition}

\begin{proof}
The mean identity follows from the tower property:
\[
  \mathbb E[\widetilde X]
  =
  \mathbb E[\mathbb E[X\mid J]]
  =
  \mathbb E[X].
\]
For the covariance, apply the law of total covariance:
\[
  \operatorname{Cov}(X)
  =
  \operatorname{Cov}\!\left(\mathbb E[X\mid J]\right)
  +
  \mathbb E\!\left[\operatorname{Cov}(X\mid J)\right].
\]
Since \(\widetilde X=\mathbb E[X\mid J]\), rearranging gives
\eqref{eq:bf-eot-barycentric-covariance-contraction}.  The second term is
positive semidefinite, so the barycentric covariance is no larger in Loewner
order.  Equality requires the conditional covariance term to vanish almost
surely.
\end{proof}

\paragraph{Implication for filtering values.}
This proposition is the finite-\(N\) reason that a barycentric equal-weight cloud
should not be treated as a categorical-resampling cloud.  Categorical resampling
draws particles from the weighted empirical law and is unbiased for test
functions conditional on the weighted cloud.  Barycentric projection instead
replaces each conditional cloud by its conditional mean.  It preserves affine
test functions, including the first moment, but it generally changes quadratic
and nonlinear test functions.  In a linear-Gaussian state-space model, the next
predictive likelihood increment depends on the predictive covariance through
\(S_t=H P_{t\mid t-1}H'+R\).  A covariance-contracted reset can therefore change
the finite-\(N\) scalar being estimated even when Sinkhorn marginal residuals are
small, except in special cases where the lost covariance lies in observation-null
directions or otherwise cancels from the downstream likelihood increment.  The
exact Kalman likelihood remains a valid comparator for a no-OT weighted particle
approximation, but it is not a direct oracle for the barycentric-reset scalar
unless the reset route also preserves the relevant moments or the comparison is
explicitly redefined.

\paragraph{Implication for SV, SIR, and nonlinear models.}
The same issue is not special to LGSSM.  In stochastic-volatility models, future
observation densities are nonlinear functions of latent volatility, so the
finite-\(N\) likelihood scalar can change when barycentric projection compresses
the spread of the volatility particles.  In SIR and other nonlinear compartmental
models, the transition and observation maps are nonlinear in the state, so
preserving only the particle mean is not enough to preserve infection tails,
state constraints, threshold behavior, or future likelihood curvature.  A value
or gradient computed through the barycentric-reset filter can still be a correct
gradient of the executed barycentric scalar; it is not automatically the gradient
of the categorical-resampling particle-filter likelihood, the exact filtering
likelihood, or a moment-preserving approximation.  Consequently, SIR or SV
gradient checks must be same-scalar checks unless an additional evidence gate
shows that barycentric bias is negligible for the declared model, particle count,
regularization, and horizon.

The distinction matters for the later custom-gradient chapter
(Chapter~\ref{ch:bf-ledh-pfpf-ot-custom-gradient}).  That chapter does not
differentiate an abstract teacher floating outside the executed computation.  It
must differentiate the declared finite Sinkhorn route together with the barycentric
map that actually produces the equal-weight cloud used by the filtering scalar.

\section{Remedy options for barycentric covariance loss}
\label{sec:bf-eot-barycentric-remedies}

The remedy has to be stated at the level of the reset map, because the
covariance defect is not a vague numerical pathology of Sinkhorn.  It is a
deterministic consequence of replacing each column of a transport plan by its
conditional mean.  This subsection makes that statement as a finite-dimensional
matrix calculation.

For the algebra in this subsection, assume first that the Sinkhorn marginal
constraints have been solved exactly.  Finite iteration and floating-point
residuals add normalization errors that should be checked separately, but they
are not the mechanism discussed here.  Let
\[
  X = [x_1,\ldots,x_N]\in\mathbb R^{d_x\times N},
  \qquad
  w\in\mathbb R^N_+,\qquad
  \mathbf 1^\top w=1,
  \qquad
  W=\diag(w).
\]
The weighted empirical measure before reset is
\[
  \nu_w=\sum_{i=1}^N w_i\delta_{x_i}.
\]
Any deterministic equal-weight reset whose output particles are linear
combinations of the input particles can be written as
\begin{equation}
\label{eq:bf-eot-reset-transform}
  Y = XD,
  \qquad
  y_j=\sum_{i=1}^N x_i D_{ij},
\end{equation}
for some \(N\times N\) transform matrix \(D\).  The barycentric Sinkhorn reset is
the special case
\begin{equation}
\label{eq:bf-eot-barycentric-D}
  D_B = N\Pi_{\varepsilon,K},
\end{equation}
because the target marginal is \(u_j=1/N\).  The row and column marginal
constraints on \(\Pi_{\varepsilon,K}\) become the first-order transform
constraints
\begin{equation}
\label{eq:bf-eot-first-order-transform-contract}
  D^\top \mathbf 1 = \mathbf 1,
  \qquad
  D\mathbf 1 = Nw.
\end{equation}
The first condition says that every output particle is an affine combination of
the input particles.  The second says that the equal-weight output cloud has the
same mean as the weighted input cloud:
\begin{equation}
\label{eq:bf-eot-transform-mean}
  \bar y_D
  :=
  \frac1N Y\mathbf 1
  =
  \frac1N XD\mathbf 1
  =
  Xw
  =:
  \mu_w.
\end{equation}

The second moment is where the problem lives.  Define the weighted input
covariance and the equal-weight output covariance by
\begin{align}
\label{eq:bf-eot-weighted-covariance}
  \Sigma_w
  &:=
  \sum_{i=1}^N w_i(x_i-\mu_w)(x_i-\mu_w)^\top
  =
  X(W-ww^\top)X^\top,\\
\label{eq:bf-eot-transform-covariance}
  \Sigma_D
  &:=
  \frac1N\sum_{j=1}^N (y_j-\mu_w)(y_j-\mu_w)^\top
  =
  \frac1N X(D-w\mathbf 1^\top)(D-w\mathbf 1^\top)^\top X^\top.
\end{align}
A deterministic reset is therefore \emph{second-order exact for the current
cloud} when it satisfies \eqref{eq:bf-eot-first-order-transform-contract} and
\begin{equation}
\label{eq:bf-eot-second-order-cloud-contract}
  \Sigma_D=\Sigma_w.
\end{equation}
A stronger, cloud-independent sufficient condition is
\begin{equation}
\label{eq:bf-eot-second-order-matrix-contract}
  (D-w\mathbf 1^\top)(D-w\mathbf 1^\top)^\top
  =
  N(W-ww^\top).
\end{equation}
Equations
\eqref{eq:bf-eot-first-order-transform-contract}--\eqref{eq:bf-eot-second-order-matrix-contract}
are the finite-dimensional version of the second-order accurate ensemble-transform
condition used by Acevedo, de Wiljes, and Reich: the output ensemble matches both
the importance-weighted mean and covariance; see their Definition~3.1 and
Sections~5--6 for the Sinkhorn-based second-order ETPF construction
\citep{acevedodewiljesreich2017secondorder}.
The factor \(N\) in \eqref{eq:bf-eot-second-order-matrix-contract} is not a
convention-free decoration: it follows from using the biased equal-weight output
covariance \(N^{-1}\sum_j(y_j-\mu_w)(y_j-\mu_w)^\top\) together with a transform
matrix satisfying \(D\mathbf 1=Nw\).

\begin{proposition}[Why the barycentric Sinkhorn reset is not second-order exact]
\label{prop:bf-eot-barycentric-not-second-order}
Assume \(D_B=N\Pi_{\varepsilon,K}\) is nonnegative and satisfies
\eqref{eq:bf-eot-first-order-transform-contract}.  Interpret \(J\) as a uniform
output index and define a source index \(I\) by
\[
  \mathbb P(I=i\mid J=j)=D_{B,ij}.
\]
Then \(X_I\) has the weighted empirical law \(w\), the barycentric output is
\(Y_J=\mathbb E[X_I\mid J]\), and
\begin{equation}
\label{eq:bf-eot-D-covariance-gap}
  \Sigma_w-\Sigma_{D_B}
  =
  \mathbb E\!\left[\operatorname{Cov}(X_I\mid J)\right]
  \succeq 0.
\end{equation}
Consequently \(D_B\) satisfies the second-order cloud contract
\eqref{eq:bf-eot-second-order-cloud-contract} only if the conditional covariance
inside every output column vanishes as a state-space covariance.  Equivalently,
for each output column \(j\), all source particles with \(D_{B,ij}>0\) must have
the same state value, up to duplicate particle locations.  In generic diffuse
entropic couplings this condition fails.
\end{proposition}

\begin{proof}
The column sums in \eqref{eq:bf-eot-first-order-transform-contract} make
\(D_{B,\cdot j}\) a probability vector.  Since
\((1/N)D_B\mathbf 1=w\), the unconditional law of \(I\) is \(w\).  The output
particle in column \(j\) is exactly
\[
  y_j=\sum_i x_iD_{B,ij}=\mathbb E[X_I\mid J=j].
\]
Applying the law of total covariance gives
\[
  \operatorname{Cov}(X_I)
  =
  \operatorname{Cov}(\mathbb E[X_I\mid J])
  +
  \mathbb E[\operatorname{Cov}(X_I\mid J)].
\]
Because \(J\) is uniform,
\[
  \operatorname{Cov}(\mathbb E[X_I\mid J])
  =
  \frac1N\sum_{j=1}^N (y_j-\mu_w)(y_j-\mu_w)^\top
  =
  \Sigma_{D_B}.
\]
The left side is \(\Sigma_w\), which proves \eqref{eq:bf-eot-D-covariance-gap}.
\end{proof}

This is why the bug is not solved by asking Sinkhorn to converge more tightly.
A smaller marginal residual makes \(D_B\) a better approximation to the declared
entropic barycentric transform.  It does not change the identity
\eqref{eq:bf-eot-D-covariance-gap}, because that identity is about the output
map \(D_B\mapsto XD_B\), not about whether the coupling has been balanced
accurately.  A real remedy must therefore change, or explicitly keep, the
mathematical object being estimated.  The following contracts are mutually
different finite-\(N\) estimators.

\paragraph{Contract A: keep the same barycentric scalar.}
Set \(D=D_B\) and accept that \(\Sigma_D\preceq\Sigma_w\).  The reported value and
gradient are then only claims about the finite barycentric-reset scalar.  The
correct derivative check is a same-scalar finite-difference check of that exact
executed computation.  This contract repairs value-gradient naming, not the
covariance loss.  In notation, if the filter returns \(L_B(\theta)\), where every
reset uses \(Y_B(\theta)=X(\theta)D_B(\theta)\), then the admissible gradient
claim is
\[
  \nabla_\theta L_B(\theta),
\]
not the gradient of the categorical particle-filter likelihood and not the
gradient of an exact Kalman or nonlinear filtering likelihood.

\paragraph{Contract B: stochastic categorical output.}
Given the same nonnegative \(D_B\), draw \(I_j\) independently or by a chosen
resampling scheme from
\(\mathbb P(I_j=i\mid J=j)=D_{B,ij}\) and output \(z_j=x_{I_j}\).  Then for any
test function \(\varphi\) on the particle support,
\begin{equation}
\label{eq:bf-eot-stochastic-unbiased-test}
  \mathbb E\!\left[
    \frac1N\sum_{j=1}^N\varphi(z_j)\,\middle|\,D_B,X
  \right]
  =
  \sum_{i=1}^N w_i\varphi(x_i).
\end{equation}
This restores the unbiased empirical-measure property of categorical resampling
in expectation
\citep{gordon1993novel,doucet2001sequential,chopin2020introduction}, but the
realized reset is random and discontinuous as a function of weights and particles.
It is therefore a different estimator and not a pathwise differentiable
replacement for \(D_B\).

\paragraph{Contract C: second-order deterministic transform.}
Replace \(D_B\) by
\begin{equation}
\label{eq:bf-eot-second-order-correction}
  D^\sharp = D_B+\Delta,
  \qquad
  \Delta^\top\mathbf 1=0,
  \qquad
  \Delta\mathbf 1=0,
\end{equation}
and choose \(\Delta\) so that
\begin{equation}
\label{eq:bf-eot-second-order-correction-cov}
  \frac1N X(D^\sharp-w\mathbf 1^\top)
  (D^\sharp-w\mathbf 1^\top)^\top X^\top
  =
  X(W-ww^\top)X^\top.
\end{equation}
The stronger matrix-level target is
\begin{equation}
\label{eq:bf-eot-second-order-correction-matrix}
  (D^\sharp-w\mathbf 1^\top)(D^\sharp-w\mathbf 1^\top)^\top
  =
  N(W-ww^\top).
\end{equation}
One concrete reference object satisfying the matrix target is the NETF-type
square-root transform
\begin{equation}
\label{eq:bf-eot-netf-reference-transform}
  D_{\mathrm{NETF}}(Q)
  =
  w\mathbf 1^\top
  +
  \sqrt{N}\,(W-ww^\top)^{1/2}Q,
  \qquad
  Q\in\mathbb R^{N\times N},
  \qquad
  Q^\top Q=QQ^\top=I_N,
  \qquad
  Q\mathbf 1=\mathbf 1.
\end{equation}
Here \((W-ww^\top)^{1/2}\) denotes the principal symmetric square root.  Indeed,
\((W-ww^\top)\mathbf 1=0\), so the square root also annihilates \(\mathbf 1\), and
\(D_{\mathrm{NETF}}(Q)^\top\mathbf 1=\mathbf 1\),
\(D_{\mathrm{NETF}}(Q)\mathbf 1=Nw\), and
\eqref{eq:bf-eot-second-order-matrix-contract} holds:
\[
  (D_{\mathrm{NETF}}-w\mathbf 1^\top)
  (D_{\mathrm{NETF}}-w\mathbf 1^\top)^\top
  =
  N(W-ww^\top)^{1/2}QQ^\top(W-ww^\top)^{1/2}
  =
  N(W-ww^\top).
\]
Thus Contract C is mathematically explicit: the filter is no longer using the
barycentric transform \(D_B\); it is using a new transform \(D^\sharp\) whose
mean and covariance are the declared finite-\(N\) moments.  Inside this contract,
the NETF-type transform above and a second-order corrected ETPF are distinct
finite-\(N\) choices.  The NETF-type transform is a reference moment-preserving
object, not a promise of closeness to \(D_B\).  A second-order corrected ETPF
instead tries to stay close to the OT transform while enforcing the same moment
equations
\citep{reich2013nonparametric,acevedodewiljesreich2017secondorder}.
The unavoidable cost is that \(D^\sharp\) need not be nonnegative.  Hence the
corrected particles need not lie in the convex hull of the pre-reset particles,
and bounded, positive, simplex-valued, or otherwise constrained state spaces need
an explicit constraint policy.

\paragraph{Contract D: affine state-space moment restoration.}
Instead of correcting the \(N\times N\) transform matrix, one may correct the
already formed barycentric cloud.  Let \(Y_B=XD_B\) and
\(\Sigma_B=\Sigma_{D_B}\).  If there exists a matrix \(A_m\in\mathbb R^{d_x\times d_x}\)
such that
\begin{equation}
\label{eq:bf-eot-affine-moment-map}
  A_m\Sigma_B A_m^\top = \Sigma_w,
\end{equation}
then
\begin{equation}
\label{eq:bf-eot-affine-moment-output}
  y_j^\sharp
  =
  \mu_w + A_m(y_j-\mu_w),
  \qquad j=1,\ldots,N,
\end{equation}
has equal-weight mean \(\mu_w\) and covariance \(\Sigma_w\).  A full-rank example
is \(A_m=\Sigma_w^{1/2}\Sigma_B^{-1/2}\).  If \(\Sigma_B\) is singular, exact
affine restoration can exist only if
\(\operatorname{rank}(\Sigma_w)\le \operatorname{rank}(\Sigma_B)\);
jitter, truncation, or pseudoinverses define an additional approximation policy,
not an identity.  This contract is simple to test in an LGSSM, but it is a new
finite target and requires its own value gate, same-scalar gradient gate, and
state-constraint audit.

\paragraph{Contract E: positive spread, residual noise, and affine restoration.}
The preceding contracts leave a gap between two desirable but incompatible
properties.  Contract C gives an exact second-order deterministic transform, but
the transform can be signed.  Contract D gives a simple state-space covariance
repair, but it can be ill-conditioned when the barycentric cloud has already
collapsed in one or more directions.  A natural intermediate construction is to
split the job into three explicit stages:
\[
  D_B \longrightarrow D^+
  \longrightarrow \widetilde Y
  \longrightarrow Y^\star .
\]
Here \(D^+\) is still a nonnegative first-order transform, \(\widetilde Y\)
adds a small reparameterized residual in the missing covariance directions, and
\(Y^\star\) applies the final affine moment restoration.  This is not a
drop-in form of the second-order ETPF of
\citet{acevedodewiljesreich2017secondorder}, nor is it the nonlinear
transport-map framework of \citet{spantini2022coupling}.  It is a new finite
reset contract assembled from familiar components: positive ensemble transforms
\citep{reich2013nonparametric,corenflos2021differentiable}, second-order and
moment-matching ensemble updates
\citep{lei2011moment,todter2015secondorder,kirchgessner2017smoother,acevedodewiljesreich2017secondorder},
stochastic ensemble perturbations \citep{vanleeuwen2020consistent}, nonlinear
and conditional transport-map viewpoints
\citep{taghvaeihosseini2022bayesot,spantini2022coupling,hosseini2025conditional,zeng2026coupling},
learned measure-to-measure ensemble maps \citep{bach2026learning}, and a final
square-root covariance map.  The exact three-stage composition below was not
found in the inspected sources.  The citations support the ingredients and the
nearby design space; they do not prove that this contract is stable, unbiased,
or preferable for LEDH-PFPF-OT.  Those are separate value, gradient,
conditioning, and constraint gates.

The construction is easiest to state on the linear support of the weighted
cloud.  Write
\[
  \mathcal S_w:=\operatorname{range}(\Sigma_w),
  \qquad
  P_w:=\hbox{the orthogonal projector onto }\mathcal S_w .
\]
Let \(r_w:=\operatorname{rank}(\Sigma_w)\).  Since \(\Sigma_w\) is the covariance
of \(N\) centered weighted particles, \(r_w\le N-1\); if a modified construction
asks for a rank larger than \(N-1\), no equal-weight cloud of \(N\) particles can
realize it.  All square roots below are principal square roots on their stated
supports.  If
the rank of \(\Sigma_w\), \(G_+\), or \(\widetilde\Sigma\) changes along a
parameter path, the corresponding square-root, pseudoinverse, or support
projector derivative may fail to be classical.  A differentiable implementation
therefore needs either a fixed-rank contract or an explicit spectral floor and
smoothed spectral projector policy.  This is a numerical assumption, not a
consequence of optimal transport.

The first stage chooses
\begin{equation}
\label{eq:bf-eot-positive-spread-transform}
  D^+\in\mathcal D_1^+
  :=
  \{D\in\mathbb R_+^{N\times N}:D^\top\mathbf 1=\mathbf 1,\,
  D\mathbf 1=Nw\}.
\end{equation}
The notation \(D^+\) does not mean that the matrix is invertible.  The relevant
rank condition is a state-space condition on the cloud
\[
  Y^+ := XD^+,
  \qquad
  \Sigma_+
  :=
  \frac1N
  (Y^+-\mu_w\mathbf 1^\top)(Y^+-\mu_w\mathbf 1^\top)^\top .
\]
The selection rule for \(D^+\) may be a sharper Sinkhorn plan, a spread-penalized
positive transport solve, or a positive projection of a partial second-order
correction.  The mathematical contract is weaker and more auditable than any
particular optimizer: \(D^+\) must remain in \(\mathcal D_1^+\), must be close
enough to the declared transport target for the application, and must pass a
spectral gate on \(\Sigma_+\).  If the gate fails, this contract has not produced
a usable reset.

\begin{proposition}[The residual covariance to be repaired is positive]
\label{prop:bf-eot-positive-spread-gap}
Let \(D^+\in\mathcal D_1^+\).  Define
\[
  G_+ := \Sigma_w-\Sigma_+ .
\]
Then
\begin{equation}
\label{eq:bf-eot-positive-spread-gap-formula}
  G_+
  =
  \frac1N
  \sum_{j=1}^N
  \sum_{i=1}^N
  D^+_{ij}
  (x_i-y^+_j)(x_i-y^+_j)^\top
  \succeq 0.
\end{equation}
Equality holds if and only if each output column of \(D^+\) has zero conditional
state covariance, in the same sense as
Proposition~\ref{prop:bf-eot-barycentric-not-second-order}.  Moreover
\(\operatorname{range}(G_+)\subseteq\operatorname{range}(\Sigma_w)\).
\end{proposition}

\begin{proof}
The proof is the same total-covariance argument as
Proposition~\ref{prop:bf-eot-barycentric-not-second-order}, with \(D_B\)
replaced by \(D^+\).  Nonnegativity and the column-sum constraint make each
column \(D^+_{\cdot j}\) a conditional probability vector.  The row-sum
constraint gives the unconditional law \(w\).  Therefore the covariance of the
weighted source draw decomposes into the covariance of the column barycentres
plus the mean conditional covariance inside the columns.  The first term is
\(\Sigma_+\), and the second is exactly the right-hand side of
\eqref{eq:bf-eot-positive-spread-gap-formula}.  This proves positive
semidefiniteness.  Since \(0\preceq G_+\preceq \Sigma_w\), every vector in
\(\ker(\Sigma_w)\) is also in \(\ker(G_+)\); for symmetric positive
semidefinite matrices this is equivalent to
\(\operatorname{range}(G_+)\subseteq\operatorname{range}(\Sigma_w)\).
\end{proof}

The second stage adds a differentiable residual in the directions represented by
\(G_+\).  Let \(\Xi\in\mathbb R^{d_x\times N}\) be an auxiliary base-noise
matrix, independent of the model parameters, satisfying
\[
  \mathbb E[\Xi]=0,
  \qquad
  \Xi\mathbf 1=0,
  \qquad
  \mathbb E\!\left[\frac1N\Xi\Xi^\top\right]=I_{d_x}.
\]
One concrete reparameterized choice is
\[
  H_N:=I_N-\frac1N\mathbf 1\mathbf 1^\top,
  \qquad
  \Xi:=\sqrt{\frac{N}{N-1}}\, ZH_N,
  \qquad
  Z_{ij}\stackrel{\mathrm{iid}}{\sim}N(0,1),
  \qquad N\ge2.
\]
The matrix \(Z\) is drawn independently of \(X\) and \(D^+\).  Then
\(\Xi\mathbf 1=0\), and
\(\mathbb E[N^{-1}\Xi\Xi^\top]=I_{d_x}\) because \(H_N\) is an idempotent
projector with trace \(N-1\).  A fixed quasi-random or orthogonal design may be
used instead, but then its empirical covariance is part of the finite contract
and must be reported.  For \(0\le\rho\le1\) and \(\tau\ge0\), define
\begin{equation}
\label{eq:bf-eot-residual-noise-cloud}
  \widetilde Y
  :=
  Y^+
  +
  B_{\rho,\tau}\Xi,
  \qquad
  B_{\rho,\tau}
  :=
  \sqrt{\rho}\,(G_+ + \tau P_w)^{1/2}.
\end{equation}
The parameter \(\rho\) controls how much of the gap is injected before the
affine map.  The parameter \(\tau\) is not a proof device; it is a numerical
floor on \(\mathcal S_w\).  Since
\(\operatorname{range}(G_+)\subseteq\mathcal S_w\), the residual covariance
\(G_+ + \tau P_w\) is positive semidefinite and has no component orthogonal to
the weighted cloud support.  If \(\tau\) is set too small, the affine whitening
in the next stage can amplify the auxiliary noise violently.  Therefore
\(\tau\) and \(\rho\) must be chosen by conditioning gates, not by the desire to
make the perturbation invisible.

\begin{proposition}[Expected effect of the residual stage]
\label{prop:bf-eot-residual-expected-covariance}
Assume \(D^+\in\mathcal D_1^+\), \(\mathbb E[\Xi\mid X,D^+]=0\),
\(\Xi\mathbf 1=0\), and
\(\mathbb E[N^{-1}\Xi\Xi^\top\mid X,D^+]=I_{d_x}\).  Conditional on \(X\) and \(D^+\),
the residual cloud \(\widetilde Y\) in
\eqref{eq:bf-eot-residual-noise-cloud} has mean \(\mu_w\), and its covariance
around \(\mu_w\) satisfies
\begin{equation}
\label{eq:bf-eot-residual-expected-covariance}
  \mathbb E\!\left[
    \frac1N(\widetilde Y-\mu_w\mathbf 1^\top)
    (\widetilde Y-\mu_w\mathbf 1^\top)^\top
    \,\middle|\,X,D^+
  \right]
  =
  \Sigma_+ + \rho(G_+ + \tau P_w).
\end{equation}
In particular, when \(\rho=1\) and \(\tau=0\), the conditional expected
covariance is \(\Sigma_w\).
\end{proposition}

\begin{proof}
Because \(D^+\mathbf 1=Nw\), the equal-weight mean of \(Y^+\) is \(\mu_w\).
Because \(\Xi\mathbf 1=0\), the equal-weight mean of \(B_{\rho,\tau}\Xi\) is
zero.  Expanding the covariance of \(\widetilde Y\) around \(\mu_w\) gives the
covariance of \(Y^+\), two cross terms, and the residual covariance.  The cross
terms have conditional expectation zero because the auxiliary noise has
conditional mean zero given \(X,D^+\).  The residual covariance has conditional
expectation
\[
  B_{\rho,\tau}
  \mathbb E\!\left[\frac1N\Xi\Xi^\top\,\middle|\,X,D^+\right]
  B_{\rho,\tau}^\top
  =
  B_{\rho,\tau}B_{\rho,\tau}^\top
  =
  \rho(G_+ + \tau P_w),
\]
by the definition of the principal square root.  This proves
\eqref{eq:bf-eot-residual-expected-covariance}.
\end{proof}

The proposition is deliberately an expectation statement.  For a realized finite
noise matrix \(\Xi\), the covariance of \(\widetilde Y\) need not equal
\(\Sigma_w\) even when \(\rho=1\) and \(\tau=0\).  This is not a defect in the
algebra; it is the price of using a reparameterized stochastic residual rather
than a discontinuous categorical resampling step.  The next affine stage turns
the realized cloud into an exact finite second-moment object only when the
support condition in Proposition~\ref{prop:bf-eot-residual-affine-restoration}
holds and the exact Moore--Penrose inverse square root is used.

\begin{proposition}[A small residual can repair support, but not sample-rank limits]
\label{prop:bf-eot-residual-support-repair}
Assume \(r_w\le N-1\), \(0<\rho\le1\), and \(\tau>0\).  Draw \(Z\) with
entries having a joint density, for example independent standard Gaussian
entries, and set
\(\Xi=\sqrt{N/(N-1)}\,ZH_N\).  Conditional on \(X\) and \(D^+\), the residual
matrix \(B_{\rho,\tau}\Xi\) has column span equal to \(\mathcal S_w\) with
probability one.  Consequently
\[
  \operatorname{range}(\Sigma_w)
  \subseteq
  \operatorname{range}(\widetilde\Sigma)
\]
with probability one.  If \(r_w>N-1\), no choice of \(N\) equal-weight output
particles can realize covariance \(\Sigma_w\), because every centered
\(d_x\times N\) cloud has rank at most \(N-1\).
\end{proposition}

\begin{proof}
Since \(\tau>0\),
\[
  G_+ + \tau P_w
\]
is strictly positive definite on \(\mathcal S_w\) and zero on
\(\mathcal S_w^\perp\).  Hence \(B_{\rho,\tau}\) has range \(\mathcal S_w\) and
rank \(r_w\).  Let \(U\in\mathbb R^{d_x\times r_w}\) have orthonormal columns
spanning \(\mathcal S_w\).  The centering matrix \(H_N\) has rank \(N-1\), and
\(U^\top ZH_N\) has rank \(r_w\) almost surely when \(r_w\le N-1\), because its
\(r_w\times (N-1)\) coordinate representation on
\(\operatorname{range}(H_N)\) has a joint density and at least one
\(r_w\times r_w\) minor is a nonzero polynomial in those variables.
Multiplication by the nonsingular
restriction of \(B_{\rho,\tau}\) on \(\mathcal S_w\) therefore produces a
residual cloud whose span is exactly \(\mathcal S_w\) almost surely.

The deterministic centered cloud \(Y^+-\mu_w\mathbf 1^\top\) also lies in
\(\mathcal S_w\).  The event that adding this fixed matrix to the full-rank
random residual reduces the rank on \(\mathcal S_w\) is again a polynomial zero
event.  To see that the relevant polynomial is not identically zero, choose any
centered matrix \(R_0\) with column span \(\mathcal S_w\); then the same
\(r_w\times r_w\) minor of
\((Y^+-\mu_w\mathbf 1^\top)+tR_0\) is nonzero for all sufficiently large
\(|t|\).  Hence the zero set has probability zero under the joint-density
assumption.  Thus the realized centered cloud
\(\widetilde Y-\mu_w\mathbf 1^\top\) has rank \(r_w\) on \(\mathcal S_w\) almost
surely, which is the displayed support inclusion.  The final statement follows
because any centered cloud \(Y-\bar y\mathbf 1^\top\) has columns summing to zero
and therefore rank at most \(N-1\).
\end{proof}

The third stage removes the remaining finite-sample mismatch by affine
restoration.  Define the realized residual covariance
\[
  \widetilde\Sigma
  :=
  \frac1N
  (\widetilde Y-\mu_w\mathbf 1^\top)
  (\widetilde Y-\mu_w\mathbf 1^\top)^\top .
\]
If \(\widetilde\Sigma\) is nonsingular on the support of \(\Sigma_w\), set
\begin{equation}
\label{eq:bf-eot-residual-affine-map}
  A_{\rho,\tau}
  :=
  \Sigma_w^{1/2}\,\widetilde\Sigma^{\dagger,-1/2},
  \qquad
  Y^\star
  :=
  \mu_w\mathbf 1^\top
  +
  A_{\rho,\tau}
  (\widetilde Y-\mu_w\mathbf 1^\top),
\end{equation}
where \(\widetilde\Sigma^{\dagger,-1/2}\) denotes the Moore--Penrose inverse of
the principal square root.

\begin{proposition}[Affine restoration after the residual stage]
\label{prop:bf-eot-residual-affine-restoration}
Assume \(D^+\in\mathcal D_1^+\), \(\Xi\mathbf 1=0\), and
\[
  \operatorname{range}(\Sigma_w)
  \subseteq
  \operatorname{range}(\widetilde\Sigma).
\]
Then the output \(Y^\star\) in \eqref{eq:bf-eot-residual-affine-map} has
equal-weight mean \(\mu_w\) and equal-weight covariance \(\Sigma_w\).
\end{proposition}

\begin{proof}
The mean statement follows from \(D^+\mathbf 1=Nw\) and \(\Xi\mathbf 1=0\).
For the covariance, let \(P_{\widetilde\Sigma}\) be the orthogonal projector
onto \(\operatorname{range}(\widetilde\Sigma)\).  By the definition of the
Moore--Penrose inverse square root,
\[
  \widetilde\Sigma^{\dagger,-1/2}
  \widetilde\Sigma
  \widetilde\Sigma^{\dagger,-1/2}
  =
  P_{\widetilde\Sigma}.
\]
The range assumption implies that this projector acts as the identity on
\(\operatorname{range}(\Sigma_w)\).  Since the range of
\(\Sigma_w^{1/2}\) is \(\operatorname{range}(\Sigma_w)\), this also implies
\(P_{\widetilde\Sigma}\Sigma_w^{1/2}=\Sigma_w^{1/2}\).  By symmetry of
\(P_{\widetilde\Sigma}\) and \(\Sigma_w^{1/2}\), it also implies
\(\Sigma_w^{1/2}P_{\widetilde\Sigma}=\Sigma_w^{1/2}\).  Hence
\[
  \frac1N
  (Y^\star-\mu_w\mathbf 1^\top)
  (Y^\star-\mu_w\mathbf 1^\top)^\top
  =
  \Sigma_w^{1/2}
  P_{\widetilde\Sigma}
  \Sigma_w^{1/2}
  =
  \Sigma_w .
\]
\end{proof}

An implementation may replace \(\widetilde\Sigma^{\dagger,-1/2}\) by a floored
or otherwise regularized inverse square root for conditioning.  That defines an
approximate restoration policy, not Proposition~\ref{prop:bf-eot-residual-affine-restoration}.
The approximation must report its spectral floor, covariance residual, and
conditioning diagnostics separately from the exact algebra above.

Contract E is attractive because it does not ask the nonnegative transform to do
the impossible.  Positivity is used only to build a noncollapsed first-order
cloud; residual noise supplies controlled missing directions; the affine map
enforces the declared second moment.  It is also dangerous for exactly the same
reason.  The final particles need not remain in the convex hull of the source
particles after the residual and affine stages, and a small residual can be made
large by \(\widetilde\Sigma^{-1/2}\).  The contract is therefore admissible only
with explicit gates:
\[
  \operatorname{rank}_{\tau_{\mathrm{eig}}}(\widetilde\Sigma)=r_w
  \quad\hbox{or}\quad
  \|(I-P_{\widetilde\Sigma,\tau_{\mathrm{eig}}})P_w\|\le \epsilon_{\mathrm{supp}},
  \qquad
  \lambda_{\min}^{\tau}(\widetilde\Sigma)>\tau_{\mathrm{eig}},
  \qquad
  \kappa^{\tau}(\widetilde\Sigma)<\kappa_{\max},
  \qquad
  \frac{\|\Sigma_w-\Sigma_+\|}{\|\Sigma_w\|}\le c_{\mathrm{gap}}
  \quad\hbox{or an explained exception.}
\]
where \(P_{\widetilde\Sigma,\tau_{\mathrm{eig}}}\) denotes the numerical support
projector obtained from the same eigenvalue threshold used by the rank gate.
The spectral quantities
\(\lambda_{\min}^{\tau}(\widetilde\Sigma)\) and
\(\kappa^{\tau}(\widetilde\Sigma)\) are computed only on this retained
\(\tau_{\mathrm{eig}}\)-support.  Eigenvalues below the threshold are treated as
discarded numerical null directions and must be charged to the support-projector
gate above; they must not simultaneously veto the fallback through a full
positive-spectrum condition.  If the implementation claims the exact
Moore--Penrose restoration rather than a floored numerical restoration, then the
full positive spectrum must be used and the support-projector fallback is no
longer sufficient.
It must also record the auxiliary-noise law or fixed design, random seeds,
Monte Carlo standard errors, same-scalar finite-difference checks, and
constraint violations after the affine map.  In SIR, SV, and other constrained
models, the construction should normally be tested first in validated
unconstrained or transformed coordinates.

\paragraph{Contract E--Chol: ridged square-root gauge for differentiability.}
The support-exact form above is mathematically natural but a poor default
coordinate system for a manual gradient implementation.  It asks the code to
differentiate eigenvectors, support projectors, principal square roots, and
Moore--Penrose inverse square roots.  When eigenvalues repeat, eigenvectors are
not unique.  When an eigenvalue crosses the numerical floor, the retained rank
changes.  Thus an eigendecomposition implementation can be perfectly reasonable
for a forward diagnostic while still being a bad coordinate chart for a
gradient-bearing reset.  This is the same engineering lesson as the
square-root sigma-point discussion in Chapter~\ref{ch:bf-svd-sigma-point}: the
covariance object may be smooth even when a particular spectral coordinate
representation is not.  Classical numerical linear algebra also treats Cholesky
factorization as the stable coordinate chart for strictly positive definite
covariances; see, for example, \citet{higham2002accuracy}.

A practical differentiable variant is therefore to replace the support-exact
eigensystem route by a strictly positive definite, ridged Cholesky gauge.  This
is a different numerical contract, not a proof that the eigensystem contract was
unnecessary.  Choose user-visible ridge parameters
\(\lambda_{\rm abs}>0\) and \(\lambda_{\rm rel}\ge0\), and set
\begin{equation}
\label{eq:bf-eot-contract-e-chol-ridge}
  \lambda_0
  :=
  \max\!\left\{
    \lambda_{\rm abs},
    \lambda_{\rm rel}\frac{\operatorname{tr}(\Sigma_w)}{d_x}
  \right\}.
\end{equation}
The implementation should then use a bounded escalation policy
\(\lambda_k=\lambda_0\gamma^k\), \(\gamma>1\), stopping if the Cholesky factors
below are not finite or do not have positive diagonal entries after the allowed
attempts.  The actual \(\lambda_k\) used is part of the algorithm output.  It is
not harmless metadata: too small a ridge can make the triangular solve explode,
while too large a ridge changes the reset into a biased full-rank covariance
contract.

With \(\lambda=\lambda_k\), define
\[
  B_\lambda
  :=
  \sqrt{\rho}\,\operatorname{chol}(G_+ + \lambda I),
  \qquad
  \widetilde Y
  :=
  Y^+ + B_\lambda\Xi .
\]
Then form
\[
  \widetilde\Sigma
  :=
  \frac1N
  (\widetilde Y-\bar{\widetilde y}\mathbf 1^\top)
  (\widetilde Y-\bar{\widetilde y}\mathbf 1^\top)^\top ,
  \qquad
  L_w:=\operatorname{chol}(\Sigma_w+\lambda I),
  \qquad
  L_{\widetilde{}}:=\operatorname{chol}(\widetilde\Sigma+\lambda I).
\]
The affine operator is not formed with an inverse explicitly.  It is the
triangular-solve operator
\begin{equation}
\label{eq:bf-eot-contract-e-chol-affine}
  A_\lambda
  :=
  L_w L_{\widetilde{}}^{-1},
  \qquad
  Y^\star_\lambda
  :=
  \mu_w\mathbf 1^\top
  +
  A_\lambda(\widetilde Y-\bar{\widetilde y}\mathbf 1^\top).
\end{equation}
The recentering in \eqref{eq:bf-eot-contract-e-chol-affine} is essential: it is
what preserves the target mean independently of the ridge.

\begin{proposition}[Ridged Cholesky restoration]
\label{prop:bf-eot-contract-e-chol-restoration}
Assume the Cholesky factors in
\eqref{eq:bf-eot-contract-e-chol-affine} exist.  Then
\[
  A_\lambda(\widetilde\Sigma+\lambda I)A_\lambda^\top
  =
  \Sigma_w+\lambda I,
\]
and \(Y^\star_\lambda\) has equal-weight mean \(\mu_w\).  The covariance of
\(Y^\star_\lambda\) around \(\mu_w\) satisfies
\[
  \frac1N
  (Y^\star_\lambda-\mu_w\mathbf 1^\top)
  (Y^\star_\lambda-\mu_w\mathbf 1^\top)^\top
  -
  \Sigma_w
  =
  \lambda(I-A_\lambda A_\lambda^\top).
\]
\end{proposition}

\begin{proof}
The mean statement follows immediately from the centering in
\eqref{eq:bf-eot-contract-e-chol-affine}.  Since
\(\widetilde\Sigma+\lambda I=L_{\widetilde{}}L_{\widetilde{}}^\top\),
\[
  A_\lambda(\widetilde\Sigma+\lambda I)A_\lambda^\top
  =
  L_wL_{\widetilde{}}^{-1}
  L_{\widetilde{}}L_{\widetilde{}}^\top
  L_{\widetilde{}}^{-\top}L_w^\top
  =
  L_wL_w^\top
  =
  \Sigma_w+\lambda I.
\]
Subtracting
\(A_\lambda(\lambda I)A_\lambda^\top\) from both sides gives the displayed
unridged covariance residual.
\end{proof}

Contract E--Chol should be opt-in until its value and gradient behavior has
been tested beyond tiny reset fixtures.  Its advantage is that the manual VJP
can be written through Cholesky factors and triangular solves, without
eigenvector gauges or rank-switching projectors.  Its cost is the explicit
\(\lambda\)-bias above.  A user-facing function must therefore expose
\(\lambda_{\rm abs}\), \(\lambda_{\rm rel}\), escalation multiplier, and maximum
attempts; it must record the realized ridge and the covariance residual; and it
must warn that the ridge is a tuning parameter, not a universal constant.

\paragraph{Contract F: no reset or delayed reset.}
If the effective sample size is acceptable, carry the weighted cloud forward and
do not apply \(D_B\).  This avoids the deterministic covariance loss at that step.
It does not solve weight degeneracy in long horizons or high dimensions, but it is
a clean diagnostic: if a no-reset weighted route agrees with an exact LGSSM
comparator while the barycentric route fails, the reset contract rather than the
transition or likelihood code is implicated.

\begin{center}
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}p{0.20\textwidth} p{0.23\textwidth} p{0.22\textwidth} p{0.20\textwidth}@{}}
\toprule
Remedy & What it fixes & Main cost & Required audit \\
\midrule
Same-scalar barycentric contract & claim mismatch for the executed scalar & does not restore covariance & same-scalar finite differences and moment-drift report \\
Stochastic resampling & unbiased empirical-measure semantics in expectation & discontinuous random branch & estimator contract, seed/randomness policy, MCSE \\
Second-order transform & mean and covariance matching & possible negative transform entries and nonphysical states & moment gate, constraint gate, custom VJP gate \\
Affine moment restoration & local covariance restoration diagnostic & new finite target and rank/jitter choices & LGSSM value/gradient gate and state-constraint audit \\
Positive residual-affine hybrid & covariance restoration with positive first stage & novel randomized finite target and conditioning sensitivity & spectrum gate, MCSE, same-scalar gradient gate, constraint audit \\
No-reset threshold & avoids unnecessary deterministic collapse & leaves weight degeneracy unresolved & ESS policy and weighted-cloud comparator \\
\bottomrule
\end{tabular}
\setlength{\tabcolsep}{6pt}
\normalsize
\end{center}

\paragraph{Implication for SIR, SV, and learned teachers.}
Yes, this issue can affect SIR, stochastic-volatility, and other nonlinear-model
gradients.  The effect is harder to expose than in an LGSSM because there is no
closed-form Kalman likelihood oracle, but the mechanism is the same: a
mean-preserving reset can still change nonlinear transition, observation, and
future-likelihood functionals.  Formally, let \(\Phi_t\) denote the rest of the
filtering computation after a reset at time \(t\).  The barycentric route and a
moment-preserving route evaluate different functions,
\[
  L_B(\theta)=\Phi_t\!\left(X_t(\theta)D_B(\theta);\theta\right),
  \qquad
  L_\sharp(\theta)=
  \Phi_t\!\left(X_t(\theta)D^\sharp(\theta);\theta\right),
\]
unless \(\Phi_t\) is insensitive to the covariance directions removed by
\eqref{eq:bf-eot-D-covariance-gap}.  Therefore equality of autodiff and finite
differences for \(L_B\) certifies only the executed barycentric route.  It does
not certify \(L_\sharp\), a categorical-resampling route, or an exact filtering
likelihood.  A learned teacher trained to imitate the current barycentric teacher
\(D_B\) will inherit this moment contract rather than repair it.  A learned or
retained teacher can be a remedy only if its target is changed to one of the
declared objects above, for example \(D^\sharp\), \(D_{\mathrm{NETF}}\), the
stochastic estimator, or the affine-corrected cloud \(Y^\star\), and the
downstream filtering diagnostics are rerun under that new finite target.

\section{Differentiation contracts}
\label{sec:bf-eot-differentiation}

The final major distinction of the chapter is between three different derivative
objects:
\begin{enumerate}[label=(\roman*)]
  \item derivative of the exact regularized OT optimizer,
  \item derivative of the finite unrolled Sinkhorn computation,
  \item derivative of some scalar built from that finite computation.
\end{enumerate}
In code, the second and third are the most common.  A gradient is only the
correct gradient of the exact executed scalar if the computational graph used in
autodiff matches the scalar value actually reported by the code.  The next chapter
specializes this point to the declared LEDH-PFPF-OT scalar and its custom VJP.

\begin{proposition}[Stopped transport normalization gives a partial derivative]
\label{prop:bf-eot-stopped-normalization-partial}
Let a finite Sinkhorn transport step use normalized particles
\[
  u_i(x)={x_i-\bar x\over s(x)},\qquad
  \bar x={1\over N}\sum_{k=1}^N x_k,
\]
where \(s(x)\) is the scale used to build the transport cost.  Let
\[
  L(x)=\Psi\bigl(x,u(x)\bigr)
\]
be any downstream scalar depending on both the carried particles \(x\) and the
transport-normalized particles \(u(x)\).  If a reverse pass treats \(\bar x\)
and \(s(x)\) as constants, then it computes the partial derivative of
\(\Psi(x,u)\) with respect to \(x\) under the artificial map
\[
  u_i={x_i-c\over s},
\]
not the total derivative of \(L(x)\).  Therefore, unless the omitted terms are
zero or a different objective is explicitly declared, this stopped derivative is
wrong relative to the claim that it is the score of the executed finite scalar.
\end{proposition}

\begin{proof}
Write \(a_i=\partial L/\partial u_i\).  The true differential through
\[
  u_i(x)={x_i-\bar x\over s(x)}
\]
is
\[
  dL_u
  =
  \sum_i a_i^\top {dx_i-d\bar x\over s(x)}
  -
  \left(\sum_i a_i^\top (x_i-\bar x)\right){ds(x)\over s(x)^2}.
\]
Since \(d\bar x=N^{-1}\sum_k dx_k\), the contribution to the adjoint of
\(x_i\) is
\[
  {a_i\over s(x)}
  -
  {1\over N}\sum_k {a_k\over s(x)}
  -
  {1\over s(x)^2}
  \left(\sum_k a_k^\top (x_k-\bar x)\right)\nabla_{x_i}s(x).
\]
The stopped rule keeps only \(a_i/s(x)\).  It drops the mean term and the scale
term.  Hence it is the derivative of an artificial partial-derivative problem
with fixed \(c\) and fixed \(s\), not the total derivative of \(L(x)\).
\end{proof}

The same warning applies to Sinkhorn cost keys, adaptive starting
\(\varepsilon_0\), finite-iteration potentials, and any other quantity computed
from the current particles.  If the scalar value recomputes such a quantity
when \(\theta\) changes, then its score includes the corresponding chain-rule
term.  A hand-written VJP may use internal stops only to avoid double counting,
but it must add back every blocked derivative.  Otherwise the result is a
partial derivative and must be named as such.

\begin{center}
\small
\begin{tabular}{p{0.24\textwidth} p{0.30\textwidth} p{0.33\textwidth}}
\toprule
Derivative object & What is being differentiated? & What can go wrong? \\
\midrule
Exact regularized optimizer & the ideal entropic OT solution & not what finite code usually executes directly \\
Finite unrolled Sinkhorn route & the declared numerical balancing procedure & residuals or stabilization may change the executed object \\
Downstream scalar built from finite Sinkhorn & the actual reported value used in filtering or HMC-adjacent code & value-gradient mismatch if autodiff follows a different graph \\
\bottomrule
\end{tabular}
\end{center}

\section{Filtering-loop insertion point}
\label{sec:bf-eot-loop}

Inside a filter, the finite EOT/Sinkhorn equal-weighting step sits after the
weighted particle cloud has been formed and before an equally weighted cloud is
passed onward:
\begin{equation}
\label{eq:bf-eot-loop}
  \{x_i,w_i\}_{i=1}^N
  \xrightarrow{\mathrm{EOT/Sinkhorn}}
  \{\tilde x^{(j)},1/N\}_{j=1}^N.
\end{equation}
If the broader filtering algorithm carries auxiliary particle-local state, this
chapter does not yet assert how that state is transported.  Its contract is
narrower: it specifies the equal-weighting rule for the particle locations and the
mass-routing matrix from which that equal-weight cloud is derived.

\section{Verification contract}
\label{sec:bf-eot-verification}

A reader implementing this route should verify at least:
\begin{enumerate}[label=(\arabic*)]
  \item row and column residuals are reported and decrease with additional
  iterations or stronger stabilization,
  \item the initialization, \(\varepsilon\), iteration budget, and stabilization
  policy are recorded,
  \item the barycentric cloud has the declared shape and normalization,
  \item any reported gradient is tied to the exact computational graph that was
  actually executed,
  \item any stronger filtering or HMC claim is kept separate from these local
  numerical checks.
\end{enumerate}

A failed residual check is evidence about the finite solver or stabilization path,
not about the abstract EOT object.  A successful residual check is evidence that
the numerical route is behaving as declared, not that it preserves the original
categorical target.

\section{What this chapter does and does not claim}
\label{sec:bf-eot-boundary}

This chapter teaches the main differentiable OT computational route in a near-
implementation style.  The derivation above fixes a specific ladder:
\begin{enumerate}[label=(\roman*)]
  \item deterministic OT chooses an exact coupling for the unregularized baseline,
  \item entropic OT changes that teacher object,
  \item Sinkhorn balancing computes a finite numerical approximation to the entropic
    teacher,
  \item barycentric projection turns that finite coupling into the equal-weight
    cloud consumed by later filtering code.
\end{enumerate}
Within that ladder, exact-online and streamed execution strategies remain on the
reference lane when they preserve the same scaling equations and the same
barycentric output contract.  Later retained-teacher chapters ask whether the same
teacher can be compressed or approximated while still judged against this reference
object.  Still later changed-object chapters ask a different question again: what
happens when the transport object itself is replaced by a learned map, altered
support, or new target contract.

It does \emph{not} claim:
\begin{itemize}
  \item that entropic OT is the same object as unregularized OT,
  \item that finite Sinkhorn is the exact regularized optimizer,
  \item that a barycentric equal-weight cloud is the same object as categorical
  ancestor sampling,
  \item that autodiff through the executed graph is automatically the gradient of
    the original filtering likelihood.
\end{itemize}
Its role is to fix the entropic teacher object, the finite numerical route, the
output cloud construction, and the differentiation contract that the recovered
custom-gradient bridge and later learned-transport chapters build on.

\FloatBarrier
