\chapter{Retained-Teacher Neural OT Warm Starts}
\label{ch:bf-dpf-learned-ot}

The preceding chapters fixed the entropic OT and Sinkhorn layer, together with
its recovered custom-gradient bridge, as a fully declared teacher-side
computational object.  The present chapter introduces the narrowest neural
extension of that teacher: do not replace the teacher; learn how to start it
better.

This is the main implementation-bearing neural chapter of the OT block.  Its
question is narrower than ``can a network learn transport?''  It asks instead:
can a network predict a good internal solver state so that a retained corrective
Sinkhorn phase needs less work while still protecting the declared teacher
object?

This narrowing is deliberate.  The monograph begins neural OT with retained-
teacher acceleration because it is the least radical learned extension of the
existing OT block.  The weighted cloud, the entropic teacher, and the barycentric
output rule all remain declared in advance; only the route used to reach that
teacher object is learned or accelerated.

\section{What problem this chapter solves}
\label{sec:bf-learned-ot-objects}

Let \(X=(x_1,\ldots,x_N)^\top\in\mathbb R^{N\times d_x}\) be the weighted
source cloud and let \(w\in\Delta_N\) be its normalized weights.  Let
\(u=(1/N,\ldots,1/N)\) be the equal target marginal, and let
\(C(X,Z)\in\mathbb R^{N\times N}\) be the declared transport cost matrix.  The
previous chapter already fixed the finite entropic-OT/Sinkhorn route as the
reference computational teacher for turning this weighted cloud into an
equal-weight cloud.

The practical question now is not whether the teacher should be changed, but
whether its execution path can be improved.  The chapter therefore asks:
\begin{quote}
Can we learn a solver-relevant latent state that makes the retained corrective
Sinkhorn solve cheaper while still returning to the same declared finite teacher
object after correction?
\end{quote}

\paragraph{How to read the rest of this chapter.}
The chapter has one primary lane and one secondary lane.  The primary lane is the
warm-start story: define the teacher, define the latent state the student predicts,
run the corrective retained solver, and judge the corrected cloud against the
teacher.  Only after that core story is in place does the chapter widen into a
shorter survey of broader retained-teacher families such as Nystr"om,
positive-feature, and low-rank routes.

\begin{center}
\small
\begin{tabular}{p{0.27\textwidth} p{0.61\textwidth}}
\toprule
Object & Meaning in this chapter \\
\midrule
exact entropic teacher \(\Pi_\varepsilon^\star\) & optimizer of the declared regularized OT problem \\
finite numerical teacher \(\Pi_{\varepsilon,K}\) & entropic teacher after a declared finite number of Sinkhorn repairs \\
latent warm-start target \(h_\tau\) & solver-relevant coordinate such as duals, log-scalings, or scaling vectors \\
corrected student output \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\) & coupling produced after student initialization and retained corrective Sinkhorn steps \\
barycentric output \(B(\Pi)\) & equal-weight cloud later code actually consumes \\
\bottomrule
\end{tabular}
\end{center}

\section{Running warm-start cell}
\label{sec:bf-learned-ot-running-cell}

Keep the same three-particle setting as the previous OT chapters: a weighted
source cloud, an equal target marginal, a declared quadratic cost, a
regularization level \(\varepsilon\), and a finite Sinkhorn budget \(K\).  The
next section will restate the teacher formally; for now the important point is
that the retained finite teacher is the coupling
\(\Pi_{\varepsilon,K}(X,w)\) produced by that declared Sinkhorn route.

A warm-start learner does not try to predict the transported output cloud first.
It tries to predict a latent teacher coordinate, such as dual variables,
log-scalings, or scaling vectors, that initializes the corrective solve.  Write
\begin{equation}
\label{eq:bf-learned-ot-student-map}
  h_\psi = S_\psi(X,w;\varepsilon,N,d_x),
\end{equation}
where \(h_\psi\) is later translated into the initialization of the retained
corrective Sinkhorn solver.

For a concrete exposition, it is useful to keep one canonical latent target in
mind.  Suppose the student predicts initial log-scalings
\begin{equation}
\label{eq:bf-learned-ot-logscaling-student}
  h_\psi
  =
  (\ell_a^{(0)},\ell_b^{(0)}),
\end{equation}
which are then exponentiated to give
\begin{equation}
\label{eq:bf-learned-ot-scaling-init}
  a^{(0)}=\exp(\ell_a^{(0)}),
  \qquad
  b^{(0)}=\exp(\ell_b^{(0)}).
\end{equation}
The corrective retained-teacher route then performs the same balancing equations as
before,
\begin{equation}
\label{eq:bf-learned-ot-corrective-updates}
  a^{(k+1)} = w \oslash (K_\varepsilon b^{(k)}),
  \qquad
  b^{(k+1)} = u \oslash (K_\varepsilon^\top a^{(k+1)}),
\end{equation}
until the declared correction budget is exhausted.  This makes the chapter's
teacher-preserving logic explicit: the learner may change the starting point of the
finite Sinkhorn route, but the corrected route still returns to the same declared
kernel, the same marginal constraints, and the same barycentric output rule.

The operational contrast is then simple:
\begin{itemize}
  \item a naive initialization changes only how quickly the solver reaches the
  finite teacher,
  \item a learned initialization also changes only that execution path,
  \item the corrected coupling and barycentric output are still judged against
  the same declared finite teacher object.
\end{itemize}

\paragraph{Running-example walkthrough.}
The warm-start story can now be read in one seven-step ladder:
\begin{enumerate}[label=(\arabic*)]
  \item start from a weighted cloud \((X,w)\), a declared cost, and the equal target
    marginal \(u\),
  \item define the retained finite teacher coupling \(\Pi_{\varepsilon,K}\),
  \item choose a latent teacher coordinate to predict, such as log-scalings,
  \item let the student predict that latent coordinate,
  \item translate the prediction into the initialization of the retained corrective
    Sinkhorn routine,
  \item run the declared number of corrective updates to obtain the corrected
    coupling \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\),
  \item form the barycentric equal-weight cloud and compare it against the teacher
    output through residual checks.
\end{enumerate}
The rest of the chapter merely fills in this ladder mathematically and then asks
how adjacent retained-teacher families fit into the same audit posture.

\paragraph{One concrete warm-start recipe.}
To keep the chapter operational, fix one explicit reading.  Suppose the teacher
latent target is the pair of initial log-scalings
\((\ell_a^{(0)},\ell_b^{(0)})\), the student network predicts exactly that pair,
and deployment uses a declared corrective budget \(K_{\mathrm{corr}}\).  Then one
practical recipe is:
\begin{enumerate}[label=(\roman*)]
  \item solve the retained teacher route on training clouds and store the resulting
    target log-scalings,
  \item train the student to predict those log-scalings from \((X,w)\),
  \item exponentiate the predicted pair to form \((a^{(0)},b^{(0)})\),
  \item run \(K_{\mathrm{corr}}\) corrective Sinkhorn updates,
  \item compare the corrected barycentric cloud and residuals against the teacher.
\end{enumerate}
This is only one possible retained-teacher design, but it is concrete enough that a
first-year reader can see what is stored, what is predicted, and what is still
protected by correction.

\paragraph{Plain-language takeaway.}
A warm-start learner is not asked to invent the transport result.  It is asked
to guess a better place from which the retained teacher solver should begin.

\paragraph{What the rest of the chapter is for.}
A first-year reader should treat everything through the end of the residual
hierarchy as the primary implementation story of the chapter.  The later Nystr"om,
positive-feature, low-rank, and minibatch sections are broader retained-teacher
family context: they are useful because they preserve the same audit posture, but
they are not new primary objects to learn first.  If the notation starts to feel
dense, return to the seven-step warm-start ladder above and ask how each later
family changes only the route to the teacher rather than the teacher object itself.

\section{What the teacher computes}
\label{sec:bf-learned-ot-hierarchy}

There is no single teacher unless the variant is named.  In this chapter, the
main teacher is the retained EOT/Sinkhorn route from the previous chapter.  The
reader therefore needs one short local recap of that teacher before any learned
warm start is introduced.

Given weighted source cloud \(X=(x_1,\ldots,x_N)^\top\), source marginal
\(w\in\Delta_N\), equal target marginal
\(u=(1/N,\ldots,1/N)\), declared cost matrix \(C\), and regularization
\(\varepsilon>0\), the entropic teacher object is
\begin{equation}
\label{eq:bf-learned-ot-entropic-primal}
  \Pi_\varepsilon^\star
  =
  \arg\min_{\Pi\in\mathcal U(w,u)}
  \left\{
    \langle \Pi,C\rangle
    + \varepsilon\sum_{i,j}\Pi_{ij}(\log\Pi_{ij}-1)
  \right\}.
\end{equation}
Its Gibbs/scaling form is
\begin{equation}
\label{eq:bf-learned-ot-gibbs-factorization}
  \Pi_\varepsilon^\star
  =
  \diag(a)K_\varepsilon\diag(b),
  \qquad
  (K_\varepsilon)_{ij}=\exp(-C_{ij}/\varepsilon),
\end{equation}
with scaling vectors satisfying the balancing equations
\begin{equation}
\label{eq:bf-learned-ot-balancing}
  a = w \oslash (K_\varepsilon b),
  \qquad
  b = u \oslash (K_\varepsilon^\top a).
\end{equation}
After a declared finite number \(K\) of Sinkhorn repairs, the finite numerical
teacher is
\begin{equation}
\label{eq:bf-learned-ot-finite-teacher}
  \Pi_{\varepsilon,K}
  =
  \diag(a^{(K)})K_\varepsilon\diag(b^{(K)}).
\end{equation}
The corresponding barycentric teacher output is
\begin{equation}
\label{eq:bf-learned-ot-barycentric-teacher}
  Y_\tau(X,w,\varepsilon)
  =
  B(\Pi_\tau(X,w,\varepsilon)),
  \qquad
  B(\Pi)_j = \frac{1}{u_j}\sum_{i=1}^N \Pi_{ij}x_i.
\end{equation}
In the equal-weight case, this is
\begin{equation}
\label{eq:bf-learned-ot-barycentric-equal}
  B(\Pi)_j = N\sum_{i=1}^N \Pi_{ij}x_i.
\end{equation}

These formulas identify the three most important teacher objects in this chapter:
\begin{enumerate}[label=(\roman*)]
  \item the exact regularized coupling \(\Pi_\varepsilon^\star\),
  \item the finite numerical teacher \(\Pi_{\varepsilon,K}\),
  \item the barycentric teacher output \(Y_\tau=B(\Pi_\tau)\).
\end{enumerate}
The chapter's neural route is only intended to accelerate the finite teacher, not
to redefine these objects.

\paragraph{Object audit.}
The learned quantity is a latent solver state.  The corrected transport object is
still the finite teacher-side coupling or its barycentric output.  In the
vocabulary of the OT block, this is an execution-path approximation, not a
transport-object change.

\paragraph{Canonical reading convention.}
For the rest of this chapter, the safest first-year reading convention is:
\begin{itemize}
  \item \(\Pi_\varepsilon^\star\) is the exact entropic teacher,
  \item \(\Pi_{\varepsilon,K}\) is the retained finite numerical teacher,
  \item \(h_\tau\) is the latent teacher coordinate the student is asked to learn,
  \item \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\) is the corrected student
  output,
  \item all residuals are interpreted relative to that corrected student output
  versus the chosen finite teacher.
\end{itemize}
Whenever the notation becomes crowded, return to that five-line convention rather
than trying to remember every symbol historically.

\paragraph{Teacher notation shortcut.}
To keep the notation stable, the chapter will use the following shorthand from this
point onward:
\begin{itemize}
  \item exact teacher = \(\Pi_\varepsilon^\star\),
  \item finite teacher = \(\Pi_{\varepsilon,K}\),
  \item generic teacher-side cloud = \(Y_\tau=B(\Pi_\tau)\), with \(\tau\) always
  meaning the chosen retained teacher route,
  \item corrected student output = \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\).
\end{itemize}
So when a later section says ``compare against the teacher,'' the default first-year
reading should be: compare the corrected student output against the chosen finite
retained teacher unless the exact teacher is explicitly named.

\section{Algorithm: retained-teacher warm-started Sinkhorn}
\label{sec:bf-learned-ot-teacher}

\begin{algorithm}[ht]
\caption{\texttt{train\_and\_deploy\_warmstarted\_sinkhorn}}
\begin{algorithmic}[1]
\Require training distribution \(\mathcal D\) over weighted particle clouds, declared cost construction, regularization schedule \(\varepsilon\), teacher Sinkhorn solver, architecture class \(S_\psi\), deployment-time corrective budget \(K_{\mathrm{corr}}\)
\Ensure trained predictor \(S_\psi\), corrected deployment coupling \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\), barycentric output cloud \(\widetilde X_{\psi,\mathrm{corr}}\), and declared residual diagnostics
\State Generate teacher problems by solving the retained EOT/Sinkhorn route on sampled clouds from \(\mathcal D\)
\State Extract the latent teacher target to be predicted, e.g. dual/scaling coordinates
\State Fit the predictor \(S_\psi\) on this teacher data using the declared loss family
\State For a new weighted cloud \((X,w)\), compute the student prediction \(h_\psi=S_\psi(X,w;\varepsilon,N,d_x)\)
\State Translate \(h_\psi\) into the initial state of the retained corrective Sinkhorn solver
\State Run \(K_{\mathrm{corr}}\) corrective Sinkhorn updates to obtain \(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\)
\State Form the barycentric equal-weight output cloud \(\widetilde X_{\psi,\mathrm{corr}}=B(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}})\)
\State Record row/column residuals, teacher-student discrepancy, and any downstream scalar checks
\end{algorithmic}
\end{algorithm}

\paragraph{What the student is allowed to change.}
Only the initialization of the retained teacher solver, or another explicitly
named latent teacher coordinate.

\paragraph{What the teacher still protects.}
The declared transport constraints, the EOT objective, and the corrected
barycentric output after refinement.

\paragraph{Why this chapter keeps log-scalings as the running latent target.}
The chapter names duals, log-scalings, and scaling vectors because all three occur
naturally in entropic OT implementations.  But the running exposition uses
log-scalings because they are concrete, per-particle, and easy to translate back
into the retained corrective solver.  A first-year reader should therefore treat the
log-scaling example as the chapter's canonical student target unless a later section
explicitly announces a different latent object.

\section{Evidence contract for retained-teacher learning}
\label{sec:bf-learned-ot-training-objective}

Let \(h_\tau\) denote the teacher latent target and let
\(\mathcal D(N,d_x,\varepsilon,m,\eta)\) be the training distribution over
problem instances.  A simple supervised teacher-student fitting problem is
\begin{equation}
\label{eq:bf-learned-ot-training-objective}
  \psi^\star
  \in
  \arg\min_\psi
  \mathbb E_{(X,w,\varepsilon,m)\sim\mathcal D}
  \left[
    \mathcal L\{S_\psi(X,w;\varepsilon,N,d_x),h_\tau(X,w,\varepsilon)
    \}
  \right].
\end{equation}
In practice, the deployment claim is not about this loss alone.  It is about
whether the corrected route preserves the teacher output well enough on held-out
and stress-test clouds.

\subsection{Permutation symmetry as a design check}
\label{sec:bf-learned-ot-symmetry}

Particle labels are arbitrary.  For a permutation matrix
\(P\in\mathbb R^{N\times N}\), the predictor should at least satisfy a declared
permutation symmetry.  In the output-cloud case the natural condition is
\begin{equation}
\label{eq:bf-learned-ot-equivariance}
  S_\psi(PX,Pw;\varepsilon,N,d_x)
  =
  P\,S_\psi(X,w;\varepsilon,N,d_x)
  \quad
  \text{or the corresponding latent-state analogue.}
\end{equation}
The distinction matters because different latent targets transform differently.  If
the student predicts a per-particle object such as initial log-scalings, then
permutation equivariance is the natural symmetry.  If the student predicts a
compressed global latent statistic, a weaker invariance or gauge-fixed analogue may
be more appropriate.  In either case, the chapter's main design check is the same:
relabeling particles should not change the mathematical content of the predicted
teacher state.

\subsection{Residual hierarchy}
\label{sec:bf-learned-ot-residuals}

A retained-teacher route should be judged by a hierarchy of residuals rather than
by one discrepancy alone.  The first checks are the marginal feasibility residuals
of the corrected route,
\begin{equation}
\label{eq:bf-learned-ot-row-col-residuals}
  r_{\mathrm{row}}
  =
  \|\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}\mathbf 1-w\|_1,
  \qquad
  r_{\mathrm{col}}
  =
  \|\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}}'\mathbf 1-u\|_1.
\end{equation}
These ask whether the corrected route still satisfies the declared transport
constraints.  They do \emph{not} by themselves show that the corrected student has
reproduced the teacher's transported cloud.

The next deployment residual is the teacher-student discrepancy after correction:
\begin{equation}
\label{eq:bf-learned-ot-residual}
  R_{\psi,\tau}(X,w,\varepsilon)
  =
  B(\widehat\Pi_{\psi,\varepsilon,K_{\mathrm{corr}}})
  -
  B(\Pi_\tau(X,w,\varepsilon)).
\end{equation}
This barycentric discrepancy measures how well the corrected student reproduces the
chosen teacher cloud.  It still does not by itself control filtering or HMC target
error.  For kernel-compression families such as Nystr"om or positive-feature
Sinkhorn, one may also monitor the transported-cloud discrepancy more directly,
\begin{equation}
\label{eq:bf-learned-ot-transport-discrepancy}
  \|\widetilde P X - P_\tau X\|,
\end{equation}
because the transport application is the actual object later code consumes.

\paragraph{Mini-example of the hierarchy.}
The logic of these residuals is easiest to remember by a simple contrast.  A route
may satisfy the row and column sums almost exactly, so \(r_{\mathrm{row}}\) and
\(r_{\mathrm{col}}\) look excellent, yet still move mass through a factorized or
restricted representation whose transported cloud differs visibly from the chosen
teacher output.  In that case feasibility has been preserved, but teacher parity
has not.  Conversely, a corrected route might match the teacher cloud very closely
while a downstream scalar still drifts because that scalar is more sensitive than the
cloud discrepancy alone suggests.  This is why the residual hierarchy must be read
in order: feasibility first, teacher-cloud parity second, downstream scalar checks
third.

A practical retained-teacher claim should therefore keep the following evidence
contract explicit:
\begin{enumerate}[label=(\arabic*)]
  \item the predicted latent teacher object is named explicitly,
  \item the corrected route satisfies the declared row/column residual contract,
  \item any factorized kernel or coupling approximation is checked against the
    teacher at the transported-cloud level,
  \item the corrected cloud discrepancy remains below the declared threshold on the
    tested envelope,
  \item any downstream scalar or HMC-adjacent target is checked only after the
    corrected output is formed.
\end{enumerate}
A useful first-year rule of thumb is: if the marginal constraints are good but the
transported-cloud discrepancy is visibly non-small on the toy or dense reference
problems, the warm start is not yet trustworthy no matter how fast it runs.  If the
transported cloud is close but a downstream scalar still drifts materially, then the
chapter's local parity checks have passed only at the cloud level and the learned
route still should not be promoted for that scalar use case.

\subsection{Where broader retained-teacher families fit}
\label{sec:bf-learned-ot-retained-family}

The chapter's core implementation story stops with the retained warm-start lane
above.  The remaining subsections should be read as broader retained-teacher family
context: they answer the same teacher-preservation question, but they are not new
primary chapter threads.  Their role is to show how adjacent scalable-entropic
families can still be interpreted through the same retained-teacher audit when the
first test remains parity against the declared finite teacher.

Warm-start predictors are the narrowest retained-teacher lane.  Other scalable
OT families --- such as Nystrom, positive-feature, low-rank, or minibatch
teacher-side approximations --- may still belong to the same lane only when
their first audit question is teacher preservation rather than speed alone
\citep{altschuler2019nystrom,scetbon2020positivefeatures,forrow2019factored,scetbon2021lowrank,scetbon2022lowrank,genevay2016stochastic,genevay2018samplecomplexity,nguyen2022bomb,nguyen2021partialminibatch}.
Once the approximation is no longer naturally judged by teacher-preservation
residuals, it belongs in the later target-changing chapters instead of this one.

\paragraph{How to read the family context that follows.}
If the next subsections start to feel denser than the warm-start core, the safest
first-year reading rule is simple: treat them as alternate ways to preserve the same
teacher-side audit posture, not as new primary chapter threads.  In other words,
keep asking the same question: what compressed route is being proposed, and how is
its corrected output compared back to the finite retained teacher?

\subsection{Exact online and GPU Sinkhorn as the reference lane}
\label{sec:bf-learned-ot-reference-lane}

Before turning to approximate retained-teacher families, the reader should keep one
semantics-preserving reference lane in view.  Exact online or GPU-specialized
Sinkhorn methods preserve the same entropic teacher object while changing only how
the kernel reductions are executed in memory or on hardware
\citep{feydy2019geomloss,charlier2021keops,ye2026flashsinkhorn,xiao2026fastsinkhorn}.
For squared Euclidean costs,
\begin{equation}
\label{eq:bf-learned-ot-reference-sqeuclidean}
  C_{ij}
  =
  \|x_i-z_j\|_2^2
  =
  \|x_i\|_2^2+\|z_j\|_2^2-2x_i'z_j,
\end{equation}
and a stabilized potential update can be written schematically as a row-wise
log-sum-exp reduction,
\begin{equation}
\label{eq:bf-learned-ot-reference-logupdate}
  f_i
  \leftarrow
  -\varepsilon\log\sum_j
  \exp\left((g_j-C_{ij})/\varepsilon+\log u_j\right).
\end{equation}
The same teacher-side barycentric output is then recovered from the resulting finite
coupling.  This route is the reference lane because it preserves semantics while
avoiding full dense materialization where possible.  It is not yet a new transport
approximation class.  It also does not remove the dominant all-pairs interaction
structure.  That is exactly why the later factorized families matter: they are not
introduced before the reader has a semantics-preserving reference route in hand.

\subsection{Nystrom kernel approximations}
\label{sec:bf-learned-ot-nystrom}

A natural retained-teacher route is to approximate the Gibbs kernel rather than
the final transport object \citep{altschuler2019nystrom}.  For a Gaussian-like
kernel, write
\begin{equation}
\label{eq:bf-learned-ot-nystrom-gibbs}
  K_{ij}
  =
  k_\eta(x_i,x_j)
  =
  \exp(-\eta\|x_i-x_j\|^2).
\end{equation}
The Nystr"om idea is to choose a smaller landmark set
\begin{equation}
\label{eq:bf-learned-ot-nystrom-landmarks}
  X_r=\{\tilde x_1,\ldots,\tilde x_r\}\subset X
\end{equation}
and define the factor matrices
\begin{equation}
\label{eq:bf-learned-ot-nystrom-factors}
  V_{ij}=k_\eta(x_i,\tilde x_j),
  \qquad
  A_{jj'}=k_\eta(\tilde x_j,\tilde x_{j'}).
\end{equation}
Then the dense kernel is approximated by
\begin{equation}
\label{eq:bf-learned-ot-nystrom-kernel}
  \widetilde K = V A^{-1}V'.
\end{equation}
This is not just a storage identity.  It changes how the expensive kernel action is
computed.  If \(A=LL'\) is a Cholesky factorization, then a matrix-vector product
is applied as
\begin{equation}
\label{eq:bf-learned-ot-nystrom-matvec}
  \widetilde K v
  =
  V\left(L'^{-1}\left(L^{-1}(V'v)\right)\right),
\end{equation}
so the dominant kernel multiply is now factorized rather than dense.  After the
factorization has been built, each multiply costs \(O(Nr)\) rather than
\(O(N^2)\).

Because Sinkhorn balancing uses kernel-vector products, the same factorization
feeds directly into the retained finite teacher route.  The approximate teacher-side
plan becomes
\begin{equation}
\label{eq:bf-learned-ot-nystrom-plan}
  \widetilde P
  =
  D_1\widetilde K D_2
  =
  D_1 V A^{-1}V' D_2,
\end{equation}
and the barycentric transport application becomes
\begin{equation}
\label{eq:bf-learned-ot-nystrom-transport}
  \widetilde P X
  =
  D_1 V A^{-1}V' D_2 X.
\end{equation}
This is the key retained-teacher ladder for the family:
\begin{enumerate}[label=(\roman*)]
  \item approximate the kernel action,
  \item run the same teacher-side balancing logic on the factorized action,
  \item compare the corrected transport output against the dense or streamed teacher.
\end{enumerate}
A practical Nystr"om route therefore needs more than the factorization formula
itself.  The rank is usually increased adaptively until the approximation is judged
good enough.  A simple entrywise diagnostic from the scalable-OT note is
\begin{equation}
\label{eq:bf-learned-ot-nystrom-diag-error}
  \|K-\widetilde K\|_\infty
  =
  1-\min_i \widetilde K_{ii},
\end{equation}
which asks whether the approximate kernel still reproduces the self-similarity
scale of the original kernel.  The deeper lesson is geometric: Nystr"om helps when
the post-flow Gibbs kernel has low effective rank, but it can fail when the rank
needed to represent the kernel grows quickly with ambient or intrinsic dimension,
or when the entropic regularization is too small for low-rank structure to remain
visible.

For retained-teacher use, the decisive checks are still teacher-facing rather than
speed-facing.  In addition to marginal residuals, one should monitor the
transported-cloud discrepancy
\begin{equation}
\label{eq:bf-learned-ot-nystrom-cloud-discrepancy}
  \|\widetilde P X-P_\tau X\|,
\end{equation}
because the transported cloud is the actual object later code consumes.  The
primary diagnostics are therefore not speed alone but kernel-approximation error,
marginal residuals after correction, and transported-cloud discrepancy against the
chosen teacher route.

\subsection{Positive-feature Sinkhorn}
\label{sec:bf-learned-ot-positive-feature}

A related route factorizes the kernel through positive features rather than through
landmark columns \citep{scetbon2020positivefeatures}.  Choose a nonnegative
feature map
\begin{equation}
\label{eq:bf-learned-ot-positive-map}
  \phi_\theta(x)\in\mathbb R_+^r
\end{equation}
and define the induced kernel
\begin{equation}
\label{eq:bf-learned-ot-positive-kernel}
  K_\theta(x_i,x_j)
  =
  \langle \phi_\theta(x_i),\phi_\theta(x_j)\rangle.
\end{equation}
The corresponding cost representation is then
\begin{equation}
\label{eq:bf-learned-ot-positive-cost}
  c_\theta(x_i,x_j)
  =
  -\varepsilon\log K_\theta(x_i,x_j).
\end{equation}
At the matrix level, define source and target feature collections by
\begin{align}
\label{eq:bf-learned-ot-positive-matrices}
  \Xi
  &=
  [\phi_\theta(x_1),\ldots,\phi_\theta(x_N)],\\
  \mathrm Z
  &=
  [\phi_\theta(z_1),\ldots,\phi_\theta(z_N)],
\end{align}
so the sample kernel factorizes as
\begin{equation}
\label{eq:bf-learned-ot-positive-factorization}
  K_\theta = \Xi'\,\mathrm Z.
\end{equation}
Balancing steps can therefore be expressed through feature-space products rather
than through a materialized dense kernel.  After the feature matrices are formed,
the dominant kernel-vector products scale like \(O(r(N+N))\), so the attraction is
again execution-path efficiency rather than a new transport doctrine.

The same factorization also has a useful integral reading.  One may view the kernel
as arising from a positive feature integral
\begin{equation}
\label{eq:bf-learned-ot-positive-integral}
  k(x,y)
  =
  \int_U \phi(x,u)'\phi(y,u)\,d\rho(u),
\end{equation}
and then approximate that integral by a finite feature family,
\begin{equation}
\label{eq:bf-learned-ot-positive-finite}
  \phi_\theta(x)
  =
  \frac{1}{\sqrt r}
  (\phi(x,u_1),\ldots,\phi(x,u_r)).
\end{equation}
This explains why the factorized kernel should still be read as an approximation to
the teacher-side kernel action, not automatically as the exact original kernel.
The corresponding factored transport
application is
\begin{equation}
\label{eq:bf-learned-ot-positive-transport}
  PX
  =
  \diag(u)\Xi'\,\mathrm Z\diag(v)X.
\end{equation}
This gives the same pedagogical ladder as Nystr"om:
\begin{enumerate}[label=(\roman*)]
  \item replace dense kernel application by a factorized kernel action,
  \item run the retained balancing logic through the factors,
  \item judge the corrected route by teacher-preservation residuals rather than by
    runtime alone.
\end{enumerate}
One can also write the induced entropic dual in factorized form,
\begin{equation}
\label{eq:bf-learned-ot-positive-dual}
  W_{\varepsilon,c_\theta}(\mu,\nu)
  =
  \max_{\alpha,\beta}
  a'\alpha+b'\beta
  - \varepsilon(\Xi e^{\alpha/\varepsilon})'\,\mathrm Z e^{\beta/\varepsilon}
  + \varepsilon,
\end{equation}
which makes the same point from the dual side: the expensive dense interaction is
replaced by feature-space products.  The caution is sharper than for a pure
engineering variant.  A positive-feature route changes the effective kernel
representation and therefore must be judged by teacher-preservation residuals,
marginal checks, and transported-cloud parity, not by runtime alone.  It is
retained-teacher only to the extent that the corrected route is still explicitly
audited against the chosen teacher object.

\subsection{Direct low-rank coupling approximations}
\label{sec:bf-learned-ot-low-rank}

Kernel approximations are not the only retained-family option.  One may also
factorize the coupling itself \citep{forrow2019factored,scetbon2021lowrank,scetbon2022lowrank,scetbon2023unbalanced}.  The key conceptual distinction is:
\begin{quote}
Approximating the kernel asks how to preserve the same teacher-side transport from
a compressed kernel action.  Constraining the coupling asks whether a restricted
family of couplings is already sufficient.
\end{quote}
The second route is more than a fast matvec; it changes the feasible family being
searched.

A useful formal entry point is the nonnegative rank of a nonnegative matrix,
\begin{equation}
\label{eq:bf-learned-ot-nonnegative-rank}
  \operatorname{rank}_{+}(M)
  =
  \min\left\{
  r:\;M=\sum_{k=1}^r M_k,\;
  \operatorname{rank}(M_k)=1,\;M_k\ge0
  \right\}.
\end{equation}
The rank-constrained coupling family is then
\begin{equation}
\label{eq:bf-learned-ot-low-rank-polytope}
  \Pi_{a,b}(r)
  =
  \{P\in\Pi(a,b):\operatorname{rank}_{+}(P)\le r\}.
\end{equation}
A generic factorized parameterization is
\begin{equation}
\label{eq:bf-learned-ot-low-rank-coupling}
  P
  =
  Q\,\diag(1/g)R',
\end{equation}
with factor constraints
\begin{equation}
\label{eq:bf-learned-ot-low-rank-constraints}
  Q\in\Pi(a,g),
  \qquad
  R\in\Pi(b,g),
  \qquad
  g\in\Delta_r.
\end{equation}
These constraints mean that the factors themselves carry source mass \(a\), target
mass \(b\), and an intermediate latent mass profile \(g\).  Equivalently, one may
speak of the factored feasible family
\begin{equation}
\label{eq:bf-learned-ot-low-rank-feasible-family}
  FC_{a,b}(r)
  =
  \{(Q,R,g): Q\in\Pi(a,g),\;R\in\Pi(b,g),\;g\in\Delta_r\},
\end{equation}
so the low-rank optimization problem can be read either as a search over
rank-constrained couplings or as a search over factored feasible triples.  In the
same notation, the objective itself is
\begin{equation}
\label{eq:bf-learned-ot-low-rank-objective}
  \min_{P\in\Pi_{a,b}(r)}\langle C,P\rangle
  =
  \min_{(Q,R,g)\in FC_{a,b}(r)}
  \left\langle C,Q\diag(1/g)R'\right\rangle.
\end{equation}
Transport application
then becomes
\begin{equation}
\label{eq:bf-learned-ot-low-rank-transport}
  PX
  =
  Q\diag(1/g)R'X,
\end{equation}
so storage and barycentric application are both executed through the low-rank
factors rather than through a dense coupling.

A richer extension introduces a small latent transport plan \(T\) between the two
factor families.  In schematic form, the induced coupling becomes
\begin{equation}
\label{eq:bf-learned-ot-low-rank-latent}
  P
  =
  Q\diag(1/g_Q)
  T
  \diag(1/g_R)R',
\end{equation}
which allows mass to route through a compact latent transport instead of only
through diagonal latent components.  A practical rank-adequacy question then
appears: is the chosen rank large enough?  The note's stronger low-rank lines use
reduced-gradient or dual-slack diagnostics for that purpose.  Even without carrying
those full certificates into this chapter, the reader should keep the right lesson
in view: low-rank coupling is not only a storage trick; it is an optimization over a
restricted feasible family, so parity against the teacher must be checked whenever
rank is fixed in advance.

For the present chapter, the important boundary is interpretive.  If the first test
remains teacher-preservation parity after correction, then the family can still be
read through the retained-teacher lens.  If the prose instead starts defending the
restricted family as a different final transport object in its own right, the
discussion has crossed toward the changed-object lane treated more naturally in the
next chapter.

\subsection{Stochastic and minibatch cautions}
\label{sec:bf-learned-ot-minibatch}

Stochastic or minibatch transport ideas are tempting because they reduce local work
per update \citep{genevay2016stochastic,genevay2018samplecomplexity,nguyen2022bomb,nguyen2021partialminibatch}.  But a minibatch-optimal plan is not automatically a valid full-
ensemble deterministic resampling map.  Solving transport on sampled subclouds or
local batches may be a useful approximation strategy, yet the first question
remains whether the resulting corrected route preserves the declared teacher well
enough on the full weighted ensemble.

For this reason, minibatch variants belong to the retained-teacher lane only when
a later correction stage or explicit audit shows that full-ensemble marginals,
transported-cloud behavior, and downstream scalars still track the declared
teacher route.  The practical validation sequence should therefore stay stricter
than a paper-summary reading:
\begin{enumerate}[label=(\arabic*)]
  \item first check small dense transported-particle parity, not only scalar OT
    values,
  \item then check row and column marginals plus positivity/non-finite failures,
  \item then inspect representation diagnostics such as rank, feature dimension,
    sparsity, projection count, or block locality,
  \item only after value parity consider gradient smoke tests and larger-
    \(N,d_x\) runtime or memory ladders,
  \item and only after those local checks interpret downstream filtering evidence.
\end{enumerate}
If the minibatch construction is instead promoted as its own randomized or local
replacement transport, it should be read through the later structure-changing
chapters rather than through the present retained-teacher one.

\begin{center}
\small
\begin{tabular}{p{0.18\textwidth} p{0.22\textwidth} p{0.22\textwidth} p{0.25\textwidth}}
\toprule
Family & What is approximated? & What stays teacher-protected? & First residual to check \\
\midrule
Warm start & solver initialization or latent state & finite Sinkhorn teacher and barycentric output & corrected cloud discrepancy \\
Nystrom / positive-feature & kernel action or feature representation & declared teacher-side transport after correction & marginal and teacher-preservation residual \\
Low-rank coupling & feasible coupling representation & declared output object only if parity remains the first test & coupling/cloud parity under restriction \\
Minibatch / stochastic routes & local or sampled transport work & teacher status only if full-ensemble behavior is still audited & full-ensemble marginal and cloud behavior \\
\bottomrule
\end{tabular}
\end{center}

\subsection{How the broader retained-teacher approximations fit here}
\label{sec:bf-learned-ot-broader-retained}

The same residual logic gives the chapter a place to classify the broader
scalable-OT approximations from the survey.  If a Nystrom factorization, a
positive-feature kernel, or a low-rank coupling factorization is proposed as a
retained-teacher acceleration, then the first audit question is still teacher
preservation: how close is the transported cloud or coupling application to the
declared teacher route from Chapter~\ref{ch:bf-dpf-entropic-ot-sinkhorn}?

For example, a kernel-factor route may replace the dense Gibbs kernel by a
factorized approximation while still producing a coupling application of the same
schematic form.  A low-rank coupling route may store and apply transport through
structured factors rather than a dense matrix.  A stochastic or minibatch route
may reduce computation by solving smaller transport subproblems.  In all such
cases, this chapter's doctrine is the same: they are still retained-teacher
candidates only if the approximation error is measured against the teacher object
rather than hidden behind speedups alone.

That is why the residual hierarchy should be read broadly.  The diagnostic is not
restricted to one neural warm-start map.  It is the first reusable test for any
approximation family that claims to accelerate the teacher without changing the
meaning of the transported output.  Once the approximation stops being naturally
judged by teacher-preservation residuals, it belongs in the later target-changing
chapters instead of this retained-teacher one.

\section{Filtering-loop insertion point}
\label{sec:bf-learned-ot-loop}

Inside the filtering loop, the retained-teacher neural route sits exactly where
Chapter~\ref{ch:bf-dpf-entropic-ot-sinkhorn} placed the finite EOT/Sinkhorn
layer:
\begin{equation}
\label{eq:bf-learned-ot-loop}
  \{x_i,w_i\}_{i=1}^N
  \xrightarrow{\mathrm{student\ warm\ start + retained\ Sinkhorn}}
  \{\tilde x^{(j)},1/N\}_{j=1}^N.
\end{equation}
It accelerates the equal-weighting layer after the weighted cloud has already
been formed.  It does not redefine the earlier particle-weighting rule itself.

\section{Verification and non-claims}
\label{sec:bf-learned-ot-verification}
\label{sec:bf-learned-ot-boundary}

A practical implementation should still check a few local quantities carefully:
\begin{enumerate}[label=(\arabic*)]
  \item teacher generation is declared and reproducible,
  \item the predicted latent teacher object is named explicitly,
  \item the corrected route satisfies the row/column residual contract,
  \item teacher-student cloud discrepancy remains below the declared threshold on
  the tested envelope,
  \item the downstream scalar is checked whenever the learned route is embedded
  in a filtering likelihood or HMC target.
\end{enumerate}

These checks are mathematical and numerical, not bureaucratic.  Their purpose is
only to ensure that the reader can tell whether the implementation is still
computing the declared teacher-student object or has drifted to a different one.

This chapter does \emph{not} claim:
\begin{itemize}
  \item that the student replaces the teacher object,
  \item that small teacher-student residuals automatically imply posterior or HMC
  correctness,
  \item that broader neural OT families are unimportant, only that they answer a
  different implementation question.
\end{itemize}
Its role is narrower: fix one teacher-student acceleration route in a form that a
reader can nearly implement directly.

\FloatBarrier
