\chapter{Dynamic Geodesic and Operator Learners, and the Filtering Target Contract}
\label{ch:bf-neural-ot-dynamic-target-contract}

The previous chapter treated direct map-learning families such as ICNN/Brenier
and Monge-gap approaches.  The present chapter changes the learned object again.
Instead of asking the network to learn one static transport map, it asks what
happens when the learned object is a dynamic transport path, a velocity field, or
an operator acting on families of measures.

This chapter is intentionally secondary.  Its job is not to become another core
implementation chapter.  Its job is to explain why these broader learned families
carry an even heavier target-status burden than retained-teacher acceleration or
static direct-map learning.

\section{What changes when the learned object becomes richer?}
\label{sec:bf-neural-ot-dynamic-orientation}

The easiest way to read this chapter is to keep the following contrast in mind.
\begin{itemize}
  \item The retained-teacher neural chapter learned how to solve a declared
  transport teacher faster.
  \item The direct-map chapter learned one final transport map.
  \item This chapter asks what happens when the learned object is richer than one
  final map: a path, a velocity field, or an operator on entire families of
  measures.
\end{itemize}

So the main burden is not just implementation difficulty.  The chapter's stable
question is narrower and more useful:
\begin{quote}
If the learned object is richer than one endpoint map, what scalar is downstream
code actually using, and what gradient interpretation still remains justified?
\end{quote}
For the rest of this chapter, the recurring OT-block terms should be read in one
stable way:
\begin{itemize}
  \item ``teacher'' means the earlier dense or retained-teacher reference route from
  the OT block,
  \item ``target-changing'' means the learned object is no longer automatically that
  same reference transport object,
  \item ``same-scalar parity'' means the reported gradient must differentiate the
  exact scalar value that is actually built from the learned object used by the
  code.
\end{itemize}

\begin{center}
\small
\begin{tabular}{p{0.19\textwidth} p{0.22\textwidth} p{0.22\textwidth} p{0.27\textwidth}}
\toprule
Family in this chapter & Learned object & What changes first? & First audit question \\
\midrule
Dynamic geodesic learners & path or velocity field & endpoint plus trajectory & is the reported scalar tied to that same path object? \\
Operator learners & family-level measure map & one problem becomes a class of problems & what target is induced across the problem family? \\
Target-changing boundary families & projected, local, or replacement transport & the transport object itself & what object is now acting on the filtering ensemble? \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{How to read the rest of this chapter.}
The sequence is intentionally staged.  First comes a tiny toy contrast between an
endpoint map and a richer path object.  Next comes the general object/scalar/
gradient audit language that the chapter will reuse.  Only after that does the
chapter look at dynamic geodesic learners, operator learners, and finally the
projected or localized boundary families that should usually be read as target-
changing in the BayesFilter monograph.  To keep the chapter concrete, the reader
should imagine one simple downstream scalar throughout: a filtering-style scalar that
reduces the learned transport output into one value, either from the terminal cloud
alone or from the full richer object.  The chapter keeps asking how that scalar and
its gradient change as the learned object becomes richer.

\paragraph{One repeated filtering-style scalar to keep in mind.}
To keep the chapter concrete, imagine one recurring scalar
\(\widetilde L\) built from the learned transport output before a later reduction in
the filtering pipeline.  In the endpoint-map case, \(\widetilde L\) might depend
only on the terminal equal-weight cloud.  In the dynamic-path case, \(\widetilde
L\) might add a path-action penalty.  In the operator or projected/localized case,
\(\widetilde L\) might depend on a family-level output or a projected/local block
transport summary.  The whole chapter is about how the meaning of \(\widetilde L\)
and therefore of its gradient changes when the learned object changes.

\section{Running path-versus-map example}
\label{sec:bf-neural-ot-dynamic-toy-contrast}

A tiny one-dimensional toy example makes the distinction concrete.  Suppose the
source cloud is
\[
  X_{\mathrm{src}}=\{-1,1\}
\]
and the target cloud is
\[
  X_{\mathrm{tgt}}=\{0,2\}.
\]
A static-map learner tries to learn one rule such as
\begin{equation}
\label{eq:bf-neural-ot-dynamic-static-map}
  T(x)=x+1,
\end{equation}
which moves the source points directly to the target points.

A dynamic path learner instead tries to describe how the particles move as a
function of pseudo-time \(\lambda\in[0,1]\).  Two simple paths with the same
endpoint map are
\begin{align}
\label{eq:bf-neural-ot-dynamic-path-example}
  x_A(\lambda)&=x+\lambda,
  &0\le \lambda\le 1,\\
  x_B(\lambda)&=x+\lambda^2,
  &0\le \lambda\le 1.
\end{align}
Both paths satisfy
\[
  x_A(0)=x_B(0)=x,
  \qquad
  x_A(1)=x_B(1)=x+1=T(x).
\]
So they agree on the final endpoint transformation.  But they disagree as soon as
later code cares about more than the endpoint.  For example,
\begin{equation}
\label{eq:bf-neural-ot-dynamic-midpoint-functional}
  \Phi_{\mathrm{mid}}(x(\cdot))=x(1/2)
\end{equation}
gives
\[
  \Phi_{\mathrm{mid}}(x_A)=x+1/2,
  \qquad
  \Phi_{\mathrm{mid}}(x_B)=x+1/4,
\]
and the path-action functional
\begin{equation}
\label{eq:bf-neural-ot-dynamic-action-functional}
  \mathcal A(x(\cdot))
  =
  \int_0^1 |\dot x(\lambda)|^2\,d\lambda
\end{equation}
produces
\[
  \mathcal A(x_A)=1,
  \qquad
  \mathcal A(x_B)=\int_0^1 (2\lambda)^2\,d\lambda = \frac{4}{3}.
\]
Thus two learned objects can agree on the endpoint map while still inducing
different path-dependent scalars.

This distinction matters because later code need not use only the endpoint.
At the level of notation, the endpoint map in both examples is still
\begin{equation}
\label{eq:bf-neural-ot-dynamic-endpoint-map-summary}
  T(x)=x(1)=x+1,
\end{equation}
but a path-dependent scalar may instead use the whole trajectory
\(x(\cdot)\).  For a BayesFilter-style reading, imagine a scalar that averages only
the terminal mapped particles versus a scalar that also penalizes path action or an
intermediate position.  Those two scalars may agree on the same endpoint map while
still producing different values and different gradients because they consume
fundamentally different learned objects.  Once intermediate positions, path
functionals, or operator outputs enter the scalar that is differentiated
downstream, two learners with the same endpoint map can still define different
effective targets.

\paragraph{Plain-language takeaway.}
A richer learned transport object can preserve the same endpoint while changing
what downstream code means by the value and gradient it consumes.

\section{Richer-object audit: endpoint, path, scalar, gradient}
\label{sec:bf-neural-ot-dynamic-object}
\label{sec:bf-neural-ot-dynamic-same-scalar}

The toy example motivates the general audit ladder.  Once the learned object is
richer than one endpoint map, the reader should slow down and ask three separate
questions: what object is being learned, what scalar later code builds from that
object, and what gradient can therefore be interpreted.

\paragraph{Object inventory.}
A dynamic transport learner no longer tries only to produce one final map
\(T\).  Instead it may try to learn:
\begin{itemize}
  \item a geodesic path \(\rho_\lambda\) between source and target measures,
  \item a velocity field \(v_\lambda(x)\) generating that path,
  \item or an operator \(\mathcal G\) that maps input measures to output
  trajectories, output measures, or transport-related latent coordinates.
\end{itemize}
This is a larger algorithmic change because the learned object is no longer merely
a replacement endpoint map.  It is a representation of how transport unfolds or
how an entire family of transport problems is parameterized.

\paragraph{What downstream code may consume.}
Once a richer object is available, later code may consume:
\begin{itemize}
  \item only the terminal map or endpoint cloud,
  \item intermediate positions or path summaries,
  \item action, regularity, or trajectory penalties,
  \item or operator outputs that vary across a family of related problems.
\end{itemize}
The object question therefore comes first.  One cannot decide what scalar is being
computed until one knows what learned object the scalar is built from.

\paragraph{Same-scalar warning.}
The most important target-contract warning survives every change of learned
object.  If a downstream routine uses a scalar \(\widetilde L\) built from a
learned map, path, or operator, then the reported gradient must differentiate that
same scalar.  Even if this value-gradient parity holds, the result is coherent
only for the learned target unless a correction returns to the teacher or original
target.

\paragraph{Same-scalar template.}
The chapter's main interpretive contract can be written schematically as
\begin{equation}
\label{eq:bf-neural-ot-dynamic-same-scalar-template}
  \mathcal O_\theta
  \longmapsto
  \widetilde L_\theta
  =
  \Phi(\mathcal O_\theta;\text{data}),
  \qquad
  g_\theta = \nabla_\theta \widetilde L_\theta,
\end{equation}
where \(\mathcal O_\theta\) is the learned object (map, path, or operator output),
\(\widetilde L_\theta\) is the actual downstream scalar built from that object,
and \(g_\theta\) is the reported gradient.  A concrete BayesFilter-style example to
keep in mind is a scalar that averages a terminal equal-weight cloud before passing
it into a later filtering reduction.  If a richer learner instead contributes a path
penalty or an operator-generated summary to that scalar, then the object feeding
\(\Phi\) has changed even if the endpoint cloud looks similar.  The gradient is
interpretable only if it differentiates that same declared scalar built from that
same learned object.  Even perfect value-gradient parity therefore certifies
coherence only for the learned target unless a correction returns to the earlier
teacher.

\begin{center}
\small
\begin{tabular}{p{0.24\textwidth} p{0.22\textwidth} p{0.42\textwidth}}
\toprule
Learned object & What may change & First audit question \\
\midrule
static map & endpoint cloud & does the learned endpoint preserve the intended downstream scalar? \\
dynamic path / velocity field & endpoint plus whole trajectory & is the reported gradient tied to the same scalar built from that path? \\
operator on measures or clouds & family-level input-output rule & what target is induced across the family of problems, and is it the one later code interprets? \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Reader checkpoint.}
The present chapter has crossed another object boundary.  A dynamic learner is
not just another static map family.  It changes the interpretation burden because
more than the endpoint may now matter.

\section{Dynamic geodesic learners}
\label{sec:bf-neural-ot-dynamic-geodesic}

Some neural OT papers learn Wasserstein geodesics or associated velocity fields
\citep{hanebeck2023progressiveot,gracykchen2024geonet,liuchen2026wgsolver}.  The
minimal mathematical picture is that a dynamic transport object is no longer only
an endpoint map \(T\).  It is a pseudo-time path of measures \((\rho_\lambda)_{0\le
\lambda\le 1}\) together with a velocity field \(v_\lambda(x)\) satisfying a
continuity equation of the schematic form
\begin{equation}
\label{eq:bf-neural-ot-dynamic-continuity}
  \partial_\lambda \rho_\lambda
  +
  \nabla\cdot(\rho_\lambda v_\lambda)
  =0,
\end{equation}
with endpoints
\begin{equation}
\label{eq:bf-neural-ot-dynamic-endpoints}
  \rho_0=\mu,
  \qquad
  \rho_1=\nu.
\end{equation}
In a Benamou--Brenier-style reading, one then judges a path by a transport-action
objective of the schematic form
\begin{equation}
\label{eq:bf-neural-ot-dynamic-action}
  \int_0^1 \int \|v_\lambda(x)\|^2\,d\rho_\lambda(x)\,d\lambda.
\end{equation}
The endpoint map is now only one summary of a richer learned object; the path and
velocity field are themselves part of what later code could use.

This is mathematically appealing because particle-flow filtering already has a
pseudo-time flavor.  The attraction is therefore obvious: perhaps one could learn
reusable transport dynamics instead of solving a fresh static problem each time.

The caution is equally important.  A dynamic geodesic learner does not answer the
same question as a retained finite Sinkhorn solver.  It learns a richer object and
must therefore be judged against a richer scalar-and-gradient interpretation.  The
first audit question is not merely whether the endpoint fit is good.  It is whether
the downstream scalar is built from the same path object whose gradient is later
reported.

\paragraph{How to read the algorithm below.}
The following routine is schematic.  It is included to make the richer learned
object concrete, not to promote dynamic-path learning as the monograph's default
BayesFilter implementation route.

\begin{algorithm}[ht]
\caption{\texttt{train\_dynamic\_transport\_path}}
\begin{algorithmic}[1]
\Require source-cloud sampler, target-cloud sampler, pseudo-time grid \(0=\lambda_0<\cdots<\lambda_K=1\), learned velocity-field or path parameterization, rollout integrator, endpoint loss, path regularizers, optimizer, stopping rule, and validation suite
\Ensure trained path or velocity-field parameters, deployable terminal map or path object, and pathwise validation diagnostics
\State Initialize the path or velocity parameters and optimizer state
\While{the stopping rule has not been met}
  \State Draw source and target cloud minibatches
  \State Initialize the source particles at pseudo-time \(\lambda_0\)
  \For{each pseudo-time step \(k=0,\ldots,K-1\)}
    \State Evaluate the learned velocity field or path increment at \(\lambda_k\)
    \State Advance the particles using the declared rollout or integration rule
  \EndFor
  \State Compute the endpoint loss against the target cloud at \(\lambda=1\)
  \State Compute any declared pathwise regularizers, such as smoothness or action penalties
  \State Backpropagate the total loss and update the parameters
\EndWhile
  \State On held-out clouds, evaluate endpoint fit, path regularity,
  discretization sensitivity, and downstream scalar sensitivity
  \State Return the trained path/velocity object together with the declared
  diagnostics
\end{algorithmic}
\end{algorithm}

\section{Operator learners}
\label{sec:bf-neural-ot-dynamic-operator}

A related family learns operators on spaces of measures
\citep{lu2021deeponet,li2021fourier,kovachki2023neuraloperator}.  In such
approaches, the input is itself a measure or empirical cloud, and the output is
either another measure, a path of measures, or a set of latent coordinates
describing the transport.  A useful empirical-cloud signature is
\begin{equation}
\label{eq:bf-neural-ot-dynamic-operator-signature}
  \mathcal G_\theta:
  (\mu,\text{context})
  \longmapsto
  \text{endpoint cloud, path object, or transport latent state}.
\end{equation}
Concretely, one may imagine an operator that takes an empirical cloud together
with problem context and returns a terminal transformed cloud, a whole pseudo-time
path, or transport-related latent coordinates for a family of related filtering
problems.  For example, one operator might take a weighted particle cloud together
with observation context and output a full pseudo-time path whose terminal state is
then reduced into a filtering scalar.  Another might output only a terminal cloud.
The chapter's point is that these two outputs can induce different scalars and
therefore different gradient meanings even when they start from the same input
ensemble.  In that sense, the operator section should be read as the family-level
analogue of the earlier path-vs-endpoint toy example: the repeated scalar
\(\widetilde L\) is still the anchor, but the object fed into it has changed.

The main conceptual advantage is reuse across families of problems.  The main
conceptual cost is that the learned object is now even less tied to one declared
teacher solve.  So the first question is no longer only ``does this recover one
finite teacher route well enough?''  It is ``what family-level target is this
operator inducing across the class of problems to which it is applied?''  That
makes these methods valuable as research-direction chapters, but not as the
default first implementation route in this monograph.

\section{Target-changing boundary families}
\label{sec:bf-neural-ot-dynamic-boundary-families}
\label{sec:bf-neural-ot-dynamic-sliced-localized}

Several important high-dimensional families fit this chapter because they should
not be read as compressed dense-teacher solvers by default.  The common theme is
simple: they may still produce useful transport surrogates, but the object they
compute is no longer automatically the same full-state dense transport object that
earler chapters treated as the reference lane.

\paragraph{Projected families: sliced and subspace OT.}
Sliced and subspace OT methods attack high dimension by projecting the measures
onto one-dimensional directions or selected subspaces before solving lower-
dimensional transport problems
\citep{kolouri2019generalized,deshpande2019maxsliced,paty2019subspace,nguyen2025sliced}.
The crucial warm-up fact is one-dimensional OT itself.  For one-dimensional
measures, the monotone transport map is
\begin{equation}
\label{eq:bf-neural-ot-dynamic-monotone-map}
  T(x)=F_\nu^{-1}(F_\mu(x)),
\end{equation}
and the one-dimensional Wasserstein distance is
\begin{equation}
\label{eq:bf-neural-ot-dynamic-oned-wasserstein}
  W_p^p(\mu,\nu)
  =
  \int_0^1
  \left|F_\mu^{-1}(z)-F_\nu^{-1}(z)\right|^p\,dz.
\end{equation}
For empirical distributions, this one-dimensional transport is computed by sorting
projected samples.  For a direction \(\theta\in S^{d-1}\), let \(\theta_\#\mu\)
denote the pushforward of \(\mu\) under \(x\mapsto\langle\theta,x\rangle\).  The
sliced Wasserstein construction then has the schematic form
\begin{equation}
\label{eq:bf-neural-ot-dynamic-sliced}
  \operatorname{SW}_p^p(\mu,\nu)
  =
  \int_{S^{d-1}} W_p^p(\theta_\#\mu,\theta_\#\nu)\,d\theta,
\end{equation}
with empirical directional approximation
\begin{equation}
\label{eq:bf-neural-ot-dynamic-sliced-empirical}
  \widehat{\operatorname{SW}}_p^p(\mu,\nu)
  =
  \frac{1}{L}\sum_{\ell=1}^L
  W_p^p\bigl((\theta_\ell)_\#\mu,(\theta_\ell)_\#\nu\bigr).
\end{equation}
A fully empirical one-dimensional term is computed by sorting,
\begin{equation}
\label{eq:bf-neural-ot-dynamic-sliced-sorting}
  W_2^2(D^\theta,F^\theta)
  =
  \frac{1}{|D|}\sum_i
  \left\|D^\theta_{\pi_D^\theta(i)}-F^\theta_{\pi_F^\theta(i)}\right\|_2^2.
\end{equation}
One can also sharpen random directional averaging into a best-direction criterion,
for example by the max-sliced quantity
\begin{equation}
\label{eq:bf-neural-ot-dynamic-max-sliced}
  \operatorname{MSW}_p^p(\mu,\nu)
  =
  \max_{\theta\in S^{d-1}}W_p^p(\theta_\#\mu,\theta_\#\nu).
\end{equation}
In plain language, these methods first solve easier projected transports and then
use those projected solutions to define a higher-dimensional surrogate.  That is
why they belong in this boundary chapter: a projected or reconstructed transport is
not automatically the same object as the full-state dense coupling from the earlier
OT chapters.  The first interpretive question is therefore not only teacher
preservation.  It is what projected or reconstructed transport object is now acting
on the filtering ensemble.  In the running toy spirit, the key contrast is: the
reference lane earlier in the OT block acted through one full-state transport object,
whereas a sliced route acts through many lower-dimensional projected ones whose
average or reconstruction must still be interpreted with care.  If the recurring
scalar \(\widetilde L\) is built from those projected outputs, then the gradient is
only a gradient of that projected-object scalar, not automatically of the earlier
dense reference scalar.

\subsection{Sliced, subspace, and localization families as target-changing lanes}
\label{sec:bf-neural-ot-dynamic-sliced-localized}

The scalable-OT survey gives this chapter a more precise boundary around several
important high-dimensional families.  Sliced and subspace OT methods attack high
dimension by projecting the measures onto one-dimensional directions or selected
subspaces before solving lower-dimensional transport problems.  Localization and
block-structured OT instead exploit state-space structure by transporting local
blocks or using localized costs rather than one global cloud-wide coupling.

These methods are mathematically important, but in the BayesFilter monograph they
should not be described as mere compressed versions of the dense entropic teacher.
A projected one-dimensional or subspace transport, or a block-localized transport,
is generally a different transport object from the full-state dense coupling used
in the previous chapters.  That means the main burden here is target
interpretation rather than just solver engineering.

In that sense, sliced/subspace and localization families fit the present chapter
better than the retained-teacher chapter.  Their primary promise is not simply to
reach the same teacher object faster.  Their promise is to define a more tractable
transport rule in high dimension.  The price is that the resulting output cloud
must usually be read as a new resampling approximation unless a separate argument
shows how the projected or localized transport reconstructs the intended
full-state object.

This chapter therefore treats these families as target-changing or
structure-changing candidates by default.  They may still be useful and even
necessary in high dimension, but their first audit question is no longer only
teacher-preservation residual.  It is: what transport object is now being applied
to the filtering ensemble, and what scalar target would a later optimizer or HMC
routine actually differentiate if this object were embedded downstream?

\paragraph{Structured families: localization and block transport.}
Localization and block-structured OT instead exploit state-space structure by
transporting local blocks or using localized costs rather than one global cloud-
wide coupling.  A schematic block decomposition is
\begin{equation}
\label{eq:bf-neural-ot-dynamic-block-state}
  x_i=(x_i^{(1)},\ldots,x_i^{(B)}),
\end{equation}
followed by either blockwise transports
\begin{equation}
\label{eq:bf-neural-ot-dynamic-block-transport}
  P^{(b)}\in\Pi(a^{(b)},b^{(b)})
\end{equation}
or shared particle-level couplings with additive local costs
\begin{equation}
\label{eq:bf-neural-ot-dynamic-block-cost}
  C_{ij}
  =
  \sum_{b=1}^B C_{ij}^{(b)}.
\end{equation}
The first option says: transport each block with its own local rule.  The second
says: keep one particle-level coupling, but define that coupling through a cost
that is itself localized across scientifically meaningful state blocks.  In plain
language, the computed object may therefore be a structured local transport rule
rather than one dense global coupling.  This can be scientifically sensible when
the state decomposition itself is justified, but it still changes the object whose
scalar consequences downstream code will inherit.  So the reader must ask not only
whether the method is faster or more stable, but which structured transport object
is actually being applied and what stronger dense-parity claim is therefore no
longer automatic.

\paragraph{Boundary markers only.}
Several additional families remain useful only as boundary markers here:
measure-to-measure regression operators, local Wasserstein regression, and
manifold or Riemannian neural OT
\citep{vandergrift2026m2mtransformers,girshfeld2025nlwr,micheli2026rnot}.  They
show how quickly the learned object can drift from the current finite-cloud
Euclidean teacher, but they do not by themselves answer the present implementation
question.

\section{Same-scalar rule revisited}
\label{sec:bf-neural-ot-dynamic-same-scalar}

\paragraph{What the remaining sections now do.}
The previous sections identified richer families by the object they learn and by
how quickly that object can drift away from the earlier dense reference route.  The
remaining sections do not introduce new families.  They instead collect the audit
questions that help the reader decide what interpretation remains mathematically
supported once the learned object is richer than a static endpoint map.

\paragraph{Mini audit example.}
Suppose a learner reports a gradient for a scalar that uses both the terminal cloud
and the path-action penalty from the running example.  If the code used to compute
the scalar includes \(\mathcal A(x(\cdot))\) but the reported gradient differentiates
only the terminal map \(T(x)=x(1)\), then same-scalar parity has failed even if the
endpoint cloud looks correct.  The strongest justified interpretation in that case
is only that the endpoint map is reasonable as a transport summary; it is not that
the reported gradient is the derivative of the actual executed scalar.  The same
logic applies to operator learners and projected/localized families: once the
filtering-style scalar uses the richer object, the audit must stay tied to that same
object all the way through value and gradient evaluation.

\section{How to audit a richer learned object}
\label{sec:bf-neural-ot-dynamic-promotion}

For richer learned objects, a few mathematical consistency checks are especially
important:
\begin{enumerate}[label=(\roman*)]
  \item whether endpoint or pathwise improvement is accompanied by stability of
  the downstream scalar,
  \item whether the learned object behaves robustly under weight, dimension, or
  particle-count shift,
  \item whether a reported gradient is the gradient of the same scalar value that
  is later interpreted downstream,
  \item whether any claim of target preservation is supported by an explicit
  correction mechanism rather than by smoothness or speed alone.
\end{enumerate}
These are not governance gates.  They are simply the mathematical checks needed to
understand what object is being computed, what scalar is built from that object,
and what interpretation remains identifiable from the reported gradient.

\begin{algorithm}[ht]
\caption{\texttt{audit\_dynamic\_transport\_target\_contract}}
\begin{algorithmic}[1]
\Require learned output object (map, path, or operator), downstream scalar-construction routine, reported gradient routine, optional teacher or corrected baseline scalar, shift-test suite, and declared tolerances for scalar and gradient mismatch
\Ensure a consistency report containing same-scalar parity status, shift-robustness status, any failed conditions, and the strongest mathematically supported interpretation of the learned object
\For{each evaluation problem in the declared suite}
  \State Construct the learned output object for that problem
  \State Compute the downstream scalar \(\widetilde L\) from that learned object
  \State Compute the reported gradient of that same scalar
  \If{a teacher or corrected baseline exists}
    \State Compute the baseline scalar and, where applicable, the baseline gradient
    \State Record scalar and gradient gaps relative to the baseline
  \EndIf
  \State Check that the reported gradient is tied to the same scalar
  \(\widetilde L\)
  \State Evaluate shift robustness under particle-count, weight, and context
  variation
  \State Record any failed scalar, gradient, or robustness condition
\EndFor
\State Return the full consistency report and the strongest interpretation of the
learned object consistent with those results
\end{algorithmic}
\end{algorithm}

\paragraph{Reader checkpoint.}
This procedure does not approve or reject a method in an institutional sense.  It
simply organizes the chapter's local question: what object was learned, what scalar
was built from it, and what gradient interpretation is still justified for that
same scalar?

\section{Relation to the full HMC target chapter}
\label{sec:bf-neural-ot-dynamic-handoff}

This chapter stops at the target-status burden created by broader dynamic and
operator learners.  It does not settle the HMC question.  That full taxonomy still
belongs to Chapter~\ref{ch:bf-dpf-hmc-target-suitability}, which interprets the
scalar and gradient contracts rung by rung across the whole DPF ladder.  Put more
sharply: the present chapter settles the local object-identification question,
while the later HMC chapter settles the broader sampler-interpretation question.

\paragraph{OT-block takeaway.}
The OT sequence has now split into three broad implementation postures.  The
entropic OT chapter fixed the semantics-preserving reference lane.  The retained-
teacher chapter kept that object but learned how to reach it faster.  The direct-
map and dynamic/operator chapters then moved into target-changing territory by
learning endpoint maps, paths, or measure-level operators directly.  The next HMC
and debugging chapters therefore do not ask which family is most elegant in the
abstract; they ask what scalar each family induces and what evidence is needed
before any stronger claim is made.

\section{What this chapter does and does not claim}
\label{sec:bf-neural-ot-dynamic-boundary}

This chapter explains why the broadest neural OT families should be read as
secondary research directions in this monograph.  It does \emph{not} claim:
\begin{itemize}
  \item that geodesic, operator, and adjacent measure/manifold families solve the
  same implementation problem as retained-teacher acceleration,
  \item that a richer learned object is automatically more useful for
  BayesFilter,
  \item that these families are already ready for the same implementation status
  as the earlier chapters.
\end{itemize}
Its role is to close the learned-transport part responsibly by keeping the
broadest learned families in view while handing the full target-status question
forward to the HMC interpretation and verification chapters.

\FloatBarrier
