\chapter{ICNN, Brenier Maps, and Monge-Gap Direct Map Learning}
\label{ch:bf-neural-ot-direct-map}

The previous chapter studied the narrowest neural extension of the entropic OT
teacher: retain the teacher and learn how to start its solver better.  The
present chapter changes the question.  Instead of accelerating a declared
solver, it asks how one can learn the transport map itself.

This is a more ambitious route.  The potential gain is obvious: if a good map is
learned directly, one may bypass repeated iterative solves at deployment time.
The cost is equally important: the learned object is no longer an initialization
for a retained teacher.  It is the transport map itself, so every modeling and
training choice now acts directly on the output cloud.

This chapter therefore has one dominant task: teach the reader what direct-map
learning is trying to learn, why Brenier/ICNN and Monge-gap methods are natural
representatives of that lane \citep{amos2017icnn,makkuva2020otmapping,uscidda2023mongegap}, and what evidence is needed once no corrective
teacher solve remains in control.

\section{What changes when the map itself is learned?}
\label{sec:bf-neural-ot-direct-orientation}

The best way to read this chapter is to compare it with the retained-teacher
warm-start chapter.
\begin{itemize}
  \item In the warm-start chapter, the teacher still decides what the final
  transport plan and output cloud are; the learner only helps the teacher solve
  its own problem faster.
  \item In the present chapter, the learner itself proposes the transport map.
  There may be no corrective Sinkhorn phase at all.
\end{itemize}
So the central question is no longer “did we accelerate the same teacher?”  It
is “what map object have we learned, and under what assumptions should we trust
it?”

\paragraph{Plain-language takeaway.}
If the previous chapter was about helping an expensive trusted calculation, this
chapter is about replacing that calculation by a learned map.  That is why the
mathematical structure and the evidence contract both change.

\paragraph{How to read the rest of this chapter.}
The sequence now splits into two direct-map routes.  First comes the
Brenier/ICNN family, which keeps a strong structural prior by learning a convex
potential whose gradient defines the map.  Then comes the Monge-gap family,
which keeps the same direct-map goal but relaxes the structural prior and asks
instead how OT-like the learned endpoint map remains under the chosen cost.
Both routes differ from the previous chapter in the same decisive way: there is
no corrective teacher solve standing behind the final output object.

\section{Running direct-map cell}
\label{sec:bf-neural-ot-direct-running-cell}

A tiny one-dimensional cell is enough to carry the main ideas.  Let the source
be the two-point cloud
\[
  X_{\mathrm{src}}=\{-1,1\}
  \quad\text{with equal weights }\frac{1}{2},
\]
and let the target be
\[
  X_{\mathrm{tgt}}=\{0,2\}
  \quad\text{with equal weights }\frac{1}{2}.
\]
The obvious translation map is
\begin{equation}
\label{eq:bf-neural-ot-direct-shift-map}
  T^\star(x)=x+1.
\end{equation}
Applying \(T^\star\) to the source cloud sends \(-1\mapsto 0\) and
\(1\mapsto 2\), so the pushforward equals the target cloud exactly.

For quadratic cost,
\begin{equation}
\label{eq:bf-neural-ot-direct-shift-cost}
  \frac{1}{2}\sum_{x\in X_{\mathrm{src}}}|x-T^\star(x)|^2
  =
  \frac{1}{2}(1^2+1^2)=1.
\end{equation}
This simple cell already shows the difference between a direct map and a retained
teacher solve.  The output cloud is obtained directly by evaluating \(T\) on the
source particles; there is no intermediate coupling matrix or balancing loop in
this picture.

A second map on the same endpoint cloud shows why direct-map evidence is more
subtle.  Consider
\begin{align}
\label{eq:bf-neural-ot-direct-good-bad-maps}
  T_{\mathrm{good}}(-1)&=0,
  &T_{\mathrm{good}}(1)&=2,\\
  T_{\mathrm{bad}}(-1)&=2,
  &T_{\mathrm{bad}}(1)&=0.
\end{align}
Both maps reach the same endpoint set \(\{0,2\}\), but for quadratic cost the
good map has cost
\[
  \frac{1}{2}(1^2+1^2)=1,
\]
whereas the bad map has cost
\[
  \frac{1}{2}(3^2+1^2)=5.
\]
So endpoint fit alone does not settle whether the learned map behaves like a
good transport map for the declared cost.

\section{Direct-map lane: object changed, evidence changed}
\label{sec:bf-neural-ot-direct-object}

A direct map-learning method tries to produce a deterministic transport map
\begin{equation}
\label{eq:bf-neural-ot-direct-map-object}
  T:
  \mathbb R^{d_x}\to\mathbb R^{d_x}.
\end{equation}
The source measure \(\mu\) and target measure \(\nu\) are related by the
pushforward relation
\begin{equation}
\label{eq:bf-neural-ot-direct-pushforward}
  T_\#\mu = \nu,
\end{equation}
meaning that if one samples \(X\sim\mu\) and then computes \(T(X)\), the law of
that transformed sample is \(\nu\).  Equivalently, for every test function
\(f\) for which the integrals exist,
\begin{equation}
\label{eq:bf-neural-ot-direct-pushforward-testfn}
  \int f(y)\,d(T_\#\mu)(y)
  =
  \int f(T(x))\,d\mu(x).
\end{equation}
At the empirical-cloud level, this means that if the chapter starts from source
particles \(x_i\sim\mu\) and evaluates \(z_i=T(x_i)\), then the mapped cloud
\(\{z_i\}\) is the finite-sample object being asked to approximate the declared
target law or target cloud.

The important change relative to the warm-start chapter is immediate: the learned
object is no longer an internal solver coordinate.  It is the transport map
itself, or a potential from which that map is derived.  In the vocabulary of the
OT block, retained-teacher learning changes the execution path while a corrected
solve still returns to a declared transport object.  Direct-map learning changes
the transport object directly and therefore inherits a different evidence
contract.

\begin{center}
\small
\begin{tabular}{p{0.24\textwidth} p{0.23\textwidth} p{0.41\textwidth}}
\toprule
Quantity & Status in this chapter & Practical burden \\
\midrule
learned map \(T\) & primary learned object & must be validated directly, not through a retained teacher residual \\
pushforward \(T_\#\mu\) & endpoint-law target & should match the declared target law or cloud closely enough on held-out problems \\
transport-cost / Monge-gap behavior & auxiliary OT-structure evidence & helps distinguish a plausible transport map from an arbitrary endpoint fit \\
downstream scalar built from mapped particles & implementation consequence & must be checked for drift once the map is embedded in filtering or HMC-adjacent code \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Reader checkpoint.}
The direct-map route is not just a stronger warm start.  It asks the learned map
itself to carry the approximation burden.

\paragraph{Reader map for the technical core.}
Before moving deeper into the equations, it helps to keep the two direct-map routes
side by side.

\begin{center}
\small
\begin{tabular}{p{0.20\textwidth} p{0.24\textwidth} p{0.24\textwidth} p{0.20\textwidth}}
\toprule
Route & Learned object & Training signal & Evidence required \\
\midrule
Brenier / ICNN & convex potential \(\psi_\theta\) and endpoint map \(T_\theta=\nabla\psi_\theta\) & potential-based objective built from mapped source samples and target-side scoring & endpoint fit, OT-structure plausibility, downstream scalar stability \\
Monge-gap & direct endpoint map \(T_\theta\) & endpoint-fit term plus empirical Monge-gap surrogate & endpoint fit, Monge-gap behavior, downstream scalar stability \\
\bottomrule
\end{tabular}
\end{center}

The rest of the chapter should be read through those three columns: what object is
learned, what training signal drives it, and what evidence would justify trusting
that object in later filtering code.

\section{Brenier viewpoint}
\label{sec:bf-neural-ot-direct-brenier}

In the quadratic-cost Euclidean setting, a classical route is to represent the
map as the gradient of a convex potential \citep{villani2003topics,peyre2019computational,vesseron2024brenier}:
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier}
  T^\star(x)=\nabla\psi^\star(x),
  \qquad
  \psi^\star\text{ convex}.
\end{equation}
The quickest intuition is geometric.  Quadratic-cost transport prefers maps with a
monotone or, more precisely, cyclically monotone structure.  Gradients of convex
potentials are the canonical way to realize that structure.  In one dimension this
reduces to the familiar monotone rearrangement picture: sort the mass from left to
right and move it without crossings.  In higher dimension the theorem-strength
statement is subtler, but the guiding lesson is the same: the more the transport is
forced to behave like a monotone gradient field, the more a convex potential
becomes a natural object to learn.  This is why the Brenier viewpoint appears so
naturally: instead of learning an arbitrary map, learn a scalar potential whose
gradient already lives in a mathematically privileged map family.

This statement should be read with its theorem-level assumptions attached.  The
classical Brenier characterization is a continuous-measure result for quadratic
cost in Euclidean space under regularity conditions such as finite second
moments and absolute continuity of the source measure with respect to Lebesgue
measure.  In the finite empirical-cloud setting used throughout BayesFilter, a
Monge map may fail to exist or fail to be unique in that theorem-strength sense.
So the present chapter uses the Brenier viewpoint primarily as structured
motivation for a learned map family, not as a blanket theorem that every finite
cloud transport problem in the particle-filter setting is exactly represented by
a gradient of a convex potential.

In one dimension, this means the derivative of the potential defines the map.
For the translation cell above, one convenient choice is
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier-example}
  \psi^\star(x)=\frac{1}{2}x^2+x.
\end{equation}
Then
\begin{equation}
\label{eq:bf-neural-ot-direct-brenier-example-grad}
  \nabla\psi^\star(x)=x+1=T^\star(x).
\end{equation}
The second derivative of \(\psi^\star\) is \(1\), so the potential is convex.
This tiny calculation is the cleanest way to see what the Brenier/ICNN family is
trying to exploit: instead of learning the map directly, it learns a structured
potential whose gradient gives the map.

A small finite-cloud warning is useful here.  Even in tiny empirical problems, the
transport object may fail to be a single theorem-strength Monge map.  If one source
atom of mass \(1\) must feed two target atoms of mass \(1/2\) each, the natural
finite-cloud transport may split mass across targets rather than assign one unique
endpoint to that source atom.  This is exactly why the chapter treats Brenier as a
structured motivation ladder rather than as an automatic finite-cloud theorem.

\paragraph{What survives from Brenier theory in this chapter.}
The theorem-strength Brenier result is not being imported wholesale into the
finite-cloud BayesFilter setting.  What survives here is a narrower bridge.  The
quadratic-cost theory says that convex-potential gradients are a mathematically
natural map family when the continuous assumptions hold.  The monograph then uses
that fact as \emph{structured motivation} for a direct-map parameterization.  What
does \emph{not} survive automatically is a theorem that every finite empirical-cloud
transport problem in this particle-filter setting is already solved exactly by an
ICNN potential.  The reader should therefore carry forward a three-way separation:
continuous-theorem support, structured finite-cloud motivation, and the separate
empirical question of whether the learned map behaves well once embedded in the
actual BayesFilter computation.

\section{ICNN/Brenier route}
\label{sec:bf-neural-ot-direct-icnn}

Input-convex neural networks (ICNNs) provide a parametric family of scalar
functions that remain convex in designated inputs \citep{amos2017icnn,makkuva2020otmapping}.  The implementation idea is:
\begin{enumerate}[label=(\roman*)]
  \item parameterize the potential \(\psi\) by an ICNN \(\psi_\theta\),
  \item obtain the transport map by differentiating the network,
  \(
    T_\theta(x)=\nabla\psi_\theta(x),
  \)
  \item train \(\psi_\theta\) so that the induced map sends source mass toward
  the target according to the chosen OT objective.
\end{enumerate}

The middle step is the important one to slow down.  The learned object is not yet
a teacher-side coupling or a corrected barycentric cloud.  It is the potential
\(\psi_\theta\), together with the gradient-defined endpoint map
\(T_\theta=\nabla\psi_\theta\) that later code will actually apply to the source
particles.  So the training objective is not certifying ``the same object as the
retained teacher, only faster.''  It is certifying, at best, that this chosen map
family behaves plausibly under the declared direct-map evidence contract.

A useful mathematical picture to keep in view is not one universal training
formula, but the structural relation
\begin{equation}
\label{eq:bf-neural-ot-direct-conjugate}
  \psi^*(y)=\sup_x\{x^\top y-\psi(x)\},
\end{equation}
between a convex potential and its convex conjugate.  In theorem-strength
quadratic-cost OT, Brenier structure motivates learning a convex potential whose
gradient gives the map.  In practical finite-cloud implementations, however,
several objective families are used: dual or conjugate-based objectives,
empirical pushforward penalties, or regularized map-learning surrogates.  The
present chapter therefore uses the Brenier/ICNN viewpoint primarily as
structured motivation for the map family, not as a claim that one displayed
objective automatically covers every finite-cloud training loop.

One good way to read the displayed objective is to track the roles of its two
functions separately.  The network parameter \(\theta\) chooses the source-side
potential \(\psi_\theta\), whose gradient becomes the deployed map.  The function
\(\varphi\) plays the complementary target-side role: it scores where the mapped
samples land and therefore transmits target information back into the optimization.
The formula is not saying that every direct-map implementation literally fixes this
\(\varphi\) in advance.  It is saying that one common way to train a Brenier-style
map is to balance a source-to-map transport cost against a target-side scoring term
that rewards mapped samples for matching the intended target law.  For a practical
reading, a first-year reader can think of \(\varphi\) as the target-side scoring
object in the objective: whether it is held fixed, updated alternately, or learned
jointly depends on the exact implementation, but its mathematical role in the
displayed formula is always to encode how good the mapped samples look from the
target side.

The attraction of this route is that the map is not a black-box regressor.  It
is constrained to come from a convex potential, which is exactly the sort of
structure the quadratic-cost theory suggests under its stated assumptions.

A concrete canonical training objective helps fix ideas.  One representative
potential-based route is to choose a target-side potential \(\varphi\) and solve
an objective of the schematic form
\begin{equation}
\label{eq:bf-neural-ot-direct-icnn-objective}
  \min_\theta
  \mathbb E_{X\sim\mu}\!
  \left[
    \frac{1}{2}\|X-\nabla\psi_\theta(X)\|^2
    - \varphi(\nabla\psi_\theta(X))
  \right]
  +
  \mathbb E_{Y\sim\nu}[\varphi(Y)].
\end{equation}
This should not be read as the only ICNN objective in the literature.  Its role in
this chapter is narrower: it makes explicit what the learner is optimizing.  The
first expectation penalizes the source-to-map transport cost while evaluating the
target-side potential on the mapped samples.  The second expectation anchors the
target distribution through samples from \(\nu\).  At deployment time the object
used by later code is still the gradient-defined endpoint map
\(T_\theta=\nabla\psi_\theta\), not the training objective itself.  The objective
therefore certifies only what this direct-map learner was trained to do; it does
not by itself recreate the retained-teacher guarantee from the previous chapter.
A first-year reader should therefore read the objective in four rungs:
\begin{enumerate}[label=(\roman*)]
  \item choose a convex potential family \(\psi_\theta\),
  \item obtain the endpoint map by differentiation,
  \item optimize a loss built from mapped samples and target-side information,
  \item deploy only the resulting endpoint map, not the training loss itself.
\end{enumerate}

The cost is also immediate from the equations.  One must evaluate gradients of
the potential and, depending on the chosen training formulation, may need to
evaluate or approximate conjugate or dual terms.  So the route trades a
retained transport solver for a more structured but more delicate
map-learning objective.

\begin{algorithm}[ht]
\caption{\texttt{train\_icnn\_brenier\_map}}
\begin{algorithmic}[1]
\Require source-cloud sampler for \(\mu\), target-cloud sampler for \(\nu\), ICNN potential family \(\psi_\theta\), declared quadratic-cost/Brenier setting, convexity-enforcement policy, optimizer, learning-rate schedule, batch size, stopping rule, and held-out validation clouds
\Ensure trained potential \(\psi_\theta\), induced map \(T_\theta(x)=\nabla\psi_\theta(x)\), and validation diagnostics for pushforward mismatch and scalar drift
\State Initialize ICNN parameters \(\theta\) and optimizer state
\While{the stopping rule has not been met}
  \State Draw minibatches \(x_b\sim\mu\) and \(y_b\sim\nu\)
  \State Evaluate the potential values \(\psi_\theta(x_b)\)
  \State Evaluate the gradient-defined transport map \(T_\theta(x_b)=\nabla\psi_\theta(x_b)\)
  \State Compute the declared dual or potential-based training objective, including any required convex-conjugate or surrogate terms
  \State Add any declared regularization enforcing smoothness, stability, or numerical tractability
  \State Backpropagate the total loss and update \(\theta\)
  \State Re-enforce convexity according to the declared ICNN policy if it is not built directly into the parameterization
\EndWhile
  \State On held-out validation clouds, compute mapped particles \(T_\theta(x_i)\)
  \State Evaluate endpoint pushforward mismatch, transport-cost proxy, and downstream scalar sensitivity on the held-out clouds
  \State Return \(\psi_\theta\), the induced map \(T_\theta\), and the declared diagnostics
\end{algorithmic}
\end{algorithm}

\paragraph{Reader checkpoint.}
Three evidential layers should remain separate for this route:
\begin{enumerate}[label=(\roman*)]
  \item whether the learned map reaches the right endpoint law or cloud well enough,
  \item whether the learned map still behaves like a plausible OT-style map under
    the declared cost and structural prior,
  \item whether a downstream scalar built from the mapped particles remains stable
    once this learned map is inserted into filtering or HMC-adjacent code.
\end{enumerate}
A good training loss or good endpoint fit may support the first layer without yet
settling the second or third.

\paragraph{Approximation enters here.}
The learned object is the transport map itself.  Errors in the learned potential
or its gradient directly change the transported output cloud.

\paragraph{Why the next section changes emphasis rather than object.}
The Brenier/ICNN route and the Monge-gap route are both direct-map answers.  The
learned endpoint map is still the primary object in both cases.  What changes in
the next section is not the endpoint ambition but the structural prior: instead of
forcing the map to arise from one especially rigid potential family, the chapter
asks how to penalize a candidate map when it behaves unlike an OT map under the
chosen cost.

\section{Monge-gap route}
\label{sec:bf-neural-ot-direct-monge-gap}

A more flexible direct-map family regularizes a candidate map by how far it
behaves from an OT map rather than forcing a particular architecture from the
start \citep{uscidda2023mongegap}.  For cost \(c\) and reference measure \(\rho\), the Monge gap is
\begin{equation}
\label{eq:bf-neural-ot-direct-monge-gap}
  \mathcal M_\rho^c(T)
  =
  \int c(x,T(x))\,d\rho(x)-W_c(\rho,T_\#\rho).
\end{equation}
Here \(\rho\) is an explicitly declared reference measure used to probe how
OT-like the learned map is on the region of state space that matters for the
intended application.  In the simplest reading one may take \(\rho=\mu\), so
the regularizer is applied on the same source measure used for transport.  But a
chapter, experiment, or implementation may instead choose a different reference
measure to emphasize a wider training envelope or a different weighting of the
state space.  The role of \(\rho\) must therefore be named explicitly rather than
left implicit.

The key structural property is immediate from the definition.  The deterministic
map \(x\mapsto T(x)\) induces a coupling between \(\rho\) and \(T_\#\rho\), so by
the definition of optimal transport,
\begin{equation}
\label{eq:bf-neural-ot-direct-monge-gap-nonnegative}
  W_c(\rho,T_\#\rho)
  \le
  \int c(x,T(x))\,d\rho(x).
\end{equation}
In slightly fuller words: the map-induced coupling
\(\gamma_T=(\operatorname{Id},T)_\#\rho\) is one admissible coupling between the
source measure and the pushforward target.  The optimal-transport cost
\(W_c(\rho,T_\#\rho)\) is the infimum over all admissible couplings with those
same marginals, so it cannot exceed the cost of this specific map-induced one.
Therefore
\begin{equation}
\label{eq:bf-neural-ot-direct-monge-gap-nonnegative-conclusion}
  \mathcal M_\rho^c(T)\ge 0,
\end{equation}
with equality when the map-induced coupling is already optimal for transporting
\(\rho\) to \(T_\#\rho\).  This is why the Monge gap is a meaningful regularizer:
it measures how far the learned map behaves from an OT map under the chosen cost.

As a mathematical definition, the Monge gap is exact.  But in a practical
training loop, the second term is typically not available in closed form because
the pushforward measure \(T_\#\rho\) itself depends on the current learned map.
So the reader should picture a four-step ladder:
\begin{enumerate}[label=(\roman*)]
  \item define the exact Monge-gap quantity mathematically,
  \item recognize that the OT reference term is usually unavailable in closed form
    inside training,
  \item replace that unavailable term by a declared estimator or surrogate,
  \item interpret the resulting optimization only as strongly as that surrogate
    justification allows.
\end{enumerate}
A concrete empirical surrogate is obtained by drawing reference samples
\(x_i\sim\rho\), forming mapped particles \(z_i=T_\theta(x_i)\), and defining
\begin{equation}
\label{eq:bf-neural-ot-direct-monge-gap-empirical}
  \widehat{\mathcal M}_{\rho,m}^c(T_\theta)
  =
  \frac{1}{m}\sum_{i=1}^m c(x_i,z_i)
  -
  W_c\!\left(\frac{1}{m}\sum_{i=1}^m\delta_{x_i},
  \frac{1}{m}\sum_{i=1}^m\delta_{z_i}\right).
\end{equation}
The second term is now a batchwise discrete OT problem on the empirical reference
cloud and its mapped image.  This is now a finite-sample training quantity, not the
exact population Monge gap.  So any finite-sample training objective that uses an
estimated OT reference term should be read as a surrogate or approximate Monge-gap
objective unless the estimator and its properties are stated explicitly.
Operationally, the chapter’s point is: fit a useful map, but penalize it when it
behaves unlike an OT map for the chosen cost.  A useful mental model is that the
training loop is differentiating through two finite objects at once: the mapped
sample cloud and a discrete OT computation on that cloud.  If either one is a poor
proxy for the intended downstream use case, the empirical Monge-gap objective may
still decrease while the later filtering scalar drifts in an undesired direction.

The running cell already shows the idea.  The good and bad endpoint maps above
both match the endpoint law, but the bad map moves mass much less efficiently.  On
that tiny example, the direct transport cost of the good map is
\(\frac{1}{2}(1^2+1^2)=1\), while the bad map has cost
\(\frac{1}{2}(3^2+1^2)=5\).  So even before worrying about population measures or
finite-sample estimators, the Monge-gap logic is already visible: two maps can fit
the same endpoint cloud while differing sharply in how OT-like their transport
behavior is under the chosen cost.  The Monge-gap route is attractive precisely
because it can discourage such maps without forcing the full ICNN/Brenier
structural commitment.

\begin{algorithm}[ht]
\caption{\texttt{train\_monge\_gap\_regularized\_direct\_map}}
\begin{algorithmic}[1]
\Require source-cloud sampler for \(\mu\), target-cloud sampler for \(\nu\), reference-measure sampler for \(\rho\), parametric map family \(T_\theta\), cost function \(c\), endpoint-fit objective, Monge-gap weight \(\lambda_{\mathrm{MG}}\), optimizer, stopping rule, and held-out validation suite
\Ensure trained direct map \(T_\theta\), Monge-gap diagnostic history, endpoint-fit diagnostics, and downstream scalar sensitivity report
\State Initialize map parameters \(\theta\) and optimizer state
\While{the stopping rule has not been met}
  \State Draw minibatches from the source, target, and reference distributions
  \State Compute mapped source particles \(z_i=T_\theta(x_i)\)
  \State Evaluate the declared endpoint-fit loss between the mapped source batch and the target batch
  \State Compute the empirical transport-cost term on the reference samples from \(\rho\)
  \State Estimate the OT reference term needed for the empirical Monge-gap penalty; this estimate defines the practical training surrogate unless a stronger equivalence result is supplied
  \State Form the total loss as endpoint-fit term plus \(\lambda_{\mathrm{MG}}\) times the empirical Monge-gap penalty, together with any declared regularization
  \State Backpropagate the total loss and update \(\theta\)
\EndWhile
  \State On held-out validation clouds, evaluate endpoint mismatch, Monge-gap magnitude, transport-cost proxy, and downstream scalar drift
  \State Return the trained map \(T_\theta\) and the declared diagnostics
\end{algorithmic}
\end{algorithm}

\paragraph{Reader checkpoint.}
The Monge-gap route still learns a direct map, but it relaxes the strong
ICNN-only structural commitment.  The gain is flexibility.  The cost is that
the learned map is now judged relative to a regularized OT-like behavior rather
than by a retained teacher solver.

\section{Structured variants and boundaries}
\label{sec:bf-neural-ot-direct-structured}
\label{sec:bf-neural-ot-direct-scalable-boundary}

The same direct-map logic can be combined with additional structural priors,
such as sparse displacement or factorized transport structure
\citep{forrow2019factored,chenxiezhang2024sicnn,xu2025flowot,geuter2025unot}.  These are worth
keeping in view because they may be useful when the filtering update itself has
a scientifically justified sparse or low-active-subspace geometry.  But they
inherit the same chapter-wide burden: the learned endpoint map itself must be
validated, rather than interpreted through a retained corrective teacher.

This is why the direct-map route should be read as a different scalable-OT
answer from retained-teacher acceleration.  It is more radical than
teacher-preserving factorized Sinkhorn, because the final map object itself is
learned.  But it is still narrower than the dynamic or operator families of the
next chapter, because it aims to learn one endpoint map rather than an entire
transport path or measure-level operator.  The next chapter therefore does not
continue the same endpoint-map story at larger scale; it changes the learned
object again, from one final map to richer path, velocity-field, or operator
representations.

\begin{center}
\small
\begin{tabular}{p{0.18\textwidth} p{0.20\textwidth} p{0.22\textwidth} p{0.27\textwidth}}
\toprule
Family & Learned object & Corrective teacher retained? & First burden of proof \\
\midrule
Retained-teacher acceleration & solver state or latent teacher coordinate & yes & teacher-preservation residual after correction \\
Direct-map learning & endpoint map or potential-derived map & no & endpoint, OT-structure, and downstream scalar drift \\
Dynamic/operator learners & path, velocity field, or operator family & no & target-status and same-scalar coherence across richer objects \\
\bottomrule
\end{tabular}
\end{center}

\section{Why this is not the default first route}
\label{sec:bf-neural-ot-direct-why-not-default}

The direct-map families are mathematically serious and potentially powerful, but
they are not the monograph’s default first implementation route for two reasons:
\begin{enumerate}[label=(\roman*)]
  \item they learn the final map object itself rather than only accelerating a
  declared teacher solve,
  \item their strongest mathematical support depends on more specific geometric
  assumptions than the retained-teacher acceleration route requires.
\end{enumerate}

\subsection{Direct-map families as a different scalable-OT answer}
\label{sec:bf-neural-ot-direct-scalable-boundary}

The scalable-OT survey makes the boundary with the retained-teacher chapter even
clearer.  Nystrom, positive-feature, and related entropic approximations begin by
asking how to preserve or approximate a declared teacher transport object more
cheaply.  The direct-map route asks a different question: can one learn the final
transport map itself well enough that repeated teacher solves are no longer the
center of the method?

That difference is not merely architectural.  It changes the meaning of the main
residual.  In the retained-teacher lane, the first residual is naturally a
teacher-preservation discrepancy.  In the direct-map lane, the primary residuals
are pushforward mismatch, Monge-gap behavior, transport-cost proxy, and scalar
drift under the learned map.  So this chapter should be read as a target-boundary
chapter: it explains why a direct-map learner is not just ``a stronger warm
start,'' but a new answer to the scalable-OT problem with a different evidence
contract.

In practical terms, this chapter occupies the middle position between the
retained-teacher approximation lane and the broader replacement/dynamic lane.  It
is more radical than teacher-preserving factorized Sinkhorn, because the map
itself is learned.  But it is still narrower than the dynamic/operator families of
the next chapter, because it aims to learn one endpoint map rather than an entire
transport path or measure-level operator.

\section{Verification and non-claims}
\label{sec:bf-neural-ot-direct-verification}

A minimal implementation-facing reading of this chapter should distinguish:
\begin{itemize}
  \item what map object is being fit,
  \item what assumptions justify that map family,
  \item what endpoint, pushforward, transport-cost, or Monge-gap residuals are
  measured,
  \item what downstream scalar drift is checked once the learned map is embedded
  in filtering or HMC-adjacent code.
\end{itemize}

The practical residuals differ from the retained-teacher lane because no
corrective Sinkhorn phase is automatically present to return to a declared
teacher object.  So endpoint fit, pushforward mismatch, transport-cost proxy,
Monge-gap proxy, and downstream scalar sensitivity are part of the primary
evidence, not side diagnostics.

This chapter does \emph{not} claim:
\begin{itemize}
  \item that direct learned maps are interchangeable with retained-teacher EOT
  acceleration,
  \item that a map-level fit automatically preserves the same filtering scalar,
  \item that broader direct-map families are already validated for BayesFilter.
\end{itemize}
Its role is to slow down the direct-map family itself so the reader can see that
it is solving a different problem from the previous chapter.

\FloatBarrier
