\chapter[High-Dimensional Filtering Foundations]{High-Dimensional Nonlinear Filtering Foundations}
\label{ch:bf-highdim-nonlinear-foundations}

\section{The Filtering Problem In Its Least Forgiving Form}
\label{sec:bf-highdim-foundations-problem}

A high-dimensional nonlinear filter is not merely a large Kalman filter.  It is
an approximation to a sequence of conditional probability laws whose update
can become singular, multimodal, sharply curved, or numerically inaccessible.
This chapter fixes the mathematical objects that the later chapters approximate.
It also separates exact identities from approximation policies.  That separation
matters because a filter used inside optimization or Hamiltonian Monte Carlo
does not only produce state estimates; it defines a scalar likelihood target and
possibly a gradient of that target.

Let $\theta\in\Theta$ be static parameters, $x_t\in\R^n$ a latent state, and
$y_t\in\R^m$ an observation.  In discrete time a state-space model is
\begin{align}
  x_0 &\sim \mu_\theta, &
  x_t\mid x_{t-1},\theta &\sim f_\theta(\cdot\mid x_{t-1}), &
  y_t\mid x_t,\theta &\sim g_\theta(\cdot\mid x_t).
  \label{eq:bf-hd-ssm}
\end{align}
The continuous-time diffusion analogue, used by the Zakai/DMZ and tensor-train
PDE papers, is commonly written
\begin{align}
  dX_t &= a_\theta(X_t,t)\,dt + B_\theta(X_t,t)\,dV_t, \\
  dY_t &= h_\theta(X_t,t)\,dt + dW_t,
  \label{eq:bf-hd-continuous-ssm}
\end{align}
with variants for correlated noise.  The discrete recursion is the natural
contract for production filtering APIs; the continuous formulation exposes the
PDE and pathwise robust methods in \citet{davis1980multiplicative},
\citet{yau2000realtime}, \citet{yau2008realtime},
\citet{li2019ttfiltering}, and \citet{meng2025regularity}.

For a reader coming from numerical analysis, physics, or computational
chemistry, the central object is the conditional law
\(p_\theta(x_t\mid y_{1:t})\).  The input object at time \(t\) is the previous
conditional law and a new observation; the output object is a new conditional
law plus the normalizer \(Z_t\).  The exact target is the Bayes recursion below.
Every later method is an approximation to one of three objects: the law
\(p_\theta(x_t\mid y_{1:t})\), the likelihood scalar
\(\ell_T(\theta)\), or a derivative of that scalar.  A diagnostic is useful only
when it says which of these objects it is checking.

\paragraph{Running quadratic-observation cell.}
The same teaching example will be used through
Chapters~\ref{ch:bf-highdim-nonlinear-foundations}--\ref{ch:bf-highdim-candidate-synthesis}:
\begin{equation}
  x_t=\rho x_{t-1}+\eta_t,\qquad
  \eta_t\sim\calN(0,Q),\qquad
  y_t=x_t^2+\epsilon_t,\qquad
  \epsilon_t\sim\calN(0,R).
  \label{eq:bf-hd-running-cell}
\end{equation}
It is deliberately small.  Its purpose is to make the objects visible, not to
validate a macro-finance model.  With a one-step Gaussian prediction
\(X\sim\calN(0,P^-)\), the filtering density after observing \(y>0\) is
proportional to
\begin{equation}
  \exp\!\left\{-\frac{x^2}{2P^-}
  -\frac{(y-x^2)^2}{2R}\right\}.
  \label{eq:bf-hd-running-posterior}
\end{equation}
When \(R\) is small and \(y\) is away from zero, the likelihood rewards states
near both \(+\sqrt y\) and \(-\sqrt y\).  The exact filtering object is
therefore a conditional law that may have two lobes.  A mean alone, a covariance
alone, or a finite scalar likelihood alone is only one view of that object.

\small
\begin{longtable}{@{}p{0.20\linewidth}p{0.36\linewidth}p{0.34\linewidth}@{}}
\toprule
Object & Role in the exact recursion & Why the running cell exposes it \\
\midrule
\endhead
Conditional law \(p_t\) & Probability law of \(x_t\) given observations &
\(x^2\) observations can create two separated lobes. \\
Normalizer \(Z_t\) & One-step predictive likelihood contribution &
It integrates both lobes and is the scalar used in inference. \\
Likelihood \(\ell_T\) & Sum of log normalizers over time & Changing \(Z_t\)
changes the parameter posterior. \\
Sensitivity \(s_t\) & Derivative of the law or normalizer with respect to
parameters & HMC needs the gradient of the same scalar it reports. \\
\bottomrule
\end{longtable}
\normalsize

\begin{assumption}[Dominated filtering model]
\label{ass:bf-hd-dominated}
The kernels in \eqref{eq:bf-hd-ssm} admit densities with respect to reference
measures, the predictive normalizing constants below are finite and positive on
the observed path, and every differentiation under an integral sign is justified
by a local dominating function.  Continuous-time PDE statements additionally
use the regularity, ellipticity, growth, and boundary assumptions stated in the
cited source.
\end{assumption}

\section{Exact Discrete-Time Recursion And Likelihood}
\label{sec:bf-hd-exact-recursion}

Under Assumption~\ref{ass:bf-hd-dominated}, prediction and update are
\begin{align}
  p_\theta(x_t\mid y_{1:t-1})
    &= \int f_\theta(x_t\mid x_{t-1})
       p_\theta(x_{t-1}\mid y_{1:t-1})\,dx_{t-1},
       \label{eq:bf-hd-prediction}\\
  p_\theta(x_t\mid y_{1:t})
    &= \frac{g_\theta(y_t\mid x_t)p_\theta(x_t\mid y_{1:t-1})}
            {Z_t(\theta)},\qquad
  Z_t(\theta)=\int g_\theta(y_t\mid x_t)
       p_\theta(x_t\mid y_{1:t-1})\,dx_t .
       \label{eq:bf-hd-update}
\end{align}
The likelihood factors as
\begin{equation}
  p_\theta(y_{1:T})=\prod_{t=1}^T Z_t(\theta),\qquad
  \ell_T(\theta)=\sum_{t=1}^T \log Z_t(\theta).
  \label{eq:bf-hd-likelihood-factor}
\end{equation}
The derivation is short, but it is the contract that every later approximation
must respect.  Write the joint density through time \(t\) as
\[
  p_\theta(x_{0:t},y_{1:t})
  =
  \mu_\theta(x_0)
  \prod_{s=1}^t f_\theta(x_s\mid x_{s-1})
  \prod_{s=1}^t g_\theta(y_s\mid x_s).
\]
Integrating out \(x_{0:t-1}\) after conditioning on \(y_{1:t-1}\) gives
\eqref{eq:bf-hd-prediction}.  Multiplying the predictive density by the
likelihood \(g_\theta(y_t\mid x_t)\) gives an unnormalized density
\[
  \gamma_t(x_t)=g_\theta(y_t\mid x_t)
       p_\theta(x_t\mid y_{1:t-1}).
\]
Its integral is exactly the one-step predictive density of the observation,
\[
  Z_t(\theta)=p_\theta(y_t\mid y_{1:t-1}).
\]
Bayes' formula then gives \eqref{eq:bf-hd-update}.  Finally, the chain rule for
the observation density gives
\[
  p_\theta(y_{1:T})
  =p_\theta(y_1)\prod_{t=2}^T p_\theta(y_t\mid y_{1:t-1})
  =\prod_{t=1}^T Z_t(\theta).
\]
Thus a filtering implementation that changes the normalizers changes the
likelihood, not merely the state estimate.

For actual stochastic-volatility rows, the target discipline is now explicit:
there is one cumulative likelihood target,
\(
\ell_T(\theta)=\sum_{t=1}^T\log Z_t(\theta)
\),
and every admissible numerical route must approximate that same scalar.  In the
transformed observation parameterization used in the code and later chapters,
\[
  z_t=\log(y_t^2),
  \qquad
  z_t-\log(\beta^2)-h_t\sim \log(\chi_1^2),
\]
so the one-step normalizer is defined by the exact transformed actual-SV
observation law.  Direct-likelihood quadrature, fixed-cloud reweighting, and the
current Zhao--Cui comparator are therefore comparable only because they are meant
to approximate that same normalizer sequence.

The previous actual-SV ``Lane~B'' framing was wrong for this inference problem.
It replaced the transformed actual-SV observation law by a Gaussian-closure
construction built from predictive observation moments of an augmented variable.
That replacement changes the implied one-step normalizer and therefore changes
the cumulative likelihood target itself.  A filter may still be internally
self-consistent for that replacement scalar, but it is not evidence about the
intended actual-SV likelihood in \eqref{eq:bf-hd-likelihood-factor}.  Any
implementation, benchmark, or gradient check must therefore keep two questions
separate: whether a method differentiates its own declared scalar correctly, and
whether that declared scalar is the right target in the first place.

\subsection{Actual-SV Single-Target Derivation}
\label{sec:bf-hd-actual-sv-single-target-derivation}

The corrected actual-SV statement can now be written as a short derivation.
Start from the raw observation equation
\begin{equation}
  y_t=\beta e^{h_t/2}\varepsilon_t,
  \qquad
  \varepsilon_t\sim \calN(0,1),
  \qquad
  \beta>0.
  \label{eq:bf-hd-actual-sv-raw-observation}
\end{equation}
For nonzero observations define the transformed datum
\begin{equation}
  z_t=\log(y_t^2).
  \label{eq:bf-hd-actual-sv-z-definition}
\end{equation}
Squaring \eqref{eq:bf-hd-actual-sv-raw-observation} gives
\(
  y_t^2=\beta^2 e^{h_t}\varepsilon_t^2
\), and taking logs yields
\begin{equation}
  z_t-\log(\beta^2)-h_t
  =
  \log(\varepsilon_t^2).
  \label{eq:bf-hd-actual-sv-transform-identity}
\end{equation}
Since \(\varepsilon_t^2\sim\chi_1^2\), the transformed observation density is
therefore
\begin{equation}
  g_\theta(z_t\mid h_t)
  =
  p_{\log\chi_1^2}\!\left(z_t-\log(\beta^2)-h_t\right).
  \label{eq:bf-hd-actual-sv-transform-density}
\end{equation}
Thus the transformed actual-SV one-step normalizer is
\begin{equation}
  Z_t(\theta)
  =
  \int g_\theta(z_t\mid h_t)
  p_\theta(h_t\mid z_{1:t-1})\,dh_t,
  \label{eq:bf-hd-actual-sv-normalizer}
\end{equation}
and the target remains the single cumulative likelihood
\(
  \ell_T(\theta)=\sum_{t=1}^T\log Z_t(\theta)
\).

This identity fixes the admissible same-target routes.  A dense exact-transformed
reference, a direct-likelihood fixed-cloud rule, or the current factorized
scalar Zhao--Cui wrapper may differ numerically, but they remain comparable only
when they all evaluate \eqref{eq:bf-hd-actual-sv-transform-density} and then
accumulate the resulting transformed normalizers.  The direct-likelihood route
has the generic node-reweighting form
\begin{equation}
  \widehat Z_t^{\rm dir}(\theta;B)
  =
  \sum_{r=1}^M
  w_r
g_\theta\bigl(z_t\mid h_t^{(r)}\bigr),
  \label{eq:bf-hd-actual-sv-direct-normalizer}
\end{equation}
with node locations \(h_t^{(r)}\) and weights \(w_r\) supplied by the declared
predictive approximation branch \(B\).  The dense reference and the present
Zhao--Cui comparator are same-target for the same reason: their value paths also
consume the transformed observations \(z_t\) and the exact transformed density
in \eqref{eq:bf-hd-actual-sv-transform-density}.

\begin{proposition}[Augmented Gaussian closure changes the actual-SV scalar]
\label{prop:bf-hd-actual-sv-gc-wrong-target}
Suppose instead that one augments the one-step state with an explicit
observation-noise coordinate and uses the raw observation map in
\eqref{eq:bf-hd-actual-sv-raw-observation}.  If the resulting filter propagates
that augmented cloud through the raw map, replaces
\eqref{eq:bf-hd-actual-sv-transform-density} by predictive observation moments,
and accumulates a Gaussian innovation likelihood such as
\eqref{eq:bf-hd-sgqf-loglik}, then the accumulated scalar is not the transformed
actual-SV likelihood target \(\ell_T(\theta)\).
\end{proposition}

\begin{proof}
The transformed target in \eqref{eq:bf-hd-actual-sv-normalizer} is defined by
integrating the exact transformed observation density
\(g_\theta(z_t\mid h_t)\).  By contrast, the augmented Gaussian-closure route
uses the raw map \(y_t=\beta e^{h_t/2}\varepsilon_t\), forms predictive
observation moments such as \(\bar z_t\) and \(S_t\), and then accumulates a
Gaussian innovation scalar.  That replacement changes the observation-density
object inside the one-step normalizer.  Once the one-step normalizer changes, the
cumulative likelihood changes as well by \eqref{eq:bf-hd-likelihood-factor}.
Hence the Gaussian-closure route may define a different declared approximate
scalar, but it is not a same-target approximation to the transformed actual-SV
likelihood.
\end{proof}

\begin{proposition}[When a Gaussian-closure likelihood surrogate can be reasonable]
\label{prop:bf-hd-gaussian-closure-surrogate-regime}
Let \(p_t^{\rm obs}(y)\) be the predictive observation density induced by the
correct augmented input law, and let \((\bar z_t,S_t)\) be the predictive mean
and covariance used in the Gaussian innovation scalar
\eqref{eq:bf-hd-sgqf-loglik}.  Suppose:
\begin{enumerate}[label=(\roman*), leftmargin=0.28in]
  \item the sigma-point or fixed-cloud rule is applied to the correct augmented
    input variable, so that the moment statement in
    Proposition~\ref{prop:bf-structural-ukf-ut-accuracy} applies;
  \item the local small-uncertainty and smoothness hypotheses behind
    Proposition~\ref{prop:bf-structural-ukf-ut-accuracy} hold, so the predictive
    mean and covariance are accurate through the stated second-order regime; and
  \item on a neighborhood \(\mathcal U_t\) containing the realized observation
    \(y_t\), the predictive observation log density satisfies
    \begin{equation}
      \log p_t^{\rm obs}(y)
      =
      \log \varphi_{n_y}(y;\bar z_t,S_t)+r_t(y),
      \qquad y\in\mathcal U_t,
      \label{eq:bf-hd-gaussian-closure-log-density-remainder}
    \end{equation}
    with remainder bounded by
    \begin{equation}
      |r_t(y)|\le \delta_t,
      \qquad y\in\mathcal U_t.
      \label{eq:bf-hd-gaussian-closure-remainder-bound}
    \end{equation}
\end{enumerate}
Then, provided \(y_t\in\mathcal U_t\), the one-step Gaussian innovation scalar
is a local likelihood surrogate with error bounded by
\begin{equation}
  \left|
    \log p_t^{\rm obs}(y_t)
    -
    \left[-\frac12\left\{\log\det S_t+v_t^\top S_t^{-1}v_t+n_y\log(2\pi)\right\}\right]
  \right|
  \le \delta_t.
  \label{eq:bf-hd-gaussian-closure-surrogate-bound}
\end{equation}
\end{proposition}

\begin{proof}
By definition of the Gaussian density,
\[
  \log \varphi_{n_y}(y_t;\bar z_t,S_t)
  =
  -\frac12\left\{\log\det S_t+v_t^\top S_t^{-1}v_t+n_y\log(2\pi)\right\},
\]
with \(v_t=y_t-\bar z_t\).  Evaluating
\eqref{eq:bf-hd-gaussian-closure-log-density-remainder} at \(y_t\) gives
\[
  \log p_t^{\rm obs}(y_t)
  -
  \log \varphi_{n_y}(y_t;\bar z_t,S_t)
  =
  r_t(y_t),
\]
and \eqref{eq:bf-hd-gaussian-closure-surrogate-bound} follows from
\eqref{eq:bf-hd-gaussian-closure-remainder-bound}.
\end{proof}

This proposition states the admissible sense in which a UKF-style or
fixed-cloud Gaussian closure can still be useful.  The structural UKF result
controls predictive moments under the correct augmented input law; the extra
remainder assumption controls how far the full predictive observation density is
from the Gaussian density built from those moments.  Both ingredients are needed.
Accurate moments alone do not guarantee an accurate likelihood scalar when the
predictive observation law is strongly skewed, heavy-tailed, multimodal, or
otherwise far from Gaussian near the realized observation.

For actual SV this means the augmented Gaussian-closure route may be a reasonable
local surrogate likelihood under a small-uncertainty, near-Gaussian predictive
observation regime, but it still does not become the same target as the
transformed exact likelihood in \eqref{eq:bf-hd-actual-sv-normalizer}.  It is a
surrogate scalar, not a second exact route to the same inference problem.

\subsection{Actual-SV SR-UKF Augmented-Noise Adapter}
\label{sec:bf-hd-actual-sv-srukf-augmented-adapter}

This subsection gives the model-specific adapter for the factor-propagating
SR-UKF score contract in
Section~\ref{sec:bf-srukf-factor-propagating-score-contract}.  It is deliberately
an adapter for the raw-observation Gaussian-closure surrogate in
Proposition~\ref{prop:bf-hd-gaussian-closure-surrogate-regime}; it is not an
admission that the surrogate equals the transformed actual-SV target in
\eqref{eq:bf-hd-actual-sv-normalizer}.

For one panel coordinate, use the scalar latent recursion
\begin{equation}
  H_t=\gamma H_{t-1}+U_t,
  \qquad
  U_t\sim\calN(0,\sigma^2),
  \qquad
  Y_t=\beta \exp(H_t/2)E_t,
  \qquad
  E_t\sim\calN(0,1).
  \label{eq:bf-hd-actual-sv-srukf-model}
\end{equation}
The candidate score contract for this surrogate uses the unconstrained
parameterization
\begin{equation}
  \theta=(\theta_\gamma,\theta_\beta),
  \qquad
  \gamma=\Phi(\theta_\gamma),
  \qquad
  \beta=\exp(\theta_\beta),
  \qquad
  \sigma \text{ fixed}.
  \label{eq:bf-hd-actual-sv-srukf-parameterization}
\end{equation}
Here \(\Phi\) is the standard normal distribution function and
\(\varphi\) is its density, so
\begin{equation}
  \partial_{\theta_\gamma}\gamma=\varphi(\theta_\gamma),
  \qquad
  \partial_{\theta_\beta}\beta=\beta,
  \qquad
  \partial_{\theta_i}\sigma=0.
  \label{eq:bf-hd-actual-sv-srukf-parameter-derivatives}
\end{equation}
A sigma-point adapter must place points on the pre-observation uncertainty
variable
\begin{equation}
  A_t=(H_{t-1},U_t,E_t)^\top .
  \label{eq:bf-hd-actual-sv-srukf-augmented-variable}
\end{equation}
If the previous filtered approximation is
\[
  H_{t-1}\mid y_{1:t-1}\approx
  \calN(m_{t-1},c_{t-1}c_{t-1}^\top),
\]
then the declared augmented law is
\begin{equation}
  A_t\mid y_{1:t-1}
  \approx
  \calN\!\left(
    \begin{bmatrix}m_{t-1}\\0\\0\end{bmatrix},
    C_{a,t}C_{a,t}^\top
  \right),
  \qquad
  C_{a,t}
  =
  \begin{bmatrix}
    c_{t-1} & 0 & 0\\
    0 & \sigma & 0\\
    0 & 0 & 1
  \end{bmatrix}.
  \label{eq:bf-hd-actual-sv-srukf-augmented-law}
\end{equation}
The exact shape of \(c_{t-1}\) follows the scalar factor branch used by the
generic SR-UKF handoff.  The crucial point is the three-coordinate law:
lagged latent state, state innovation, and observation shock are all present.

The state and raw-observation maps sent to the generic SR-UKF backend are
\begin{align}
  X_\theta(A_t)
    &=
    \gamma H_{t-1}+U_t,
    \label{eq:bf-hd-actual-sv-srukf-transition-map}\\
  Z_\theta(A_t)
    &=
    \beta\exp(X_\theta(A_t)/2)E_t .
    \label{eq:bf-hd-actual-sv-srukf-observation-map}
\end{align}
The symbol \(Z_\theta(A_t)\) in
\eqref{eq:bf-hd-actual-sv-srukf-observation-map} is the raw predicted
observation coordinate used by the Gaussian closure; it is not the transformed
datum \(z_t=\log(y_t^2)\) from
\eqref{eq:bf-hd-actual-sv-z-definition}.

For a placed point \(a^{(j)}=(h_-^{(j)},u^{(j)},e^{(j)})\), the local map
derivatives required by
\eqref{eq:bf-srukf-propagated-point-first} are
\begin{align}
  D_a X_\theta(a^{(j)})
    &=
    \begin{bmatrix}\gamma&1&0\end{bmatrix},
    &
  \partial_{\theta_\gamma}X_\theta(a^{(j)})
    &=
    \varphi(\theta_\gamma)h_-^{(j)},
    &
  \partial_{\theta_\beta}X_\theta(a^{(j)})
    &=0,
    \label{eq:bf-hd-actual-sv-srukf-transition-derivatives}\\
  D_a Z_\theta(a^{(j)})
    &=
    \beta\exp(x^{(j)}/2)
    \begin{bmatrix}
      \frac12\gamma e^{(j)}&
      \frac12 e^{(j)}&
      1
    \end{bmatrix},
    \label{eq:bf-hd-actual-sv-srukf-observation-state-derivative}\\
  \partial_{\theta_\gamma}Z_\theta(a^{(j)})
    &=
    \frac12
    \beta\exp(x^{(j)}/2)e^{(j)}
    \varphi(\theta_\gamma)h_-^{(j)},
    &
  \partial_{\theta_\beta}Z_\theta(a^{(j)})
    &=
    \beta\exp(x^{(j)}/2)e^{(j)},
    \label{eq:bf-hd-actual-sv-srukf-observation-parameter-derivatives}
\end{align}
where \(x^{(j)}=X_\theta(a^{(j)})\).  These are ordinary pointwise derivatives
of the declared maps.  They do not differentiate a principal matrix square
root, an eigensystem, an SVD basis, or an autodiff tape.

The initial stationary branch, when used, is
\begin{equation}
  H_0\sim\calN\!\left(0,\frac{\sigma^2}{1-\gamma^2}\right),
  \qquad |\gamma|<1,
  \label{eq:bf-hd-actual-sv-srukf-initial-law}
\end{equation}
with score handoff derivatives
\begin{equation}
  \partial_{\theta_\gamma}m_0=0,\qquad
  \partial_{\theta_\beta}m_0=0,\qquad
  \partial_{\theta_\gamma}
  \frac{\sigma^2}{1-\gamma^2}
  =
  \frac{2\sigma^2\gamma}{(1-\gamma^2)^2}\varphi(\theta_\gamma),
  \qquad
  \partial_{\theta_\beta}
  \frac{\sigma^2}{1-\gamma^2}=0.
  \label{eq:bf-hd-actual-sv-srukf-initial-derivatives}
\end{equation}
If the factor-propagating backend consumes the scalar stationary factor
\(c_0=\sigma(1-\gamma^2)^{-1/2}\) rather than the variance, the corresponding
fixed-\(\sigma\) branch derivative is
\begin{equation}
  \partial_{\theta_\gamma}c_0
  =
  \sigma\gamma(1-\gamma^2)^{-3/2}\varphi(\theta_\gamma),
  \qquad
  \partial_{\theta_\beta}c_0=0.
  \label{eq:bf-hd-actual-sv-srukf-initial-factor-derivatives}
\end{equation}
If a nonstationary initialization is chosen, the initialization law and its
derivatives must replace \eqref{eq:bf-hd-actual-sv-srukf-initial-law}--
\eqref{eq:bf-hd-actual-sv-srukf-initial-factor-derivatives} explicitly.

Given the placed points from
\eqref{eq:bf-hd-actual-sv-srukf-augmented-law}, the generic SR-UKF backend
computes
\[
  \bar x_t,\quad \bar y_t,\quad
  P_{xx,t,\star},\quad S_{t,\star},\quad P_{xy,t,\star}
\]
from the propagated pairs
\((X_\theta(a_t^{(j)}),Z_\theta(a_t^{(j)}))\).  The declared one-step
surrogate contribution is
\begin{equation}
  \widehat\ell_t^{\rm rawGC}(\theta)
  =
  -\frac12
  \left[
    \log(2\pi)+\log S_{t,\star}
    +(y_t-\bar y_t)^2S_{t,\star}^{-1}
  \right],
  \label{eq:bf-hd-actual-sv-srukf-surrogate-loglik}
\end{equation}
or its vector-panel analogue when several independent coordinates are combined.
Its analytical score is exactly the specialization of
\eqref{eq:bf-srukf-score-first} with \(v_t=y_t-\bar y_t\), followed by the
filtered-state derivative handoff
\eqref{eq:bf-srukf-filtered-mean-first}--
\eqref{eq:bf-srukf-filtered-factor-first}:
\begin{equation}
  \partial_{\theta_i}
  \sum_{t=1}^T\widehat\ell_t^{\rm rawGC}(\theta)
  =
  \sum_{t=1}^T
  \partial_{\theta_i}
  \widehat\ell_t^{\rm rawGC}(\theta),
  \qquad i\in\{\gamma,\beta\}.
  \label{eq:bf-hd-actual-sv-srukf-score-handoff}
\end{equation}
This is the score of the raw actual-SV Gaussian-closure surrogate scalar.  It
is not the score of \(\ell_T(\theta)\) in
\eqref{eq:bf-hd-actual-sv-normalizer}.

A two-coordinate collapsed-process specialization may be used only after this
three-coordinate adapter has been audited.  The required proof obligation is
the scalar Gaussian law equivalence
\begin{equation}
  (H_t^-,E_t)
  =
  (\gamma H_{t-1}+U_t,E_t)
  \sim
  \calN\!\left(
    \begin{bmatrix}\gamma m_{t-1}\\0\end{bmatrix},
    \begin{bmatrix}
      \gamma^2 c_{t-1}c_{t-1}^\top+\sigma^2 & 0\\
      0 & 1
    \end{bmatrix}
  \right),
  \label{eq:bf-hd-actual-sv-srukf-collapsed-law-equivalence}
\end{equation}
together with derivative equivalence for the mean and factor handoff.  A
collapsed route that cannot demonstrate
\eqref{eq:bf-hd-actual-sv-srukf-collapsed-law-equivalence} and the matching
derivatives remains a separate approximation and cannot be used as the admitted
SR-UKF analytical-score route.

\begin{equation}
  \boxed{
  \begin{gathered}
  \text{No exact actual-SV likelihood claim;}\\
  \text{no same-target transformed-likelihood claim;}\\
  \text{no GradientTape score admission;}\\
  \text{no historical SVD/eigenderivative score admission;}\\
  \text{no strict-SPD principal-root derivative admission.}
  \end{gathered}}
  \label{eq:bf-hd-actual-sv-srukf-nonclaims}
\end{equation}

The practical consequence is strict.  Same-scalar derivative checks, fixed-branch
audits, or optimizer/HMC experiments are meaningful only after the value path is
a route to \eqref{eq:bf-hd-actual-sv-normalizer}.  Internal consistency of a
Gaussian-closure scalar does not repair the target mismatch.  This is why the
current transformed direct-likelihood route and the current Zhao--Cui comparator
remain admissible same-target comparators, whereas the previous augmented
Gaussian-closure framing does not.

\paragraph{Reader checkpoint.}
What should now be clear: filtering updates a probability law and a scalar
normalizer at the same time.  Object preserved: the conditional law after
normalization.  Approximation enters at: the representation of the predictive
law, the integral for \(Z_t\), or both.  Exported to synthesis: any industrial
method must say whether it preserves the intended likelihood scalar, merely a
different surrogate scalar, or only a moment summary.

\begin{proposition}[Predictive score identity]
\label{prop:bf-hd-score}
Let $Z_t(\theta)=\int a_t(x,\theta)\,dx$ with
$a_t(x,\theta)=g_\theta(y_t\mid x)p_\theta(x\mid y_{1:t-1})$.  If
$a_t>0$ on the posterior support and dominated differentiation holds, then
\begin{equation}
  \nabla_\theta\log Z_t(\theta)
  =
  \int \nabla_\theta\log a_t(x,\theta)\,
       p_\theta(x\mid y_{1:t})\,dx .
  \label{eq:bf-hd-score}
\end{equation}
\end{proposition}

\begin{proof}
Under dominated differentiation,
\[
  \nabla_\theta\log Z_t(\theta)
  =\frac{\nabla_\theta Z_t(\theta)}{Z_t(\theta)}
  =
  \frac{\int \nabla_\theta a_t(x,\theta)\,dx}
       {\int a_t(x,\theta)\,dx}.
\]
On the support where \(a_t>0\),
\(\nabla_\theta a_t=a_t\nabla_\theta\log a_t\).  Substitution gives
\[
  \nabla_\theta\log Z_t(\theta)
  =
  \int \nabla_\theta\log a_t(x,\theta)
       \frac{a_t(x,\theta)}{Z_t(\theta)}\,dx .
\]
The last factor is the filtering density in
\eqref{eq:bf-hd-update}.  Hence \eqref{eq:bf-hd-score} follows.  The identity
does not say that a numerical score is correct; it says that any reported score
must correspond to the derivative of the same normalizer used in the reported
likelihood.
\end{proof}

This proposition is exact but not automatically implementable.  A numerical
filter supplies an approximate normalizer $\widehat Z_t$ and an approximate
history-dependent distribution.  If value and gradient are not derivatives of
the same scalar approximation, a downstream optimizer or HMC sampler is solving
a different problem from the one reported by the filter.

\section{Analytical Likelihood-Gradient And Filtering Sensitivities}
\label{sec:bf-hd-analytical-likelihood-gradient}

This section is a project derivation for the dominated discrete-time model in
\eqref{eq:bf-hd-ssm}.  It is not a continuous-time adjoint theorem.  The PDE
and robust-DMZ settings require their own sensitivity or adjoint analysis under
the regularity assumptions of the cited sources.

Write
\[
  p_t^-(x_t)=p_\theta(x_t\mid y_{1:t-1}),\qquad
  p_t(x_t)=p_\theta(x_t\mid y_{1:t}),
\]
and let the parameter-gradient sensitivities be
\[
  s_t^-(x_t)=\nabla_\theta p_t^-(x_t),\qquad
  s_t(x_t)=\nabla_\theta p_t(x_t).
\]
These sensitivities are vector-valued when \(\theta\) is vector-valued; the
derivation is componentwise.  Differentiating the prediction equation
\eqref{eq:bf-hd-prediction} under Assumption~\ref{ass:bf-hd-dominated} gives
\begin{equation}
  s^-_t(x_t)
  =
  \int
  \left[
  \nabla_\theta f_\theta(x_t\mid x_{t-1})
  p_{t-1}(x_{t-1})
  +
  f_\theta(x_t\mid x_{t-1})
  s_{t-1}(x_{t-1})
  \right]dx_{t-1}.
  \label{eq:bf-hd-prediction-sensitivity}
\end{equation}
The first term is the direct parameter effect on the transition density.  The
second term is the inherited sensitivity of the previous filtering law.  If
the initial distribution depends on \(\theta\), the recursion starts from
\(s_0(x_0)=\nabla_\theta p_\theta(x_0)\), or from the corresponding
post-initial-observation sensitivity if an initial update is used.

The one-step normalizer is
\begin{equation}
  Z_t(\theta)=\int g_\theta(y_t\mid x_t)p_t^-(x_t)\,dx_t .
  \label{eq:bf-hd-gradient-normalizer}
\end{equation}
Differentiating it gives
\begin{equation}
  \nabla_\theta Z_t(\theta)
  =
  \int
  \left[
  \nabla_\theta g_\theta(y_t\mid x_t)p^-_t(x_t)
  +
  g_\theta(y_t\mid x_t)s^-_t(x_t)
  \right]dx_t .
  \label{eq:bf-hd-normalizer-gradient}
\end{equation}
Therefore
\begin{equation}
  \nabla_\theta \ell_T(\theta)
  =
  \sum_{t=1}^T
  \nabla_\theta \log Z_t(\theta)
  =
  \sum_{t=1}^T
  \frac{\nabla_\theta Z_t(\theta)}{Z_t(\theta)} .
  \label{eq:bf-hd-likelihood-gradient}
\end{equation}

It remains to differentiate the normalized update.  Define
\[
  n_t(x_t,\theta)=g_\theta(y_t\mid x_t)p_t^-(x_t),
  \qquad
  p_t(x_t)=\frac{n_t(x_t,\theta)}{Z_t(\theta)}.
\]
The numerator sensitivity is
\[
  \nabla_\theta n_t(x_t,\theta)
  =
  \nabla_\theta g_\theta(y_t\mid x_t)p_t^-(x_t)
  +
  g_\theta(y_t\mid x_t)s_t^-(x_t).
\]
Using the quotient rule and
\(\nabla_\theta Z_t/Z_t=\nabla_\theta\log Z_t\),
\begin{equation}
  s_t(x_t)
  =
  \frac{
  \nabla_\theta g_\theta(y_t\mid x_t)p^-_t(x_t)
  +
  g_\theta(y_t\mid x_t)s^-_t(x_t)
  }{Z_t(\theta)}
  -
  p_t(x_t)\nabla_\theta\log Z_t(\theta).
  \label{eq:bf-hd-update-sensitivity}
\end{equation}
Integrating \eqref{eq:bf-hd-update-sensitivity} over \(x_t\) gives zero,
because a normalized density has derivative of total mass equal to zero.  This
is the sensitivity analogue of probability conservation.

\begin{proposition}[Discrete-time likelihood sensitivity recursion]
\label{prop:bf-hd-likelihood-sensitivity}
Under Assumption~\ref{ass:bf-hd-dominated}, with differentiable transition and
observation densities and finite positive \(Z_t(\theta)\), equations
\eqref{eq:bf-hd-prediction-sensitivity}--\eqref{eq:bf-hd-update-sensitivity}
give an analytical recursion for \(\nabla_\theta\ell_T(\theta)\) in the exact
discrete-time filter.
\end{proposition}

\begin{proof}
Equation~\eqref{eq:bf-hd-prediction-sensitivity} is differentiation of the
prediction integral.  Equation~\eqref{eq:bf-hd-normalizer-gradient} is
differentiation of the predictive normalizer.  Equation
\eqref{eq:bf-hd-likelihood-gradient} follows from the likelihood factorization
\eqref{eq:bf-hd-likelihood-factor} and the identity
\(\nabla\log Z=\nabla Z/Z\) for \(Z>0\).  Equation
\eqref{eq:bf-hd-update-sensitivity} is the quotient rule applied to
\(p_t=n_t/Z_t\).  Each step is componentwise in \(\theta\) and uses dominated
differentiation to exchange gradient and integral.
\end{proof}

This recursion is the analytical object needed by Chapter~\ref{ch:bf-highdim-hmc-research}.
An automatic-differentiation tape through an approximate filter is a different
object unless it differentiates exactly the same scalar displayed to the
sampler.

\section{Compact Sensitivity Score For Finite LEDH--PFPF--OT}
\label{sec:bf-hd-ledh-compact-sensitivity-score}

This section specializes the preceding sensitivity principle to the finite
LEDH--PFPF--OT scalar used by the benchmark code.  The object is not the exact
Kalman likelihood and not the derivative of an ideal infinite-iteration
transport solve.  It is the derivative of the finite computation actually run:
fixed base random numbers, a fixed resampling mask, finitely many LEDH flow
steps, finitely many Sinkhorn balancing steps, and the displayed accumulated
log normalizers.  A partial derivative that treats the transport
normalization, scale, Sinkhorn potentials, or transported particles as fixed
when they change with \(\theta\) is not the score of this finite scalar.

Let \(\theta\in\R^p\), and let \(r_{0,i}\) and \(r_{t,i}\) be the fixed
standard-normal base draws used to form the initial particles and transition
noise.  The compact method carries, at each time \(t\),
\[
  X_t=\{x_{t,i}\}_{i=1}^N,\qquad
  \lambda_t=\{\lambda_{t,i}\}_{i=1}^N,\qquad
  \dot X_t=\partial_\theta X_t,\qquad
  \dot\lambda_t=\partial_\theta \lambda_t,
\]
where \(\lambda_t\) are normalized log weights.  It does not store the full
history of LEDH matrices or transport plans.  For a diagonal LGSSM row with
\[
  A_\theta=\operatorname{diag}(\phi_1,\ldots,\phi_d),\qquad
  Q_\theta=q^2 I,\qquad
  R_\theta=r^2 I,
\]
the fixed-randomness prediction is
\begin{align}
  m_{t,i} &= A_\theta x_{t-1,i},&
  x^-_{t,i} &= m_{t,i}+q\,r_{t,i},&
  \dot m_{t,i} &= A_\theta\dot x_{t-1,i}
    +(\partial_\theta A_\theta)x_{t-1,i},&
  \dot x^-_{t,i} &= \dot m_{t,i}+(\partial_\theta q)r_{t,i}.
  \label{eq:bf-hd-ledh-compact-prediction-sensitivity}
\end{align}
These are total derivatives under the fixed base draws.  They are not
derivatives after redrawing the particles.

For a linear observation \(y_t=H x_t+\epsilon_t\), the LEDH affine map used in
the row can be written
\begin{align}
  P_\theta &= Q_\theta^{-1},&
  S_\theta &= R_\theta^{-1},&
  \Lambda_\theta &= P_\theta+H^\top S_\theta H,&
  C_\theta &= \Lambda_\theta^{-1},
  \label{eq:bf-hd-ledh-compact-ledh-matrices}\\
  \mu_{t,i} &= C_\theta\{P_\theta m_{t,i}+H^\top S_\theta y_t\},&
  B_\theta &= \operatorname{chol}(C_\theta)\operatorname{chol}(Q_\theta)^{-1},&
  z_{t,i} &= \mu_{t,i}+B_\theta(x^-_{t,i}-m_{t,i}).
  \label{eq:bf-hd-ledh-compact-ledh-map}
\end{align}
Here \(Q_\theta\) and \(R_\theta\) are positive definite, so
\(P_\theta\), \(S_\theta\), \(\Lambda_\theta\), \(C_\theta\), the Cholesky
factors, and the log determinant below are well defined.  If a numerical
implementation adds jitter or projects a covariance to a positive-definite
matrix, the derivative is the derivative of that jittered or projected finite
operation, not of the unjittered formula.
The tangent of this map is obtained by differentiating these matrix equations:
\begin{align}
  \dot P_\theta &=-P_\theta\dot Q_\theta P_\theta,&
  \dot S_\theta &=-S_\theta\dot R_\theta S_\theta,&
  \dot\Lambda_\theta&=\dot P_\theta+H^\top\dot S_\theta H,&
  \dot C_\theta&=-C_\theta\dot\Lambda_\theta C_\theta,
  \label{eq:bf-hd-ledh-compact-matrix-sensitivity}\\
  \dot\mu_{t,i}
    &=\dot C_\theta\{P_\theta m_{t,i}+H^\top S_\theta y_t\}
      +C_\theta\{\dot P_\theta m_{t,i}
      +P_\theta\dot m_{t,i}+H^\top\dot S_\theta y_t\},\\
  \dot z_{t,i}
    &=\dot\mu_{t,i}
      +\dot B_\theta(x^-_{t,i}-m_{t,i})
      +B_\theta(\dot x^-_{t,i}-\dot m_{t,i}).
  \label{eq:bf-hd-ledh-compact-ledh-sensitivity}
\end{align}
For \(L=\operatorname{chol}(C_\theta)\), the Cholesky tangent is defined by
\[
  E=L^{-1}\dot C_\theta L^{-\top},\qquad
  M_{ij}=
  \begin{cases}
    E_{ij}, & i>j,\\
    E_{ii}/2, & i=j,\\
    0, & i<j,
  \end{cases}
  \qquad
  \dot L=LM.
  \label{eq:bf-hd-ledh-compact-cholesky-jvp}
\]
Thus \(\dot B_\theta=\dot L\operatorname{chol}(Q_\theta)^{-1}
 +L\,\partial_\theta\{\operatorname{chol}(Q_\theta)^{-1}\}\).

The log-weight correction is differentiated in the same forward pass.  With
\[
  c_{t,i}
  =\lambda_{t-1,i}
   +\log f_\theta(z_{t,i}\mid x_{t-1,i})
   +\log g_\theta(y_t\mid z_{t,i})
   -\log f_\theta(x^-_{t,i}\mid x_{t-1,i})
   +\log|\det B_\theta|,
  \label{eq:bf-hd-ledh-compact-corrected-weight}
\]
define \(a_t=\operatorname{logsumexp}_i c_{t,i}\) and
\(\lambda^+_{t,i}=c_{t,i}-a_t\).  The one-step likelihood contribution and
normalized-log-weight tangent are
\begin{align}
  \dot a_t &= \sum_{i=1}^N \omega^+_{t,i}\dot c_{t,i},&
  \dot\lambda^+_{t,i} &= \dot c_{t,i}-\dot a_t,&
  \omega^+_{t,i}&=\exp(\lambda^+_{t,i}).
  \label{eq:bf-hd-ledh-compact-normalize-sensitivity}
\end{align}
If an implementation floors normalized weights by replacing
\(\omega^+_{t,i}\) with \(\max(\omega^+_{t,i},\epsilon_{\mathrm{floor}})\),
then the derivative must include that branch: active floored entries have zero
tangent for the displayed floored log weight.  Ignoring this branch is again a
different derivative.

It remains to differentiate the finite Sinkhorn equal-weighting map.  Write the
streaming softmin used by the code as
\begin{equation}
  \operatorname{sm}_\varepsilon(q;K,v)
  =
  -\varepsilon\log\sum_j
    \exp\{v_j-\|q-k_j\|^2/(2\varepsilon)\}.
  \label{eq:bf-hd-ledh-compact-softmin}
\end{equation}
For tangents \((\dot q,\dot K,\dot v,\dot\varepsilon)\), define
\[
  \pi_j(q)
  =
  \frac{\exp\{v_j-\|q-k_j\|^2/(2\varepsilon)\}}
       {\sum_l\exp\{v_l-\|q-k_l\|^2/(2\varepsilon)\}} .
\]
The softmin tangent is
\begin{align}
  \partial\operatorname{sm}_\varepsilon
  &=
  -\dot\varepsilon\log\sum_j
    \exp\{v_j-\|q-k_j\|^2/(2\varepsilon)\}
  -\varepsilon\sum_j\pi_j(q)\dot h_j,
  \label{eq:bf-hd-ledh-compact-softmin-jvp}\\
  \dot h_j
  &=
  \dot v_j
  -\frac{(q-k_j)^\top(\dot q-\dot k_j)}{\varepsilon}
  +\frac{\|q-k_j\|^2}{2\varepsilon^2}\dot\varepsilon .
  \nonumber
\end{align}
This equation is local and can be evaluated in blocks.  It does not require a
dense \(N\times N\) matrix to be stored.

The finite Sinkhorn potentials are a fixed composition of the softmin
\eqref{eq:bf-hd-ledh-compact-softmin}.  With source log weights
\(\alpha=\lambda^+_t\), target log weights \(\beta=-\log N\), scaled particles
\(u_i=(z_{t,i}-\bar z_t)/s_t\), and deterministic annealing values
\(\varepsilon_k\), the executed code carries four potentials.  Starting from
the cold-start softmins, one finite balancing step has the form
\begin{align}
  a_y^{k+1} &=\frac12\left[
    a_y^k+\operatorname{sm}_{\varepsilon_k}
      (u;u,\alpha+b_x^k/\varepsilon_k)\right],\\
  b_x^{k+1} &=\frac12\left[
    b_x^k+\operatorname{sm}_{\varepsilon_k}
      (u;u,\beta+a_y^k/\varepsilon_k)\right],\\
  a_x^{k+1} &=\frac12\left[
    a_x^k+\operatorname{sm}_{\varepsilon_k}
      (u;u,\alpha+a_x^k/\varepsilon_k)\right],\\
  b_y^{k+1} &=\frac12\left[
    b_y^k+\operatorname{sm}_{\varepsilon_k}
      (u;u,\beta+b_y^k/\varepsilon_k)\right].
  \label{eq:bf-hd-ledh-compact-sinkhorn-recursion}
\end{align}
After the declared finite number of balancing steps, the transport application
uses \(a=\operatorname{sm}_{\varepsilon}(u;u,\alpha+b_x/\varepsilon)\) and
\(b=\operatorname{sm}_{\varepsilon}(u;u,\beta+a_y/\varepsilon)\).
The tangent recursion is exactly the derivative of
\eqref{eq:bf-hd-ledh-compact-sinkhorn-recursion}, using
\eqref{eq:bf-hd-ledh-compact-softmin-jvp} at each finite iteration.  The scale
and initial annealing value are part of the same derivative:
\begin{align}
  \bar z_t &= N^{-1}\sum_i z_{t,i},&
  u_{t,i}&=(z_{t,i}-\bar z_t)/s_t,&
  \dot u_{t,i}
    &=\frac{\dot z_{t,i}-\dot{\bar z}_t}{s_t}
      -\frac{z_{t,i}-\bar z_t}{s_t^2}\dot s_t,\\
  \varepsilon_0&=\max\{(\max_{i,j}u_{t,ij}-\min_{i,j}u_{t,ij})^2,10^{-6}\}.
  \label{eq:bf-hd-ledh-compact-scale-epsilon-sensitivity}
\end{align}
When the max or min is tied, the implementation must choose and document a
subgradient convention.  Away from ties, this is the ordinary derivative.

Given final finite potentials \(a,b\), the barycentric transport row is
\begin{equation}
  \tilde z_{t,i}
  =
  \sum_j
  \exp\left\{
    \frac{a_i+b_j-\|u_i-u_j\|^2/2}{\varepsilon}
    -\rho_j+\log N+\lambda^+_{t,j}
  \right\}z_{t,j},
  \label{eq:bf-hd-ledh-compact-transport-application}
\end{equation}
where \(\rho_j\) is the column log normalizer for the same finite potentials.
Its tangent is the direct product rule:
\begin{equation}
  \dot{\tilde z}_{t,i}
  =
  \sum_j T_{ij}\left[
    \dot z_{t,j}
    + z_{t,j}
      \left(
        \dot\eta_{ij}-\dot\rho_j+\dot\lambda^+_{t,j}
      \right)
  \right],
  \label{eq:bf-hd-ledh-compact-transport-jvp}
\end{equation}
where \(T_{ij}\) is the finite transport weight in
\eqref{eq:bf-hd-ledh-compact-transport-application} and
\[
  \eta_{ij}
  =
  \frac{a_i+b_j-\|u_i-u_j\|^2/2}{\varepsilon}.
\]
The column-normalizer tangent is the softmax average of \(\dot\eta_{ij}\) over
rows \(i\).  Equation~\eqref{eq:bf-hd-ledh-compact-transport-jvp} is the
forward-mode counterpart of a manual VJP.  It computes the same total
derivative of the same finite Sinkhorn map, but it carries tangents forward in
time and therefore does not need the full filtering history.

\begin{proposition}[Compact finite LEDH--PFPF--OT score recursion]
\label{prop:bf-hd-ledh-compact-sensitivity-score}
Assume \(Q_\theta\) and \(R_\theta\) are positive definite with the stated
dimensions, every jittered or projected matrix used by the finite code remains
positive definite, and the parameter point is away from nondifferentiable max,
floor, active-resampling, covariance-projection, and Cholesky-boundary ties.
Assume also that the finite LEDH and Sinkhorn operations above are executed with
fixed base random numbers and a fixed resampling mask.  If the filter loop carries
\((X_t,\lambda_t,\dot X_t,\dot\lambda_t)\) and updates the tangents by
\eqref{eq:bf-hd-ledh-compact-prediction-sensitivity}--
\eqref{eq:bf-hd-ledh-compact-transport-jvp}, then
\[
  \dot{\widehat\ell}_T(\theta)
  =
  \sum_{t=1}^T \dot a_t
  \label{eq:bf-hd-ledh-compact-score}
\]
is the total derivative of the finite scalar
\(\widehat\ell_T(\theta)=\sum_{t=1}^T a_t(\theta)\) computed by that same
fixed-randomness LEDH--PFPF--OT run.
\end{proposition}

\begin{proof}
Each displayed update is a differentiable finite-dimensional operation away
from the stated branch boundaries: matrix inverse, Cholesky, affine maps,
Gaussian log densities, logsumexp normalization, softmin, a fixed finite
composition of Sinkhorn updates, column log normalization, and barycentric
transport.  Equations
\eqref{eq:bf-hd-ledh-compact-prediction-sensitivity}--
\eqref{eq:bf-hd-ledh-compact-transport-jvp} are the chain rule applied to this
finite composition in the same order as the value computation.  The scalar
reported by the filter is the sum of the log normalizers \(a_t\).  Therefore
the carried scalar tangent is the sum of their tangents, which is
\eqref{eq:bf-hd-ledh-compact-score}.  No reverse-time cotangent or stored
history is required for this forward-mode proof.
\end{proof}

The memory difference is simple.  The full-history reverse route stores
time-indexed LEDH auxiliaries, transported states, and weights before scanning
backward.  The compact sensitivity route stores only the current state and its
\(p\) parameter tangents.  Its leading state memory is
\[
  O(BNDp+BNp+D^2p)
\]
plus local streaming blocks for Sinkhorn, rather than a time-multiplied
history.  This is the right route for a small parameter dimension such as the
LGSSM row.  For very large \(p\), a checkpointed adjoint may be cheaper, but it
must still differentiate the same finite scalar.

\paragraph{Reader checkpoint.}
What should now be clear: the likelihood gradient is not a separate numerical
decoration.  It is the derivative of the same normalizers that define
\(\ell_T\).  Object preserved: scalar value-gradient consistency.  Approximation
enters at: any approximate \(p_t^-\), \(\widehat Z_t\), truncation, rounding, or
adaptive gate.  Exported to synthesis: HMC can start only after the scalar and
gradient contract is named.

\section{Zakai, Kushner--Stratonovich, And DMZ Equations}
\label{sec:bf-hd-zakai-dmz}

For diffusion models, the normalized conditional distribution satisfies a
nonlinear filtering equation, while the unnormalized conditional measure
satisfies a linear Zakai equation.  We use \citet{vanhandel2007filtering} as the
standard checked derivation anchor for the normalized and unnormalized
equations rather than relying on unchecked historical priority papers.  In an
informal density notation, the unnormalized density $\sigma_t$ obeys a linear
stochastic equation of the form
\begin{equation}
  d\sigma_t
  =
  \mathcal L_\theta^\ast \sigma_t\,dt
  + h_\theta^\top \sigma_t\,dY_t,
  \label{eq:bf-hd-zakai-density}
\end{equation}
where $\mathcal L_\theta^\ast$ is the Fokker--Planck adjoint associated with the
signal generator.  The normalized density is
\begin{equation}
  \rho_t(x)=\frac{\sigma_t(x)}{\int \sigma_t(u)\,du}.
  \label{eq:bf-hd-normalize-zakai}
\end{equation}
The Duncan--Mortensen--Zakai viewpoint emphasizes the density/PDE form of
\eqref{eq:bf-hd-zakai-density}.  It is attractive in high dimension because it
turns filtering into density propagation; it is frightening for exactly the same
reason, since a tensor-product grid has exponentially many degrees of freedom.

\begin{remark}[Bibliographic scope]
For the equations used below, this chapter relies on checked technical sources:
van Handel for the standard normalized and unnormalized filtering equations, and
Davis/Yau papers for the pathwise robust DMZ transformation and memoryless
algorithm.  Classical priority names are retained for orientation, but a
historical citation is not used as a substitute for an inspected theorem or
derivation.
\end{remark}

\section{Pathwise Robust DMZ And Memoryless Computation}
\label{sec:bf-hd-pathwise-dmz}

\citet{davis1980multiplicative} studies a multiplicative functional
transformation for nonlinear filtering.  In the simple scalar-observation
diffusion setting, the idea is to absorb observation-path dependence into a
path-dependent transformation of the unnormalized density.  The transformed
object evolves according to a generator/PDE that can be studied pathwise rather
than only as a stochastic differential equation.

\citet{yau2000realtime} and \citet{yau2008realtime} build a robust DMZ
algorithm around this principle.  Their central computational move is to
freeze observation increments on a time partition and repeatedly solve
observation-independent Kolmogorov-type equations, correcting by
observation-dependent multiplicative factors.  In schematic form, over
$[t_j,t_{j+1}]$ one computes
\begin{align}
  \partial_s u_j(s,x) &= \mathcal K_{y_{t_j}} u_j(s,x),
       \qquad s\in[t_j,t_{j+1}], \label{eq:bf-hd-frozen-kolmogorov}\\
  u_{j+1}(t_{j+1},x) &=
       \mathcal M(y_{t_{j+1}}-y_{t_j},x)\,u_j(t_{j+1},x),
       \label{eq:bf-hd-memoryless-update}
\end{align}
where $\mathcal K_{y_{t_j}}$ and $\mathcal M$ denote the paper-specific frozen
operator and multiplicative observation update.  The details are not optional:
the convergence and bounded-domain approximation theorems in
\citet{yau2000realtime,yau2008realtime} depend on their growth, regularity,
weak-solution, and boundary assumptions.

\begin{algorithm}
\caption{Robust DMZ offline/online template}
\label{alg:bf-hd-robust-dmz}
\begin{algorithmic}[1]
\Require Partition $0=t_0<\cdots<t_K=T$, initial density, checked PR-DMZ
  assumptions, spatial domain or truncation rule.
\State Build or choose the observation-independent Kolmogorov solver required by
  the cited theorem.
\For{$j=0,\ldots,K-1$}
  \State Freeze the observation path according to the cited rule.
  \State Solve the frozen Kolmogorov equation on $[t_j,t_{j+1}]$.
  \State Apply the multiplicative observation correction.
  \State Normalize only after preserving the unnormalized-density semantics.
  \State Record mass, positivity, boundary, and discretization residuals.
\EndFor
\end{algorithmic}
\end{algorithm}

\citet{meng2025regularity} is a recent arXiv contribution rather than a settled
classical reference.  It is therefore used here provisionally: it links pathwise
robust DMZ regularity estimates to sparse functional tensor-train
approximations and a finite-difference QTT algorithm.  The durable lesson for
the rest of the monograph is methodological rather than evidentiary: a tensor
filter must make regularity, rank/compressibility, discretization error,
rounding error, and online-assimilation stability visible before small storage
can be interpreted as a filtering result.

\section{Approximate Targets And The HMC Boundary}
\label{sec:bf-hd-approx-target}

For a prior $\pi(\theta)$, the exact parameter posterior is proportional to
\begin{equation}
  \pi(\theta\mid y_{1:T})
  \propto \exp\{\ell_T(\theta)\}\pi(\theta).
  \label{eq:bf-hd-exact-posterior}
\end{equation}
If a filter supplies $\widehat\ell_T(\theta)$, the posterior implied by the
calculation is
\begin{equation}
  \widehat\pi(\theta\mid y_{1:T})
  \propto \exp\{\widehat\ell_T(\theta)\}\pi(\theta),
  \label{eq:bf-hd-approx-posterior}
\end{equation}
unless a pseudo-marginal or other exact correction is part of the algorithm.
The corresponding gradient is
\[
  \nabla_\theta \widehat\ell_T(\theta)
  =
  \nabla_\theta\sum_{t=1}^T\log\widehat Z_t(\theta),
\]
provided the reported gradient is the derivative of the reported scalar.
Truncation, resampling, clipping, tensor rounding, adaptive stopping, and
discrete accept/reject gates must therefore state which scalar is being
differentiated.  The HMC chapter returns to this point: exact Hamiltonian
dynamics for \eqref{eq:bf-hd-approx-posterior} is not exact sampling from
\eqref{eq:bf-hd-exact-posterior}.

\section{Why Dimension Breaks Filters}
\label{sec:bf-hd-failure-modes}

High dimension is a family of failure mechanisms.

\begin{description}[leftmargin=*, style=nextline]
  \item[Particle collapse.]
  The bootstrap and SIR filters of \citet{gordon1993novel} and
  \citet{arulampalam2002tutorial} represent the filtering law by particles and
  weights.  \citet{bengtsson2008curse} and \citet{snyder2008obstacles} show
  that, under their assumptions, weight concentration can be governed by the
  variance of the log likelihood and can require enormous ensembles.  These
  results do not say every structured particle method fails; they say proposal,
  localization, and correction must be central.

  \item[Grid explosion.]
  A density grid with $M$ points per coordinate has $M^n$ entries.  Sparse grids,
  tensor trains, and transport maps are attempts to exploit smoothness,
  low-order interactions, low rank, or triangular structure.  If the posterior
  lacks that structure, the format fails.

  \item[Gaussian projection bias.]
  EKF/UKF/CKF-style filters propagate only selected moments.  Multimodality,
  skewness, sharp likelihoods, and nonlocal curvature can be invisible to the
  chosen moment rule.

  \item[Sampler geometry.]
  HMC and NUTS can fail through divergences, poor energy exploration, invalid
  regions, or a value/gradient mismatch even when the filter returns finite
  numbers.
\end{description}

\section{Defects Passed To Synthesis}
\label{sec:bf-hd-foundations-defects}

This chapter passes five defects to the synthesis chapter.

\begin{enumerate}[label=(\arabic*)]
  \item The exact object is a sequence of probability laws and normalizers.
  A later method that returns a mean but not a defensible \(Z_t\) has not
  supplied a likelihood target for parameter inference.
  \item The approximate posterior in \eqref{eq:bf-hd-approx-posterior} is a
  different target unless an exact correction is part of the method.  This is
  the first same-scalar obligation inherited by HMC.
  \item Analytical score recursions such as
  \eqref{eq:bf-hd-likelihood-gradient} are available in the dominated
  discrete-time setting.  Continuous-time/PDE score and adjoint formulas are
  not derived here and remain source-specific obligations.
  \item Density/PDE formulations require mass, positivity, boundary, and
  regularity checks.  A low residual in a transformed PDE does not by itself
  prove that the normalized density is a valid filtering law.
  \item Source-local PDE assumptions matter.  The Davis and Yau robust-DMZ
  transformations and the Meng tensor approximation results support the
  statements made under their hypotheses; they do not license an unrestricted
  high-dimensional filtering theorem for BayesFilter.
\end{enumerate}

The synthesis chapter uses these defects as gates: exact target first,
approximation target second, sampler or acceleration only downstream.

\paragraph{Exported to synthesis.}
\begingroup
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{longtable}{@{}p{0.10\linewidth}p{0.13\linewidth}p{0.12\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.08\linewidth}p{0.11\linewidth}p{0.12\linewidth}@{}}
\toprule
Object & Exact target & Approximation site & Failure mode & Diagnostic/veto &
Cost variable & Promotion gate & Citation/derivation anchor \\
\midrule
\endhead
Conditional law & \(p_\theta(x_t\mid y_{1:t})\) & density, particles,
moments, tensor, or map & mean-only or invalid law & mass, positivity,
reference moments & state dimension & law semantics named &
\eqref{eq:bf-hd-prediction}--\eqref{eq:bf-hd-update} \\
Normalizer & \(Z_t=p(y_t\mid y_{1:t-1})\) & quadrature, Monte Carlo,
tensor/PDE solve & wrong likelihood & finite positive \(Z_t\), scalar parity &
integral cost & scalar likelihood reported &
\eqref{eq:bf-hd-update}--\eqref{eq:bf-hd-likelihood-factor} \\
Likelihood gradient & \(\nabla_\theta\ell_T\) & sensitivity approximation or
autodiff path & gradient of a different scalar & finite-difference parity,
same-scalar check & filter-gradient cost & value and gradient match &
Proposition~\ref{prop:bf-hd-likelihood-sensitivity} \\
DMZ/PDE density & unnormalized filtering density & spatial discretization,
operator compression & mass, boundary, or regularity drift & residual, mass,
positivity, boundary checks & grid/rank & source assumptions checked &
\citet{davis1980multiplicative,yau2000realtime,yau2008realtime} \\
\bottomrule
\end{longtable}
\endgroup

\section{Methodological Boundary}
\label{sec:bf-hd-foundations-boundary}

The chapters that follow are literature-grounded design chapters, not production
promotion notes.  A numerical smoke test can show that a command ran, values
were finite, shapes matched, or a stop rule triggered.  It cannot establish
posterior accuracy, NAWM readiness, tensor validation, HMC convergence, GPU/XLA
readiness, or a default production policy.

\section{Bibliographic Limits}
\label{sec:bf-hd-foundations-limits}

Several historical or specialized sources were not available for direct
technical inspection in this pass: Savostyanov-specific maxvol/quasioptimality,
Stroud's book, Smolyak's original sparse-grid construction, Genz-specific
cubature foundations, and Knothe's original rearrangement source.  When later
chapters use successor or alternative sources, their claims are restricted to
what those inspected sources actually state.
