\chapter[Gaussian And High-Order Filters]{Gaussian Projection and High-Order Quadrature Filters}
\label{ch:bf-highdim-gaussian-quadrature}

\section{What This Chapter Computes}
\label{sec:bf-hd-gq-purpose}

This chapter is about a deliberately limited object.  It does not compute the
full nonlinear filtering law
\[
  p(x_t\mid y_{1:t}).
\]
It replaces that law by a Gaussian moment pair
\[
  p(x_t\mid y_{1:t}) \quad\leadsto\quad \calN(m_t,P_t),
\]
and it computes the Gaussian-weighted expectations needed to update
\((m_t,P_t)\).  The result is a useful baseline and diagnostic, not a theorem
that the posterior is Gaussian.

The notation is fixed for the rest of the chapter.  Before observing \(y_t\),
the predictive approximation is
\[
  X\sim \calN(m^-,P^-),\qquad P^-=CC^\top ,
\]
where \(C\) is a chosen covariance factor.  The observation equation is
\[
  Y=h_\theta(X)+V,\qquad V\sim\calN(0,R_\theta),
\]
with \(V\) independent of \(X\).  The exact Bayesian update would multiply the
predictive density by \(g_\theta(y_t\mid x_t)\) and renormalize.  A Gaussian
projection filter instead computes a finite list of moments of \(X\) and
\(h_\theta(X)\), then performs a Kalman-style affine correction.

Here are the terms before they are used.

\begin{description}[leftmargin=0.25in]
  \item[Sigma point.] A deterministic point chosen so that a weighted cloud has
  prescribed mean and covariance.  Julier--Uhlmann use this language for the
  unscented transform.
  \item[Cubature point.] A deterministic point in a numerical integration rule.
  In the CKF literature it is usually chosen to integrate polynomials under a
  Gaussian weight up to a stated degree.
  \item[Degree.] A degree-\(d\) rule integrates every polynomial of total degree
  at most \(d\) exactly under the stated weight.  Degree is a property of a
  quadrature formula, not a posterior-accuracy guarantee.
  \item[Gaussian weight.] The density of a standard normal variable
  \(\xi\sim\calN(0,I)\).  A Gaussian expectation is written as
  \[
    \E[\varphi(X)]
    =
    \int \varphi(m^-+C\xi)\phi(\xi)\,d\xi .
  \]
  \item[Tensor-product GHQ.] A multivariate Gauss--Hermite quadrature rule
  formed by taking all combinations of one-dimensional Gauss--Hermite nodes.
  \item[Sparse-grid index set.] A selected set of tensor-product levels.  It
  keeps some coordinate interactions and omits others to reduce point count.
  \item[Innovation likelihood.] The Gaussian density assigned to the residual
  \(v_t=y_t-\bar z_t\) with covariance \(S_t\).  It is an approximate scalar
  unless the model is in an exact linear-Gaussian case.
  \item[Approximate score.] The derivative of the reported approximate scalar.
  It is not the derivative of the exact nonlinear likelihood unless a separate
  correction or exactness proof is supplied.
\end{description}

\paragraph{Running scalar cell.}
The common example is
\begin{equation}
  x_t=\rho x_{t-1}+\eta_t,\qquad y_t=x_t^2+\epsilon_t,
  \label{eq:bf-hd-gq-running-cell}
\end{equation}
with Gaussian innovations.  If \(X\sim\calN(m^-,P^-)\), then the exact
likelihood factor is proportional to
\[
  \exp\{-(y-X^2)^2/(2R)\}.
\]
For \(y>0\) and small \(R\), the posterior can have mass near
\(+\sqrt y\) and \(-\sqrt y\).  A Gaussian filter cannot carry those two lobes.
It can only carry a mean, covariance, innovation scalar, and diagnostics.  This
simple cell will be used below as a truth-telling device: whenever a method
computes moments accurately but still cannot represent two lobes, the reader can
see the exact limitation.

\section{Gaussian Moment Projection}
\label{sec:bf-hd-moment-projection}

\paragraph{Problem setting.}
Start with the predictive Gaussian \(X\sim\calN(m^-,P^-)\).  The method first
computes the joint moments of \(X\) and the predicted observation
\[
  Z=h_\theta(X).
\]
Define
\begin{align}
  \bar z &= \E[Z],&
  S &= R_\theta+\E[(Z-\bar z)(Z-\bar z)^\top],&
  C_{xz}&=\E[(X-m^-)(Z-\bar z)^\top].
  \label{eq:bf-hd-gq-moments}
\end{align}
The Gaussian projection update is
\begin{align}
  K &= C_{xz}S^{-1},&
  m &= m^-+K(y-\bar z),&
  P &= P^- - KSK^\top .
  \label{eq:bf-hd-gq-update}
\end{align}
Jia--Xin--Cheng write the same Gaussian-approximation filtering structure as
Gaussian weighted integrals in their sparse-grid paper and high-degree CKF
paper \citep{jia2012sparsegrid,jia2013highdegree}.  The derivation below is a
BayesFilter derivation of the affine projection once those moments are given.

\begin{proposition}[Affine moment projection]
\label{prop:bf-hd-affine-projection}
Among estimators of \(X\) of the form \(a+BY\), the minimum mean-square
estimator under the joint moments in \eqref{eq:bf-hd-gq-moments} has
\(B=C_{xz}S^{-1}\), provided \(S\) is nonsingular.  Interpreting the resulting
mean and residual covariance as a Gaussian gives \eqref{eq:bf-hd-gq-update}.
\end{proposition}

\begin{proof}
Center \(\tilde X=X-m^-\) and \(\tilde Y=Y-\bar z\).  The estimator can be
written as \(m^-+B\tilde Y\).  The mean-square objective is
\[
  J(B)=\E\|\tilde X-B\tilde Y\|^2
  =\tr(P^-)-2\tr(BC_{xz}^\top)+\tr(BSB^\top).
\]
The derivative is \(\nabla_BJ=-2C_{xz}+2BS\).  The stationary point is
\(B=C_{xz}S^{-1}\).  Substituting this \(B\) gives residual covariance
\[
  \E[(\tilde X-C_{xz}S^{-1}\tilde Y)
     (\tilde X-C_{xz}S^{-1}\tilde Y)^\top]
  =P^- - C_{xz}S^{-1}C_{xz}^\top ,
\]
which is the covariance in \eqref{eq:bf-hd-gq-update}.
\end{proof}

\paragraph{What is exact and what is approximate.}
The affine theorem is exact for the moments it is given.  The approximation is
earlier: the posterior law is compressed to the moments
\((\bar z,S,C_{xz})\), and those moments may themselves be approximated by
Taylor expansion or quadrature.

\paragraph{Running scalar example.}
For \(h(x)=x^2\) and \(X\sim\calN(0,P^-)\),
\[
  C_{xz}=\Cov(X,X^2)=\E[X^3]-\E[X]\E[X^2]=0.
\]
The Gaussian update therefore leaves the mean at zero regardless of \(y\), even
when the exact posterior has two separated lobes.  The affine projection has not
made an algebra mistake.  It has solved a compressed moment problem that cannot
store the posterior shape.

\paragraph{Cost and gradient status.}
After \(\bar z,S,C_{xz}\) have been computed, the dominant dense operation is
solving with \(S\) and updating \(P\).  If the moment computation is a fixed
smooth rule, Section~\ref{sec:bf-hd-gq-gradient-contract} gives the derivative
of the approximate innovation scalar.  If the rule changes branches, the same
formula is only branch-local.

\section{Extended Kalman Filter (EKF) and Second-Order Taylor Filters}
\label{sec:bf-hd-derivative-filters}

\paragraph{Problem setting.}
Derivative filters approximate the moments in
\eqref{eq:bf-hd-gq-moments} by expanding \(h_\theta\) around \(m^-\).  This
section is a project derivation in the notation of this chapter.  It is included
because Julier--Uhlmann explicitly motivate the unscented transform as an
alternative to EKF linearization \citep{julier1997new}, but the formulas below
are not claimed as a new result from that paper.

Write \(X=m^-+\delta\), with \(\E[\delta]=0\) and
\(\E[\delta\delta^\top]=P^-\).  Let \(J=D_xh_\theta(m^-)\).  The first-order
Taylor approximation is
\[
  h_\theta(m^-+\delta)\approx h_\theta(m^-)+J\delta .
\]
Therefore the extended Kalman filter moment approximation is
\begin{align}
  \bar z_{\rm EKF} &= h_\theta(m^-),&
  S_{\rm EKF} &= R_\theta+JP^-J^\top,&
  C_{xz,{\rm EKF}} &= P^-J^\top .
  \label{eq:bf-hd-ekf-moments}
\end{align}

For a second-order Taylor filter, write the \(r\)th observation component as
\[
  h_r(m^-+\delta)
  =
  h_r(m^-)+J_r\delta+\frac12\delta^\top H_r\delta
  +\operatorname{rem}_r(\delta),
\]
where \(H_r=D_x^2h_r(m^-)\).  If the remainder is dropped and \(\delta\) is
Gaussian, then
\begin{equation}
  \bar z_r
  \approx
  h_r(m^-)+\frac12\tr(H_rP^-).
  \label{eq:bf-hd-second-order-mean}
\end{equation}
For observation components \(r,s\), the same Gaussian moment calculation gives
\begin{equation}
  \Cov(h_r(X),h_s(X))
  \approx
  J_rP^-J_s^\top
  +\frac12\tr(H_rP^-H_sP^-).
  \label{eq:bf-hd-second-order-cov}
\end{equation}
The cross covariance with \(X\) remains \(P^-J^\top\) at this expansion point,
because the third central moment of a Gaussian is zero.

\paragraph{Running scalar example.}
For \(h(x)=x^2\), \(J=2m^-\) and \(H=2\).  The EKF gives
\[
  \bar z_{\rm EKF}=(m^-)^2,\qquad
  C_{xz,{\rm EKF}}=2m^-P^-.
\]
At \(m^-=0\), it cannot move the state mean because \(C_{xz}=0\).  The
second-order mean gives \(\bar z=(m^-)^2+P^-\), and for this quadratic cell the
second-order covariance formula recovers
\[
  \Var(X^2)=2(P^-)^2+4(m^-)^2P^-.
\]
Even so, after the update the filter stores only one Gaussian.  The Taylor
filter can improve local moments without representing the two-lobed posterior.

\paragraph{What is exact and what is approximate.}
The Taylor identities are exact only up to the displayed remainder.  Dropping
the remainder, and then storing the result as one Gaussian, are the
approximations.  The first-order EKF is exact for affine \(h_\theta\); the
second-order formulas are local moment formulas, not posterior-shape formulas.

\paragraph{Computational cost.}
The EKF needs Jacobians and dense innovation solves.  The second-order filter
needs Hessian contractions such as \(\tr(H_rP^-)\) and
\(\tr(H_rP^-H_sP^-)\), which are too expensive for large dense blocks unless
the model supplies exploitable sparsity, Hessian-vector products, or a small
active subspace.

\paragraph{Limitations.}
The EKF can be misleading when the predictive mean lies at a point where the
Jacobian is uninformative, as in the scalar quadratic cell at \(m^-=0\).  The
second-order filter repairs selected local moments, but it can still be
overconfident after Gaussian projection and can be numerically impractical when
Hessian contractions are dense.

\paragraph{Analytical gradient status.}
If the Taylor moments define a fixed smooth approximate innovation scalar, the
score in Section~\ref{sec:bf-hd-gq-gradient-contract} applies after replacing
the quadrature sums by Taylor-derived \(\bar z\) and \(S\).  An iterated EKF
with data-dependent convergence branches is only a local branch calculation
unless the iteration policy is part of a declared smooth scalar.

\section{Julier-Uhlmann Unscented Transform and Unscented Kalman Filter (UKF)}
\label{sec:bf-hd-ukf}

\paragraph{Problem setting.}
The unscented transform replaces Taylor expansion with a deterministic weighted
cloud.  Julier--Uhlmann construct \(2n+1\) sigma points for a random vector with
mean and covariance, transform those points through the nonlinear map, and take
weighted mean and covariance of the transformed cloud
\citep[Eq. 12--14]{julier1997new}.

\paragraph{Original construction in BayesFilter notation.}
Let \(n=n_x\), \(P^-=CC^\top\), and choose \(\kappa\) with \(n+\kappa>0\).
Set \(\gamma=\sqrt{n+\kappa}\).  The standardized offsets and weights are
\begin{align}
  \xi^{(0)}&=0,&
  w_0&=\frac{\kappa}{n+\kappa},\\
  \xi^{(j)}&=\gamma e_j,&
  w_j&=\frac{1}{2(n+\kappa)},\qquad j=1,\ldots,n,\\
  \xi^{(j+n)}&=-\gamma e_j,&
  w_{j+n}&=\frac{1}{2(n+\kappa)},\qquad j=1,\ldots,n.
  \label{eq:bf-hd-ut-points}
\end{align}
The state-space sigma points are
\[
  \chi^{(r)}=m^-+C\xi^{(r)},\qquad
  z^{(r)}=h_\theta(\chi^{(r)}).
\]
Then the UT observation moments are
\begin{align}
  \bar z_{\rm UT}&=\sum_{r=0}^{2n}w_rz^{(r)},\\
  S_{\rm UT}&=R_\theta+
    \sum_{r=0}^{2n}w_r(z^{(r)}-\bar z_{\rm UT})
      (z^{(r)}-\bar z_{\rm UT})^\top,\\
  C_{xz,{\rm UT}}&=
    \sum_{r=0}^{2n}w_r(\chi^{(r)}-m^-)
      (z^{(r)}-\bar z_{\rm UT})^\top .
  \label{eq:bf-hd-ut-moments}
\end{align}
These moments are inserted into \eqref{eq:bf-hd-gq-update}.

\paragraph{Derivation of the moment match.}
The weights sum to one.  Symmetry gives \(\sum_rw_r\xi^{(r)}=0\).  For the
covariance,
\[
  \sum_rw_r\xi^{(r)}\xi^{(r)\top}
  =
  \sum_{j=1}^n
  \frac{1}{2(n+\kappa)}
  \{(n+\kappa)e_je_j^\top+(n+\kappa)e_je_j^\top\}
  =I .
\]
Thus the point cloud has the predictive mean and covariance exactly.  What is
not exact is the transformed distribution after \(h_\theta\), except in special
cases.

\paragraph{Running scalar example.}
Let \(m^-=0\), \(P^-=P\), and \(h(x)=x^2\).  The nonzero transformed points have
value \(P(1+\kappa)\), while the center has value zero.  The UT mean is
\[
  \bar z_{\rm UT}
  =
  \frac{1}{1+\kappa}P(1+\kappa)=P.
\]
The UT variance of \(z=x^2\) is \(\kappa P^2\).  The true Gaussian variance is
\(2P^2\), so in this scalar cell the common Gaussian choice \(1+\kappa=3\)
matches the variance.  Matching this local moment still does not create a
two-lobed posterior representation.

\paragraph{What is exact and what is approximate.}
The input point cloud matches the predictive mean and covariance exactly.  The
weighted transformed cloud is an approximation to the transformed moments of
\(h_\theta(X)\), and the final update is still a Gaussian projection.  The UKF
does not claim to represent the exact posterior law after a nonlinear
observation.

\paragraph{Computational cost and limitations.}
The UT uses \(2n+1\) function evaluations per transformed Gaussian block plus
the covariance factorization and dense solve.  Julier--Uhlmann note that
\(\kappa\) may be negative and that negative choices can create covariance
semantics problems \citep{julier1997new}.  In this chapter that becomes a
practical veto: if the weighted covariance or innovation covariance is not
positive definite, the approximate likelihood is not a valid HMC scalar.

\paragraph{Use in parameter inference.}
If the UKF likelihood is used inside HMC, the sampler is not targeting the
exact nonlinear filtering likelihood.  It is targeting the approximate
Gaussian innovation scalar derived in
Section~\ref{sec:bf-hd-gq-gradient-contract}.  The derivative is meaningful
only when \(\kappa\), the sigma-point offsets, the weights, and the covariance
factor branch are fixed smooth parts of that scalar.  If a factor floor, active
subspace, or tuning rule changes discontinuously with \(\theta\), the reported
score is at best local to that branch and should not be presented as the
gradient of a single smooth target.

\section{Arasaratnam-Haykin Cubature Kalman Filter (CKF)}
\label{sec:bf-hd-ckf}

\paragraph{Problem setting.}
The cubature Kalman filter computes the same Gaussian projection moments, but
uses a third-degree spherical--radial cubature rule.  Arasaratnam--Haykin
derive the rule for Gaussian-weighted integrals and embed it in a CKF and
square-root CKF \citep{arasaratnam2009cubature}.

\paragraph{Original construction in BayesFilter notation.}
For \(\xi\sim\calN(0,I_n)\), the CKF rule uses
\begin{align}
  \xi^{(j)}&=\sqrt n\,e_j,&
  \xi^{(j+n)}&=-\sqrt n\,e_j,&
  w_j&=w_{j+n}=\frac{1}{2n},
  \qquad j=1,\ldots,n .
  \label{eq:bf-hd-ckf-points}
\end{align}
There is no center point.  The transformed points
\(\chi^{(r)}=m^-+C\xi^{(r)}\) and \(z^{(r)}=h_\theta(\chi^{(r)})\) are inserted
into the same formulas as \eqref{eq:bf-hd-ut-moments}.

\paragraph{Derivation of the third-degree moment rule.}
The weights sum to one and symmetry gives zero mean.  The covariance is
\[
  \sum_{j=1}^n \frac{1}{2n}
  \{(\sqrt n e_j)(\sqrt n e_j)^\top
    +(-\sqrt n e_j)(-\sqrt n e_j)^\top\}
  =I .
\]
All odd monomials of degree one or three vanish by sign symmetry.  Quadratic
cross terms vanish because each point lies on a coordinate axis.  This is the
source of the third-degree cubature interpretation.  The rule is a rule for
Gaussian integrals, not a representation of the updated posterior.

\paragraph{Running scalar example.}
When \(n=1\), the CKF points are \(\xi=\pm1\) with weights \(1/2\).  For
\(m^-=0\) and \(h(x)=x^2\), both transformed observation points equal \(P\).
Hence
\[
  \bar z_{\rm CKF}=P,\qquad
  \Var_{\rm CKF}(z)=0.
\]
The exact variance of \(X^2\) is \(2P^2\).  The CKF integrates up to degree
three, but the innovation variance for a quadratic observation depends on a
fourth moment.  This is why "third-degree" is not enough to judge posterior or
likelihood quality.

\paragraph{What is exact and what is approximate.}
The third-degree cubature statement is exact for the polynomial class under the
standard Gaussian weight.  The approximation is using that rule for the
nonlinear moment integrals and then projecting the result to one Gaussian.

\paragraph{Computational cost and limitations.}
The CKF uses \(2n\) model evaluations.  Arasaratnam--Haykin emphasize that this
is linear in function evaluations, while dense square-root and solve operations
still scale cubically in the state or observation block dimension
\citep{arasaratnam2009cubature}.  CKF is therefore a compact local baseline,
not a complete cure for high dimensionality.

\paragraph{Limitations.}
The CKF can be too low-order for likelihoods whose decisive information enters
through fourth and higher moments.  It also inherits the Gaussian-projection
failure: even perfect third-degree integration does not store multimodality,
tail behavior, or support constraints.

\paragraph{Use in parameter inference.}
With a fixed cubature rule and a smooth factor branch, CKF supplies a
well-defined approximate Gaussian innovation scalar and the derivative formula
in Section~\ref{sec:bf-hd-gq-gradient-contract} differentiates that scalar.
A square-root implementation with hard floors, pivot changes, or failed
positive-definite factors does not define the same smooth function of
\(\theta\) unless those branch choices are explicitly included in the scalar.

\section{Jia-Xin-Cheng High-Degree Cubature Kalman Filter (High-Degree CKF)}
\label{sec:bf-hd-high-degree-cubature}

\paragraph{Problem setting.}
High-degree CKF changes the cubature rule while keeping the same Gaussian
projection update.  Jia--Xin--Cheng define degree-\(d\) cubature rules, derive
arbitrary-degree spherical--radial constructions, give explicit third- and
fifth-degree rules, and apply them to Gaussian approximation filtering
\citep[Definition 3.1, Theorem 3.1, Propositions 3.1--3.2]{jia2013highdegree}.

\paragraph{Original construction.}
A degree-\(d\) rule is exact for all polynomials of total degree at most \(d\)
under the stated Gaussian weight.  The paper constructs such rules by splitting
the Gaussian integral into a radial integral and a spherical integral.  The
third-degree rule reduces to the CKF rule \eqref{eq:bf-hd-ckf-points}.  The
fifth-degree rule has \(2n^2+1\) points.

To write the fifth-degree rule compactly, for each pair \(1\le a<b\le n\) set
\[
  s_{ab}^+=\frac{e_a+e_b}{\sqrt2},\qquad
  s_{ab}^-=\frac{e_a-e_b}{\sqrt2},\qquad
  \alpha=\sqrt{n+2}.
\]
The rule for \(I(g)=\E[g(\xi)]\), \(\xi\sim\calN(0,I_n)\), is
\begin{align}
  I(g)\approx&
  \frac{2}{n+2}g(0)
  +\frac{1}{(n+2)^2}
    \sum_{a<b}\sum_{\sigma\in\{-1,1\}}
    \{g(\sigma\alpha s_{ab}^+)+g(\sigma\alpha s_{ab}^-)\}
  \nonumber\\
  &+
  \frac{4-n}{2(n+2)^2}
    \sum_{j=1}^n
    \{g(\alpha e_j)+g(-\alpha e_j)\}.
  \label{eq:bf-hd-fifth-degree-rule}
\end{align}
This is the BayesFilter notation version of the explicit fifth-degree rule in
Jia--Xin--Cheng \citep[Eq. 46]{jia2013highdegree}.  After choosing these
points and weights, the filter again computes \(\bar z,S,C_{xz}\) and applies
\eqref{eq:bf-hd-gq-update}.

\paragraph{Derivation of selected moments.}
The weights in \eqref{eq:bf-hd-fifth-degree-rule} sum to one.  Symmetry makes
all odd moments vanish.  For coordinate \(i\), the second moment receives
\((4-n)/(n+2)\) from the two axis points and \(2(n-1)/(n+2)\) from the pair
points involving coordinate \(i\), hence equals one.  The fourth moments also
match:
\[
  \E_{\rm rule}[\xi_i^4]=(4-n)+(n-1)=3,
\]
and for \(i\ne j\),
\[
  \E_{\rm rule}[\xi_i^2\xi_j^2]=1.
\]
These are the moments that the third-degree CKF misses in the scalar quadratic
cell.

\paragraph{Running scalar example.}
For \(n=1\), the pair terms disappear and the fifth-degree rule becomes
\[
  \frac23 g(0)+\frac16 g(\sqrt3)+\frac16 g(-\sqrt3),
\]
the familiar three-point Gauss--Hermite rule for a standard normal.  Thus it
matches \(\E[X^2]\) and \(\Var(X^2)\) in the scalar cell, while still projecting
the posterior back to one Gaussian.

\paragraph{What is exact and what is approximate.}
The source-local exactness claim is polynomial exactness under the Gaussian
weight for the constructed rule.  The approximate objects are the nonlinear
moments outside that polynomial class, the Gaussian innovation scalar, and the
single-Gaussian posterior representation.

\paragraph{Computational cost and limitations.}
For fixed degree \(2m+1\), Jia--Xin--Cheng prove that the number of cubature
points grows polynomially in dimension with highest degree \(m\)
\citep[Proposition 3.2]{jia2013highdegree}.  For the explicit fifth-degree
rule, that is \(2n^2+1\) points.  The limitation is numerical as well as
statistical: the fifth-degree axis weight
\((4-n)/[2(n+2)^2]\) is negative for \(n>4\), and the paper records the
stability concern for high-degree rules \citep[Remark 3.6]{jia2013highdegree}.

\paragraph{Limitations.}
High degree is useful only when the decisive local integrand is smooth enough
and the active block is small enough for the point set.  Negative weights can
make covariance assembly fragile, and polynomial exactness can still miss
nonsmooth likelihoods, tail events, and multimodal posterior structure.

\paragraph{Use in parameter inference.}
A fixed high-degree rule can be differentiated as an approximate likelihood
when the degree, points, weights, model maps, and factor branch remain fixed
and smooth.  If the implementation changes degree, prunes points, floors
covariances, or switches active subspaces while the sampler moves through
parameter space, then the target itself has changed.

\section{Transformed Stochastic Volatility and Gaussian-Mixture CUT4}
\label{sec:bf-hd-sv-mixture-cut4}

\paragraph{Why the direct CUT4 path is not the SV likelihood.}
The scalar stochastic-volatility benchmark used in the Zhao--Cui validation
line can be written as
\begin{align}
  X_t &= \gamma X_{t-1}+\sigma \eta_t,\qquad
    \eta_t\sim\calN(0,1),\\
  Y_t &= \beta \exp(X_t/2)\epsilon_t,\qquad
    \epsilon_t\sim\calN(0,1).
  \label{eq:bf-hd-sv-native}
\end{align}
The observation variance is state dependent:
\[
  Y_t\mid X_t=x_t \sim
  \calN\{0,\beta^2\exp(x_t)\}.
\]
This is not the additive-Gaussian observation equation assumed at the start of
this chapter.  Therefore a direct call to a Gaussian CUT4 observation update
would test a different model.

\paragraph{The log-squared transform.}
For \(Y_t\ne0\), define
\[
  Z_t=\log(Y_t^2).
\]
Then \eqref{eq:bf-hd-sv-native} gives
\begin{align}
  Z_t
  &= \log(\beta^2)+X_t+\log(\epsilon_t^2) \nonumber\\
  &= \log(\beta^2)+X_t+U_t,
  \qquad U_t=\log(\epsilon_t^2).
  \label{eq:bf-hd-sv-log-square}
\end{align}
The transformed observation is additive, but the noise \(U_t\) is not Gaussian.
Kim--Shephard--Chib use this transformed representation together with an
approximating offset normal mixture and an importance reweighting correction in
their likelihood-based SV framework
\citep[Section~3, Table~4]{kim1998stochasticvolatility}.  The construction
below uses only the approximating mixture as a BayesFilter comparator.  It does
not implement the importance reweighting correction and therefore is not
promoted as the exact native SV likelihood.

In numerical code one often replaces \(Z_t=\log(Y_t^2)\) by
\[
  Z_t^{(c)}=\log(Y_t^2+c),\qquad c>0,
  \label{eq:bf-hd-sv-log-square-offset}
\]
to avoid an infinite value when an observation is exactly zero.  This offset is
a declared approximation and must appear in the run manifest.

\paragraph{Gaussian-mixture approximation.}
Let a finite mixture approximate the law of \(U_t\):
\[
  p_U(u)\approx \sum_{j=1}^J \pi_j
  \varphi(u;m_j,v_j),
  \qquad
  \pi_j>0,\quad \sum_{j=1}^J\pi_j=1,\quad v_j>0.
  \label{eq:bf-hd-sv-logchi-mixture}
\]
Kim--Shephard--Chib tabulate component locations \(a_j\) for the shifted
variable \(U_t+1.2704\); equivalently, the component means for \(U_t\) itself
are \(m_j=a_j-1.2704\).  The P39 implementation pins the following
seven-component table and uses the effective means \(m_j\):
\begin{equation}
\begin{array}{c|rrrr}
j & \pi_j & a_j & m_j=a_j-1.2704 & v_j\\
\hline
1 & 0.00730 & -10.12999 & -11.40039 & 5.79596\\
2 & 0.10556 &  -3.97281 &  -5.24321 & 2.61369\\
3 & 0.00002 &  -8.56686 &  -9.83726 & 5.17950\\
4 & 0.04395 &   2.77786 &   1.50746 & 0.16735\\
5 & 0.34001 &   0.61942 &  -0.65098 & 0.64009\\
6 & 0.24566 &   1.79518 &   0.52478 & 0.34023\\
7 & 0.25750 &  -1.08819 &  -2.35859 & 1.26261
\end{array}
\label{eq:bf-hd-sv-ksc-pinned-table}
\end{equation}
These constants are the declared BayesFilter approximation contract for this
phase.  The code records them in the run diagnostics and tests only their
finite-mixture consequences.  The implementation also checks that the shifted
fixture has mean close to \(-1.2704\) and variance close to \(\pi^2/2\), the
first two moments of \(\log\chi_1^2\).

Introduce a discrete component \(S_t\in\{1,\ldots,J\}\) with
\(\Pr(S_t=j)=\pi_j\).  Conditional on \(S_t=j\), the transformed observation is
linear Gaussian:
\begin{equation}
  Z_t\mid X_t=x_t,S_t=j
  \sim
  \calN\{\log(\beta^2)+x_t+m_j,\;v_j\}.
  \label{eq:bf-hd-sv-component-observation}
\end{equation}
This is the least-disruptive route back to the Gaussian quadrature machinery:
each mixture component is an additive-Gaussian observation model, and the
non-Gaussian skewness and tail shape enter through the mixture weights,
locations, and variances rather than through a new non-Gaussian cubature rule.

\paragraph{Component CUT4 update.}
CUT4 denotes the named conjugate unscented transform family of
Adurthi--Singla--Singh.  The BayesFilter implementation used in this phase is
the positive-weight SVD-placed CUT4-G rule from
Chapter~\ref{ch:bf-svd-sigma-point}.  Its code diagnostics report polynomial
degree \(5\).  For an augmented stochastic integration dimension \(q\ge3\),
the rule uses
\[
  N_{\rm CUT4}=2q+2^q
\]
standard-normal points and weights to approximate Gaussian expectations in the
projection moments.  Each component update below feeds the conditional
Gaussian model for component \(j\) into that rule and obtains a component log
normalizer and Gaussian retained moments.  CUT4 exactness is a statement about
Gaussian polynomial integration for the implemented point rule
\citep{adurthi2012cut,adurthi2012cutfiltering}; it is not by itself a proof
that the collapsed SV posterior is exact.

Suppose that before observing \(Z_t\) the predictive approximation is
\[
  X_t\mid Z_{1:t-1}\approx\calN(m_t^-,P_t^-).
\]
For component \(j\), apply a Gaussian projection update using the observation
map
\[
  h_j(x)=\log(\beta^2)+x+m_j,
  \qquad R_j=v_j.
\]
Let the component CUT4 update return
\[
  \ell_{tj},\qquad m_{tj},\qquad P_{tj},
  \label{eq:bf-hd-sv-component-cut4}
\]
where \(\ell_{tj}\) is the approximate component log normalizer and
\((m_{tj},P_{tj})\) is the component Gaussian posterior.  The mixture
normalizer is
\begin{equation}
  \ell_t
  =
  \log\sum_{j=1}^J \pi_j\exp(\ell_{tj}),
  \label{eq:bf-hd-sv-mixture-normalizer}
\end{equation}
and the posterior component weights are
\begin{equation}
  \alpha_{tj}
  =
  \frac{\pi_j\exp(\ell_{tj})}{\sum_{k=1}^J\pi_k\exp(\ell_{tk})}.
  \label{eq:bf-hd-sv-component-weights}
\end{equation}
If the filter is forced back to a single Gaussian for the next prediction, the
retained moments are
\begin{align}
  m_t &= \sum_{j=1}^J \alpha_{tj}m_{tj},\\
  P_t &=
  \sum_{j=1}^J \alpha_{tj}
  \{P_{tj}+(m_{tj}-m_t)(m_{tj}-m_t)^\top\}.
  \label{eq:bf-hd-sv-mixture-moment-collapse}
\end{align}
Equations \eqref{eq:bf-hd-sv-mixture-normalizer}--\eqref{eq:bf-hd-sv-mixture-moment-collapse}
are exact for the finite mixture approximation and the component Gaussian
updates they are given.  The approximations are the mixture replacement for
the log-chi-square law, the CUT4 component integration when the transition or
observation is nonlinear, and the collapse of the mixture posterior back to
one Gaussian.

\paragraph{Simple SV recursion.}
For the simple SV model \eqref{eq:bf-hd-sv-native}, the prediction before
component updates is
\[
  m_t^-=\gamma m_{t-1},\qquad
  P_t^-=\gamma^2P_{t-1}+\sigma^2,
  \label{eq:bf-hd-sv-simple-predict}
\]
with the stationary initial variance \(P_0^-=\sigma^2/(1-\gamma^2)\) when
\(|\gamma|<1\).  Because \eqref{eq:bf-hd-sv-component-observation} is linear in
\(x_t\), the component update is Kalman-exact for the finite mixture model; a
CUT4 implementation should agree with a dense transformed-mixture reference on
small scalar fixtures.  Agreement with the native likelihood is a separate
question, because the finite mixture is only an approximation to
\(\log(\epsilon_t^2)\).  These scalar fixtures validate the transformation,
mixture bookkeeping, log-sum-exp normalization, moment collapse, and CUT4
reduction to the Gaussian component update.  They do not establish
CUT4-specific accuracy for a nonlinear conditional component; that would
require a same-target nonlinear transformed-mixture fixture.

\paragraph{Generalized SV after Chib--Nardari--Shephard.}
Chib--Nardari--Shephard extend the SV likelihood program to generalized models
with heavy-tailed observation errors, exogenous variables in observation and
volatility equations, and jump components \citep{chib2002mcmcsv}.  In
BayesFilter notation, a broad transformed observation template is
\begin{equation}
  Z_t
  =
  a_\theta(w_t)
  + b_\theta(w_t)^\top X_t
  + c_\theta(w_t,S_t,J_t)
  + E_t,
  \qquad
  E_t\mid S_t,J_t\sim\calN(0,R_{S_t,J_t}),
  \label{eq:bf-hd-sv-generalized-template}
\end{equation}
where \(w_t\) are exogenous covariates, \(S_t\) is a mixture component, and
\(J_t\) denotes an optional jump or tail-state indicator.  Conditional on the
discrete indicators, \eqref{eq:bf-hd-sv-generalized-template} again has an
additive-Gaussian observation form.  The CUT4 recipe is therefore the same:
run a component update for each declared discrete component, combine
normalizers by log-sum-exp, and collapse retained moments only after recording
the mixture weights.

This generalized template is a derivation route, not an implementation claim.
To promote a particular CNS-style model, the project must first inspect the
relevant Chib--Nardari--Shephard primary-source equations and then specify the
component state space, priors, jump/tail indicators, exogenous regressors,
mixture table, and whether an exact correction or reweighting step is used.

\paragraph{Comparison with TT filtering.}
The existing scalar SV TT lane approximates the native SV value path on a tiny
two-observation fixture.  The Gaussian-mixture CUT4 lane approximates the
transformed mixture model.  These are adjacent but not identical algorithms.
They can be compared as an explanatory approximation gap.  They become a
same-target equivalence comparison only after a TT lane is built for the same
transformed mixture likelihood, with the same offset, mixture table, horizon,
and parameter values.

\paragraph{Claim boundary.}
The phrase "CUT4 for SV" is too broad unless the target is named.  In this
chapter it means
\[
  \hbox{CUT4 component updates inside a transformed finite-mixture SV
  approximation.}
\]
It does not mean exact native SV filtering, KSC importance reweighting,
paper-scale S\&P 500 validation, generalized CNS estimation, or a production
default for HMC.

\section{Tensor-Product Gauss-Hermite Quadrature Filtering (GHQF)}
\label{sec:bf-hd-tensor-product-ghq}

\paragraph{Problem setting.}
Tensor-product Gauss--Hermite quadrature filtering is the most literal way to
connect quadrature with Gaussian filtering.  The filter needs Gaussian
expectations such as
\[
  \E[\varphi(X)],\qquad X\sim\calN(m,P),
\]
where \(\varphi\) is the transition map, an outer product of transition
residuals, the observation map, or an observation outer product.  GHQF replaces
each such expectation by a finite weighted sum.  Jia--Xin--Cheng use this
construction as the conventional Gauss--Hermite quadrature filter baseline
before introducing sparse grids \citep[Section 2.2.1]{jia2012sparsegrid}.

For implementability, separate two objects:

\begin{description}[leftmargin=0.25in]
  \item[Standardized rule.] A list of points and weights
  \(\{(\xi^{(r)},w_r)\}_{r=1}^M\) for
  \(\xi\sim\calN(0,I_b)\).
  \item[Placed rule.] The points \(\chi^{(r)}=m+C\xi^{(r)}\) for
  \(X=m+C\xi\sim\calN(m,CC^\top)\).
\end{description}

Once a standardized rule has been built, the filter never integrates directly.
It evaluates the model at the placed points and forms weighted sums.

\paragraph{One-dimensional rule.}
Let \(\{(\zeta_a,\omega_a)\}_{a=1}^s\) be a one-dimensional rule for a standard
normal variable:
\[
  \E[\psi(\xi)]\approx \sum_{a=1}^s\omega_a\psi(\zeta_a),
  \qquad \xi\sim\calN(0,1),\qquad \sum_a\omega_a=1 .
\]
Jia--Xin--Cheng describe obtaining these nodes and weights from a symmetric
tridiagonal matrix and record the usual univariate exactness degree
\(2s-1\) \citep[Section 2.2.1]{jia2012sparsegrid}.  In code, this chapter
assumes the implementation has a routine
\[
  \textsc{OneDimGHQ}(s)\mapsto \{(\zeta_a,\omega_a)\}_{a=1}^s
\]
returning standard-normal, not physicists' \(e^{-x^2}\), weights.  If a library
returns physicists' or probabilists' Hermite nodes with a different
normalization, the implementation must rescale before using the formulas here.

\paragraph{Tensor-product construction.}
In \(b\) dimensions the tensor-product rule enumerates every multi-index
\(\alpha=(\alpha_1,\ldots,\alpha_b)\in\{1,\ldots,s\}^b\):
\begin{equation}
  \xi^\alpha=(\zeta_{\alpha_1},\ldots,\zeta_{\alpha_b}),\qquad
  w^\alpha=\prod_{j=1}^b\omega_{\alpha_j}.
  \label{eq:bf-hd-ghq-tensor-rule}
\end{equation}
For \(X\sim\calN(m,P)\) with \(P=CC^\top\),
\begin{equation}
  \widehat{\E}_{\rm GHQ}[\varphi(X)]
  =
  \sum_{\alpha\in\{1,\ldots,s\}^b}
  w^\alpha\,
  \varphi(m+C\xi^\alpha).
  \label{eq:bf-hd-ghq-expectation}
\end{equation}
This is the only quadrature object needed by the filter.  Prediction,
observation prediction, likelihood evaluation, and update are obtained by
choosing different \(\varphi\)'s.

\paragraph{Prediction step.}
Suppose the transition is
\[
  x_t=f_\theta(x_{t-1})+u_t,\qquad u_t\sim\calN(0,Q_\theta),
\]
and the previous filtering approximation is \(X_{t-1}\sim\calN(m_{t-1},P_{t-1})\).
Choose \(C_{t-1}C_{t-1}^\top=P_{t-1}\), place the GHQ points
\[
  \chi_{t-1}^{\alpha}=m_{t-1}+C_{t-1}\xi^\alpha,
  \qquad
  x_{t|t-1}^{\alpha}=f_\theta(\chi_{t-1}^{\alpha}).
\]
The predicted moments are
\begin{align}
  m_t^- &= \sum_\alpha w^\alpha x_{t|t-1}^{\alpha},\\
  P_t^- &= Q_\theta+
  \sum_\alpha w^\alpha
  (x_{t|t-1}^{\alpha}-m_t^-)
  (x_{t|t-1}^{\alpha}-m_t^-)^\top .
  \label{eq:bf-hd-ghq-predict}
\end{align}

\paragraph{Observation prediction and update.}
Choose \(C_t^-(C_t^-)^\top=P_t^-\), place a new GHQ cloud
\[
  \chi_t^\alpha=m_t^-+C_t^-\xi^\alpha,\qquad
  z_t^\alpha=h_\theta(\chi_t^\alpha).
\]
Then compute
\begin{align}
  \bar z_t&=\sum_\alpha w^\alpha z_t^\alpha,\\
  S_t&=R_\theta+
  \sum_\alpha w^\alpha(z_t^\alpha-\bar z_t)(z_t^\alpha-\bar z_t)^\top,\\
  C_{xz,t}&=
  \sum_\alpha w^\alpha(\chi_t^\alpha-m_t^-)(z_t^\alpha-\bar z_t)^\top,\\
  v_t&=y_t-\bar z_t,\qquad
  \widehat\ell_t=-\frac12\{\log\det S_t+v_t^\top S_t^{-1}v_t+n_y\log(2\pi)\}.
  \label{eq:bf-hd-ghq-update-moments}
\end{align}
If \(S_t\) is positive definite, solve \(K_t=C_{xz,t}S_t^{-1}\) and set
\begin{align}
  m_t&=m_t^-+K_tv_t,\\
  P_t&=P_t^- - K_tS_tK_t^\top .
  \label{eq:bf-hd-ghq-update}
\end{align}
In implementation, symmetrize \(P_t\leftarrow(P_t+P_t^\top)/2\) and require a
positive-semidefinite check or square-root update before exporting \(P_t\).

\begin{algorithm}
\caption{One Tensor-Product GHQF Step}
\label{alg:bf-hd-ghqf-step}
\begin{algorithmic}[1]
\Require \(m_{t-1},P_{t-1},y_t,\theta\); maps \(f_\theta,h_\theta\); covariances
\(Q_\theta,R_\theta\); one-dimensional size \(s\)
\Ensure \(m_t,P_t,\widehat\ell_t\), and diagnostics
\State Build \(\{(\xi^\alpha,w^\alpha)\}\) from \(\textsc{OneDimGHQ}(s)\) by
\eqref{eq:bf-hd-ghq-tensor-rule}
\State Factor \(P_{t-1}=C_{t-1}C_{t-1}^\top\)
\State Place prediction points \(\chi_{t-1}^\alpha=m_{t-1}+C_{t-1}\xi^\alpha\)
and evaluate \(x_{t|t-1}^\alpha=f_\theta(\chi_{t-1}^\alpha)\)
\State Compute \(m_t^-,P_t^-\) from \eqref{eq:bf-hd-ghq-predict}
\State Factor \(P_t^-=C_t^-(C_t^-)^\top\)
\State Place observation points \(\chi_t^\alpha=m_t^-+C_t^-\xi^\alpha\) and
evaluate \(z_t^\alpha=h_\theta(\chi_t^\alpha)\)
\State Compute \(\bar z_t,S_t,C_{xz,t},v_t,\widehat\ell_t\) from
\eqref{eq:bf-hd-ghq-update-moments}
\State If \(S_t\) is not positive definite, return a likelihood/covariance veto
\State Compute \(K_t=C_{xz,t}S_t^{-1}\), update \(m_t,P_t\), symmetrize \(P_t\),
and run the PSD diagnostic
\end{algorithmic}
\end{algorithm}

\paragraph{Cost and role.}
The point count is \(M=s^b\).  With \(s=5\) and \(b=20\), the method asks for
\(5^{20}\) transition evaluations and another \(5^{20}\) observation
evaluations.  That is why tensor-product GHQF is a reference construction and
small-block benchmark, not the high-dimensional default.  Mathematically, a
fixed tensor-product rule defines a differentiable approximate likelihood when
the factor branches are smooth.  Practically, the exponential point count makes
it a tool for small-block parity checks and debugging, not a candidate for the
large high-dimensional filtering path.

\section{Jia-Xin-Cheng Sparse-Grid Quadrature Filter (SGQF)}
\label{sec:bf-hd-sgqf}

\paragraph{Problem setting.}
Sparse-grid quadrature filtering keeps the same Gaussian projection filter as
GHQF but replaces the full tensor-product expectation
\eqref{eq:bf-hd-ghq-expectation} by a signed linear combination of lower-level
tensor products.  Jia--Xin--Cheng define this construction, merge repeated
points, and insert the resulting weighted cloud into the Gaussian approximation
filter \citep[Sections 2--3]{jia2012sparsegrid}.  In this monograph, the fixed
SGQF is the selected high-dimensional approximate likelihood target.

\paragraph{Univariate levels and tensor component rules.}
Let \(I_\ell\) denote a one-dimensional standard-normal quadrature rule at
level \(\ell\):
\[
  I_\ell[\psi]=\sum_{a=1}^{s_\ell}\omega_{\ell,a}\psi(\zeta_{\ell,a}),
  \qquad \ell=1,2,\ldots .
\]
Jia--Xin--Cheng allow several moment-matched choices for the univariate point
sets and discuss \(2\ell-1\) symmetric points for level \(\ell\)
\citep[Section 3.2]{jia2012sparsegrid}.  To make the algorithm below
reproducible, BayesFilter uses the following default level policy:
\[
  s_\ell=2\ell-1,\qquad
  \{(\zeta_{\ell,a},\omega_{\ell,a})\}_{a=1}^{s_\ell}
  =
  \textsc{OneDimGHQ}(2\ell-1),
  \label{eq:bf-hd-sgqf-level-policy}
\]
where \(\textsc{OneDimGHQ}\) returns standard-normal Gauss--Hermite nodes and
weights.  This default is not nested in general, except for repeated nodes such
as zero in symmetric odd rules.  Nesting is therefore not assumed.  Duplicate
node merging below combines only exactly identical standardized nodes under the
chosen numerical tolerance.  A nested moment-matched family can be substituted,
but then the node/weight family and its derivatives must be declared as part of
the scalar.

For a multi-index \(\mathbf i=(i_1,\ldots,i_b)\), define the tensor component
rule
\[
  I_{\mathbf i}[F]
  =
  (I_{i_1}\otimes\cdots\otimes I_{i_b})[F].
\]
It has standardized points
\[
  \xi^{\mathbf i,\alpha}
  =
  (\zeta_{i_1,\alpha_1},\ldots,\zeta_{i_b,\alpha_b}),
  \qquad
  w^{\mathbf i,\alpha}
  =
  \prod_{j=1}^b\omega_{i_j,\alpha_j}.
\]

\paragraph{Fixed sparse-grid index set.}
For level \(L\), Jia--Xin--Cheng write the sparse-grid rule in terms of
multi-indices whose component sums fall in a source-local level band
\citep[Eq. 26--29]{jia2012sparsegrid}.  The implementable convention in this
chapter is the direct translation of that band:
\[
  \mathcal A_{b,L}
  =
  \{\mathbf i\in\mathbb N^b:
    L\le |\mathbf i|\le L+b-1\},
  \qquad |\mathbf i|=\sum_{j=1}^b i_j ,
\]
and coefficient
\begin{equation}
  c_{\mathbf i}
  =
  (-1)^{L+b-1-|\mathbf i|}
  {b-1\choose L+b-1-|\mathbf i|}.
  \label{eq:bf-hd-sgqf-coefficient}
\end{equation}
The sparse-grid approximation is
\begin{equation}
  A_{b,L}[F]
  =
  \sum_{\mathbf i\in\mathcal A_{b,L}} c_{\mathbf i} I_{\mathbf i}[F].
  \label{eq:bf-hd-sgqf-smolyak}
\end{equation}
To see the equivalence with the source formula, write their inner set as
\(N_q^b=\{\mathbf i:|\mathbf i|=b+q\}\) and sum over
\(q=L-b,\ldots,L-1\).  Then \(|\mathbf i|=b+q\) runs from \(L\) to \(L+b-1\),
and their coefficient becomes exactly \eqref{eq:bf-hd-sgqf-coefficient}.  The
paper's Theorem 3.1 records the inherited polynomial exactness, Algorithm 1
gives the construction, and Theorem 3.2 relates UKF to a low-level SGQF member
in the paper's parameterization
\citep[Theorem 3.1, Algorithm 1, Theorem 3.2]{jia2012sparsegrid}.

\paragraph{Combined node and weight dictionary.}
Equation \eqref{eq:bf-hd-sgqf-smolyak} is not yet an implementable point cloud,
because the same standardized node can appear in several tensor components.
The implementation constructs a dictionary \(D\) keyed by standardized node
\(\xi\):
\[
  D[\xi]\leftarrow D[\xi]+c_{\mathbf i}w^{\mathbf i,\alpha}
  \quad\hbox{for every } \mathbf i\in\mathcal A_{b,L}
  \hbox{ and tensor node } \alpha .
\]
After duplicate-node merging, discard entries with numerical zero weight and
enumerate the remaining dictionary as
\[
  \{(\xi^{(r)},w_r)\}_{r=1}^{M_{b,L}} .
\]
The weights can be negative.  Therefore covariance matrices assembled from the
rule must be checked, symmetrized, and, if necessary, rejected or stabilized by
a declared square-root/PSD policy.  Silent clipping changes the scalar and
invalidates the HMC gradient unless the clipping is part of the scalar
definition.

\begin{algorithm}
\caption{ConstructFixedSparseGrid}
\label{alg:bf-hd-construct-fixed-sgq}
\begin{algorithmic}[1]
\Require Dimension \(b\), level \(L\), one-dimensional rules \(I_\ell\) for
\(\ell=1,\ldots,L+b-1\) built by \eqref{eq:bf-hd-sgqf-level-policy} unless a
different declared family is supplied
\Ensure Distinct standardized points and signed weights
\State \(D\gets\) empty map from standardized node to weight
\ForAll{\(\mathbf i\in\mathbb N^b\) with \(L\le|\mathbf i|\le L+b-1\)}
  \State \(c_{\mathbf i}\gets
  (-1)^{L+b-1-|\mathbf i|}{b-1\choose L+b-1-|\mathbf i|}\)
  \ForAll{tensor nodes
  \(\alpha\in\{1,\ldots,s_{i_1}\}\times\cdots\times\{1,\ldots,s_{i_b}\}\)}
    \State \(\xi\gets(\zeta_{i_1,\alpha_1},\ldots,\zeta_{i_b,\alpha_b})\)
    \State \(w\gets c_{\mathbf i}\prod_{j=1}^b\omega_{i_j,\alpha_j}\)
    \State Add \(w\) to \(D[\xi]\)
  \EndFor
\EndFor
\State Return all \((\xi,w)\) in \(D\) whose accumulated weight is not zero
\end{algorithmic}
\end{algorithm}

\paragraph{Filtering recursion using a fixed sparse grid.}
After Algorithm~\ref{alg:bf-hd-construct-fixed-sgq} returns
\(\{(\xi^{(r)},w_r)\}_{r=1}^{M}\), the filter uses exactly the same moment
formulas as GHQF, with \(\sum_\alpha w^\alpha\) replaced by \(\sum_{r=1}^M w_r\).
The prediction step is
\begin{align}
  \chi_{t-1}^{(r)}&=m_{t-1}+C_{t-1}\xi^{(r)},&
  x_{t|t-1}^{(r)}&=f_\theta(\chi_{t-1}^{(r)}),\\
  m_t^-&=\sum_{r=1}^Mw_rx_{t|t-1}^{(r)},&
  P_t^-&=Q_\theta+
  \sum_{r=1}^Mw_r
  (x_{t|t-1}^{(r)}-m_t^-)(x_{t|t-1}^{(r)}-m_t^-)^\top .
  \label{eq:bf-hd-sgqf-predict}
\end{align}
The observation step is
\begin{align}
  \chi_t^{(r)}&=m_t^-+C_t^-\xi^{(r)},&
  z_t^{(r)}&=h_\theta(\chi_t^{(r)}),\\
  \bar z_t&=\sum_{r=1}^Mw_rz_t^{(r)},\\
  S_t&=R_\theta+\sum_{r=1}^Mw_r
  (z_t^{(r)}-\bar z_t)(z_t^{(r)}-\bar z_t)^\top,\\
  C_{xz,t}&=\sum_{r=1}^Mw_r
  (\chi_t^{(r)}-m_t^-)(z_t^{(r)}-\bar z_t)^\top .
  \label{eq:bf-hd-sgqf-observation}
\end{align}
Then compute \(v_t,\widehat\ell_t,K_t,m_t,P_t\) as in
\eqref{eq:bf-hd-ghq-update-moments}--\eqref{eq:bf-hd-ghq-update}.

\begin{algorithm}
\caption{RunFixedSparseGridFilterStep}
\label{alg:bf-hd-fixed-sgqf-step}
\begin{algorithmic}[1]
\Require \(m_{t-1},P_{t-1},y_t,\theta\); fixed sparse-grid cloud
\(\{(\xi^{(r)},w_r)\}_{r=1}^M\); maps and covariances
\Ensure \(m_t,P_t,\widehat\ell_t\), score inputs, diagnostics
\State Factor \(P_{t-1}=C_{t-1}C_{t-1}^\top\)
\State Use \eqref{eq:bf-hd-sgqf-predict} to compute \(m_t^-,P_t^-\)
\State Symmetrize \(P_t^-\); if the square-root diagnostic fails, return veto
\State Factor \(P_t^-=C_t^-(C_t^-)^\top\)
\State Use \eqref{eq:bf-hd-sgqf-observation} to compute
\(\bar z_t,S_t,C_{xz,t}\)
\State Symmetrize \(S_t\); if \(S_t\) is not positive definite, return veto
\State Compute \(v_t=y_t-\bar z_t\) and \(\widehat\ell_t\)
\State Compute \(K_t=C_{xz,t}S_t^{-1}\), update \(m_t,P_t\), and run the PSD
diagnostic
\end{algorithmic}
\end{algorithm}

\paragraph{Why this is the high-dimensional candidate.}
For fixed level, Jia--Xin--Cheng record polynomial point growth rather than the
full tensor-product \(s^b\) growth \citep[Proposition 3.1 and Appendix
point-count examples]{jia2012sparsegrid}.  The point count can still be too
large, and signed weights can damage covariance semantics, but the fixed sparse
grid is the smallest object in this chapter that can plausibly define a
high-dimensional approximate likelihood scalar with a reproducible analytical
gradient.  It can be used as an HMC likelihood only in that precise sense: the
sampler sees the fixed sparse-grid approximate scalar, not the exact
filtering likelihood.  The index set \(\mathcal A_{b,L}\), one-dimensional
rules, duplicate-node merging, covariance stabilization policy, and
square-root branches must be fixed parts of the scalar being differentiated.

\section{Adaptive Sparse-Grid Gauss-Hermite Filter (ASGHF)}
\label{sec:bf-hd-asghf}

\paragraph{Problem setting.}
ASGHF is not a different filtering object.  It is a procedure for choosing a
sparse-grid index set before using the same Gaussian projection formulas.
Singh et al. start from sparse-grid Gauss--Hermite filtering and adaptively add
tensor-product increments in directions whose estimated contribution is large
relative to their cost \citep[Section 3]{singh2018adaptive}.  The implementable
BayesFilter interpretation is:

\[
  \hbox{ASGHF pilot run chooses an index set}
  \quad\Longrightarrow\quad
  \hbox{freeze that index set}
  \quad\Longrightarrow\quad
  \hbox{run fixed SGQF.}
\]

\paragraph{Difference increments.}
Let \(I_\ell\) be a one-dimensional Gauss--Hermite rule and set \(I_0=0\).
Define
\[
  \Delta_\ell=I_\ell-I_{\ell-1}.
\]
For a multi-index \(\lambda=(\lambda_1,\ldots,\lambda_b)\),
\[
  \Delta_\lambda F
  =
  (\Delta_{\lambda_1}\otimes\cdots\otimes\Delta_{\lambda_b})F .
\]
This increment is the new tensor-product information added by level
\(\lambda\) after lower one-dimensional levels have already been accounted for.

\paragraph{Admissible index sets.}
An index set \(\Lambda\subset\mathbb N^b\) is admissible if every backward
neighbor is present:
\[
  \lambda\in\Lambda,\ \lambda_j>1
  \quad\Longrightarrow\quad
  \lambda-e_j\in\Lambda .
\]
The algorithm maintains an old set \(O\) of accepted indices and an active
frontier \(A\) of admissible candidates.  A forward neighbor \(\lambda+e_j\)
can be added to \(A\) only when all of its backward neighbors are already in
\(O\cup A\).

\paragraph{Error and cost indicator.}
For a vector-valued filter integrand \(F\), Singh et al. use a local indicator
of the form
\begin{equation}
  g_\lambda
  =
  \max\left\{
    \psi\frac{\|\Delta_\lambda F\|_1}{\|\Delta_{\mathbf 1}F\|_1},
    (1-\psi)\frac{\varpi_{\mathbf 1}}{\varpi_\lambda}
  \right\},
  \label{eq:bf-hd-asghf-error-indicator}
\end{equation}
where \(\mathbf 1=(1,\ldots,1)\), \(\varpi_\lambda\) is the number of function
evaluations in the increment, and \(\psi\in[0,1]\) trades estimated error
against cost \citep[Section 3]{singh2018adaptive}.  For filtering, \(F\) should
be a concrete vector function of a standardized point.  BayesFilter uses a
pilot Gaussian \(X=m+C\xi\), a pilot predicted Gaussian
\(X^-=m^-+C^-\xi\), and the column-major vectorization
\[
  \operatorname{vec}(A)=(A_{11},A_{21},\ldots,A_{n1},A_{12},\ldots)^\top .
\]
The pilot integrand is
\begin{equation}
  F_{\rm pilot}(\xi)
  =
  \begin{bmatrix}
    a(\xi)/s_a\\
    \operatorname{vec}(a(\xi)a(\xi)^\top)/s_{aa}\\
    z(\xi)/s_z\\
    \operatorname{vec}(z(\xi)z(\xi)^\top)/s_{zz}\\
    \operatorname{vec}((C^-\xi)z(\xi)^\top)/s_{xz}
  \end{bmatrix},
  \quad
  a(\xi)=f_\theta(m+C\xi),\quad
  z(\xi)=h_\theta(m^-+C^-\xi).
  \label{eq:bf-hd-asghf-pilot-integrand}
\end{equation}
The positive scales \(s_a,s_{aa},s_z,s_{zz},s_{xz}\) are fixed before the
adaptive run, for example the corresponding norms under the base index
\(\mathbf 1\) floored at one.  The norm in
\eqref{eq:bf-hd-asghf-error-indicator} is then the ordinary vector
\(\ell_1\) norm of this concatenated vector.  This convention makes the
adaptive score reproducible; changing the vectorization, scales, or norm is a
different grid-selection algorithm.  Using only one scalar diagnostic can miss
an interaction that matters for covariance or likelihood.

\begin{algorithm}
\caption{SelectAdaptiveSparseGrid}
\label{alg:bf-hd-select-asghf}
\begin{algorithmic}[1]
\Require Dimension \(b\), one-dimensional rules, pilot Gaussian
\((m,P)\), pilot integrand vector \(F\), tolerance \(TOL\), maximum point budget
\(M_{\max}\), tradeoff \(\psi\)
\Ensure Frozen admissible index set \(\Lambda\), or a budget/tolerance veto
\State \(O\gets\emptyset\), \(A\gets\{\mathbf 1\}\), \(E\gets\infty\)
\While{\(E>TOL\) and estimated point count is below \(M_{\max}\)}
  \State Choose \(\lambda^\star\in A\) with largest \(g_\lambda\)
  \State Move \(\lambda^\star\) from \(A\) to \(O\)
  \For{\(j=1,\ldots,b\)}
    \State \(\mu\gets\lambda^\star+e_j\)
    \If{\(\mu\) is admissible relative to \(O\cup A\)}
      \State Evaluate \(\Delta_\mu F\), compute \(g_\mu\), and add \(\mu\) to \(A\)
    \EndIf
  \EndFor
  \State Update \(E\) from the active indicators, for example
  \(E=\sum_{\lambda\in A}g_\lambda\)
\EndWhile
\State If \(E>TOL\), return veto; otherwise return \(\Lambda=O\cup A\)
\end{algorithmic}
\end{algorithm}

\paragraph{Node and weight assembly after adaptation.}
The adaptive algorithm returns an admissible set \(\Lambda\), not a point cloud.
To freeze it, expand every tensor increment
\[
  \Delta_\lambda
  =
  \bigotimes_{j=1}^b (I_{\lambda_j}-I_{\lambda_j-1})
  =
  \sum_{\epsilon\in\{0,1\}^b}
  (-1)^{|\epsilon|}
  \bigotimes_{j=1}^b I_{\lambda_j-\epsilon_j},
  \label{eq:bf-hd-asghf-increment-expansion}
\]
with \(I_0=0\).  Terms with any \(\lambda_j-\epsilon_j=0\) are skipped.  Then
accumulate the tensor-product nodes and signed weights in the same dictionary
style as fixed SGQF.

\begin{algorithm}
\caption{AssembleFrozenAdaptiveSparseGrid}
\label{alg:bf-hd-assemble-frozen-asghf}
\begin{algorithmic}[1]
\Require Frozen admissible set \(\Lambda\), one-dimensional rules \(I_\ell\)
\Ensure Distinct standardized points and signed weights
\State \(D\gets\) empty map from standardized node to weight
\ForAll{\(\lambda\in\Lambda\)}
  \ForAll{\(\epsilon\in\{0,1\}^b\)}
    \State \(j_\ell\gets\lambda_\ell-\epsilon_\ell\) for
    \(\ell=1,\ldots,b\)
    \If{any \(j_\ell=0\)}
      \State Skip this term
    \Else
      \State \(c\gets(-1)^{\epsilon_1+\cdots+\epsilon_b}\)
      \ForAll{tensor nodes
      \(\alpha\in\{1,\ldots,s_{j_1}\}\times\cdots\times\{1,\ldots,s_{j_b}\}\)}
        \State \(\xi\gets(\zeta_{j_1,\alpha_1},\ldots,\zeta_{j_b,\alpha_b})\)
        \State \(w\gets c\prod_{\ell=1}^b\omega_{j_\ell,\alpha_\ell}\)
        \State Add \(w\) to \(D[\xi]\)
      \EndFor
    \EndIf
  \EndFor
\EndFor
\State Return all \((\xi,w)\) in \(D\) whose accumulated weight is not zero
\end{algorithmic}
\end{algorithm}

\paragraph{Frozen-grid filtering after adaptation.}
Algorithm~\ref{alg:bf-hd-select-asghf} is a grid-selection routine, not an HMC
target by itself.  To use it in a likelihood, convert the returned admissible
set \(\Lambda\) into a signed node/weight dictionary with
Algorithm~\ref{alg:bf-hd-assemble-frozen-asghf}.  Then freeze the resulting
distinct \(\{(\xi^{(r)},w_r)\}\) and run
Algorithm~\ref{alg:bf-hd-fixed-sgqf-step}.  The frozen-grid scalar is
differentiable under the same assumptions as fixed SGQF.  If
Algorithm~\ref{alg:bf-hd-select-asghf} is rerun as \(\theta\) changes during
HMC, the scalar changes by discrete branch decisions.  The sampler would then
receive gradients from one frozen cloud while evaluating values from another,
or would differentiate through a discontinuous grid-selection rule.  Therefore
ASGHF is a grid-design procedure before HMC, not a live adaptive rule inside an
HMC trajectory.

\paragraph{Limitations and diagnostics.}
ASGHF can miss weak many-way interactions because an increment may look
unimportant before its neighboring increments have been selected.  It can also
select a grid that is adequate for a pilot Gaussian but poor after filtering
moves the covariance.  Required diagnostics are: final point count, active error
estimate, PSD status of the resulting \(P_t^-\) and \(S_t\), duplicate-node
weight cancellation, and finite-difference parity after freezing the grid.
After the grid is frozen, the likelihood and gradient are those of fixed SGQF.

\section{Approximate Likelihood and Analytical Gradient}
\label{sec:bf-hd-gq-gradient-contract}

\paragraph{Problem setting.}
HMC cannot sample from a method name.  It needs a scalar potential and the
gradient of that same scalar.  For the Gaussian and quadrature filters in this
chapter, the scalar is the reported Gaussian innovation log likelihood.

For a fixed deterministic rule, define
\begin{align}
  \chi_t^{(r)}(\theta)
  &=m_t^-(\theta)+C_t^-(\theta)\xi^{(r)},\\
  z_t^{(r)}(\theta)
  &=h_\theta(\chi_t^{(r)}(\theta)).
  \label{eq:bf-hd-gq-cloud}
\end{align}
The approximate observation moments are
\begin{align}
  \bar z_t&=\sum_r w_r^{(m)}z_t^{(r)},\\
  S_t&=R_\theta+\sum_r w_r^{(c)}
    (z_t^{(r)}-\bar z_t)(z_t^{(r)}-\bar z_t)^\top,\\
  v_t&=y_t-\bar z_t.
  \label{eq:bf-hd-gq-likelihood-objects}
\end{align}
The approximate scalar is
\begin{equation}
  \widehat\ell_t(\theta)
  =
  -\frac12
  \{\log\det S_t+v_t^\top S_t^{-1}v_t+n_y\log(2\pi)\}.
  \label{eq:bf-hd-gq-loglik}
\end{equation}
This is a project derivation in BayesFilter notation, patterned after the
sigma-point likelihood derivative in Chapter~\ref{ch:bf-svd-sigma-point}.  It
is not a claim that the cited UKF, CKF, SGQF, or ASGHF papers prove HMC
correctness.

\paragraph{Point and moment derivatives.}
For parameter component \(\theta_i\), assume the offsets and weights are fixed
and the covariance factor branch is differentiable.  Equations
\eqref{eq:bf-hd-gq-point-score}--\eqref{eq:bf-hd-gq-score} are valid only when
\(w_r^{(m)},w_r^{(c)}\), and \(\xi^{(r)}\) are constant in the differentiated
parameter.  If a tuning parameter changes the weights, offsets, index set, or
degree, the corresponding derivative terms must be added to the scalar
definition and score.  Then
\begin{align}
  \dot\chi_t^{(r,i)}
  &=
  \dot m_t^{-(i)}+\dot C_t^{-(i)}\xi^{(r)},\\
  \dot z_t^{(r,i)}
  &=
  D_xh_\theta(\chi_t^{(r)})\dot\chi_t^{(r,i)}
  +\partial_i h_\theta(\chi_t^{(r)}),\\
  \dot{\bar z}_t^{(i)}
  &=
  \sum_r w_r^{(m)}\dot z_t^{(r,i)}.
  \label{eq:bf-hd-gq-point-score}
\end{align}
Let
\[
  \Delta_t^{(r)}=z_t^{(r)}-\bar z_t,\qquad
  \dot\Delta_t^{(r,i)}=\dot z_t^{(r,i)}-\dot{\bar z}_t^{(i)}.
\]
Then
\begin{align}
  \dot S_t^{(i)}
  &=
  \dot R_\theta^{(i)}
  +
  \sum_r w_r^{(c)}
  \left\{
    \dot\Delta_t^{(r,i)}\Delta_t^{(r)\top}
    +\Delta_t^{(r)}\dot\Delta_t^{(r,i)\top}
  \right\},\\
  \dot v_t^{(i)}
  &=
  \dot y_t^{(i)}-\dot{\bar z}_t^{(i)}.
  \label{eq:bf-hd-gq-S-v-score}
\end{align}
For recorded data independent of \(\theta\), \(\dot y_t^{(i)}=0\).  If
quadrature tuning parameters or weights are estimated, their derivatives must
be added explicitly; the fixed-rule formulas above do not cover hidden rule
adaptation.

\begin{proposition}[Approximate Gaussian innovation score]
\label{prop:bf-hd-gq-score}
Assume \(S_t\) is positive definite, the deterministic rule is fixed, and all
objects above are differentiable on the active branch.  If \(S_tw_t=v_t\), then
the derivative of \eqref{eq:bf-hd-gq-loglik} is
\begin{equation}
  \partial_i\widehat\ell_t
  =
  -\frac12
  \left[
    \tr(S_t^{-1}\dot S_t^{(i)})
    +2\dot v_t^{(i)\top}w_t
    -w_t^\top\dot S_t^{(i)}w_t
  \right].
  \label{eq:bf-hd-gq-score}
\end{equation}
\end{proposition}

\begin{proof}
Differentiate the three terms in \eqref{eq:bf-hd-gq-loglik}.  The log
determinant identity gives
\[
  \partial_i\log\det S_t=\tr(S_t^{-1}\dot S_t^{(i)}).
\]
For the quadratic term, use \(w_t=S_t^{-1}v_t\) and
\(\partial_i S_t^{-1}=-S_t^{-1}\dot S_t^{(i)}S_t^{-1}\):
\[
  \partial_i(v_t^\top S_t^{-1}v_t)
  =
  2\dot v_t^{(i)\top}w_t
  -w_t^\top\dot S_t^{(i)}w_t.
\]
The constant \(n_y\log(2\pi)\) has zero derivative.  Combining the derivatives
proves \eqref{eq:bf-hd-gq-score}.
\end{proof}

\subsection{Fixed-SGQF Approximate Likelihood Gradient}
\label{sec:bf-hd-fixed-sgqf-gradient}

The generic score above is useful but not enough for implementation.  The
selected high-dimensional target in this chapter is the fixed sparse-grid
Gaussian projection likelihood
\begin{equation}
  \widehat\ell_T^{\rm FSGQ}(\theta)
  =
  \sum_{t=1}^T\widehat\ell_t^{\rm FSGQ}(\theta),
  \label{eq:bf-hd-fsgq-total-likelihood}
\end{equation}
where every \(\widehat\ell_t^{\rm FSGQ}\) is produced by
Algorithm~\ref{alg:bf-hd-fixed-sgqf-step} using one frozen standardized cloud
\(\{(\xi^{(r)},w_r)\}_{r=1}^M\).  The cloud is part of the scalar definition.
The gradient below differentiates this approximate scalar, not the exact
nonlinear likelihood.

\paragraph{Differentiable square-root convention.}
At each time, let \(C(P)\) be the declared covariance square-root map used by
the implementation.  The default P9 implementation is unpivoted Cholesky on the
strictly positive-definite branch, \(P=LL^\top\), \(C=L\).  Given a symmetric
\(\dot P^{(i)}\), compute
\[
  A^{(i)}=L^{-1}\dot P^{(i)}L^{-\top}.
\]
Let \(\Phi(A)\) be the lower-triangular matrix
\[
  \Phi(A)_{jk}=
  \begin{cases}
    A_{jk}, & j>k,\\
    A_{jj}/2, & j=k,\\
    0, & j<k .
  \end{cases}
\]
Then the Cholesky sensitivity is
\begin{equation}
  \dot L^{(i)}=L\Phi(A^{(i)}).
  \label{eq:bf-hd-fsgq-cholesky-derivative}
\end{equation}
Indeed, with \(B=\Phi(A)\), \(B+B^\top=A\), so
\[
  \dot L L^\top+L\dot L^\top
  =
  L(B+B^\top)L^\top
  =
  \dot P .
\]
If the implementation uses SVD, a pivoted factor, or a custom square-root map
from the earlier derivative chapters, it must instead supply
\[
  \dot C^{(i)} = D C(P)[\dot P^{(i)}].
\]
If the square-root branch pivots, changes rank, floors eigenvalues, or clips
covariances, the derivative is only branch-local unless the branch rule is
included in the scalar and differentiated or smoothed.

\paragraph{Prediction sensitivities.}
Assume the previous filtering state and sensitivities are
\[
  m_{t-1},\quad P_{t-1},\quad
  \dot m_{t-1}^{(i)},\quad \dot P_{t-1}^{(i)} .
\]
Let \(C_{t-1}=C(P_{t-1})\) and
\(\dot C_{t-1}^{(i)}=DC(P_{t-1})[\dot P_{t-1}^{(i)}]\).  For each fixed
sparse-grid point,
\begin{align}
  \chi_{t-1}^{(r)}&=m_{t-1}+C_{t-1}\xi^{(r)},\\
  \dot\chi_{t-1}^{(r,i)}
  &=
  \dot m_{t-1}^{(i)}+\dot C_{t-1}^{(i)}\xi^{(r)}.
  \label{eq:bf-hd-fsgq-pred-point-score}
\end{align}
Propagate the transition points:
\begin{align}
  a_t^{(r)}&=f_\theta(\chi_{t-1}^{(r)}),\\
  \dot a_t^{(r,i)}
  &=
  D_xf_\theta(\chi_{t-1}^{(r)})\dot\chi_{t-1}^{(r,i)}
  +\partial_i f_\theta(\chi_{t-1}^{(r)}).
  \label{eq:bf-hd-fsgq-transition-score}
\end{align}
Then
\begin{align}
  m_t^-&=\sum_rw_ra_t^{(r)},&
  \dot m_t^{-(i)}&=\sum_rw_r\dot a_t^{(r,i)}.
  \label{eq:bf-hd-fsgq-pred-mean-score}
\end{align}
With \(d_t^{(r)}=a_t^{(r)}-m_t^-\) and
\(\dot d_t^{(r,i)}=\dot a_t^{(r,i)}-\dot m_t^{-(i)}\),
\begin{align}
  P_t^-&=Q_\theta+\sum_rw_rd_t^{(r)}d_t^{(r)\top},\\
  \dot P_t^{-(i)}
  &=
  \dot Q_\theta^{(i)}
  +
  \sum_rw_r
  \{\dot d_t^{(r,i)}d_t^{(r)\top}
  +d_t^{(r)}\dot d_t^{(r,i)\top}\}.
  \label{eq:bf-hd-fsgq-pred-cov-score}
\end{align}

\paragraph{Observation and likelihood sensitivities.}
Let \(C_t^-=C(P_t^-)\) and
\(\dot C_t^{-(i)}=DC(P_t^-)[\dot P_t^{-(i)}]\).  The observation cloud and its
sensitivities are
\begin{align}
  \chi_t^{(r)}&=m_t^-+C_t^-\xi^{(r)},\\
  \dot\chi_t^{(r,i)}
  &=
  \dot m_t^{-(i)}+\dot C_t^{-(i)}\xi^{(r)},\\
  z_t^{(r)}&=h_\theta(\chi_t^{(r)}),\\
  \dot z_t^{(r,i)}
  &=
  D_xh_\theta(\chi_t^{(r)})\dot\chi_t^{(r,i)}
  +\partial_i h_\theta(\chi_t^{(r)}).
  \label{eq:bf-hd-fsgq-obs-point-score}
\end{align}
The observation mean and innovation residual derivatives are
\begin{align}
  \bar z_t&=\sum_rw_rz_t^{(r)},&
  \dot{\bar z}_t^{(i)}&=\sum_rw_r\dot z_t^{(r,i)},\\
  v_t&=y_t-\bar z_t,&
  \dot v_t^{(i)}&=-\dot{\bar z}_t^{(i)}
  \label{eq:bf-hd-fsgq-z-v-score}
\end{align}
for recorded data independent of \(\theta\).  With
\(\Delta_t^{(r)}=z_t^{(r)}-\bar z_t\) and
\(\dot\Delta_t^{(r,i)}=\dot z_t^{(r,i)}-\dot{\bar z}_t^{(i)}\),
\begin{align}
  S_t&=R_\theta+\sum_rw_r\Delta_t^{(r)}\Delta_t^{(r)\top},\\
  \dot S_t^{(i)}
  &=
  \dot R_\theta^{(i)}
  +
  \sum_rw_r
  \{\dot\Delta_t^{(r,i)}\Delta_t^{(r)\top}
  +\Delta_t^{(r)}\dot\Delta_t^{(r,i)\top}\}.
  \label{eq:bf-hd-fsgq-S-score}
\end{align}
The cross-covariance is needed for the filter state:
\begin{align}
  C_{xz,t}&=\sum_rw_r
  (\chi_t^{(r)}-m_t^-)\Delta_t^{(r)\top},\\
  \dot C_{xz,t}^{(i)}
  &=
  \sum_rw_r
  \{(\dot\chi_t^{(r,i)}-\dot m_t^{-(i)})\Delta_t^{(r)\top}
  +(\chi_t^{(r)}-m_t^-)\dot\Delta_t^{(r,i)\top}\}.
  \label{eq:bf-hd-fsgq-cross-score}
\end{align}
For the current likelihood contribution, \(\dot C_{xz,t}^{(i)}\) is not needed.
It is needed only if the implementation propagates posterior sensitivities
through the Kalman update to the next time step.

\paragraph{Likelihood score and posterior sensitivity propagation.}
If \(S_tw_t=v_t\), the per-period score is
\begin{equation}
  \partial_i\widehat\ell_t^{\rm FSGQ}
  =
  -\frac12
  \left[
    \tr(S_t^{-1}\dot S_t^{(i)})
    +2\dot v_t^{(i)\top}w_t
    -w_t^\top\dot S_t^{(i)}w_t
  \right].
  \label{eq:bf-hd-fsgq-score}
\end{equation}
To continue the recursion, differentiate the update on the same smooth branch.
Let \(K_t=C_{xz,t}S_t^{-1}\).  Then
\begin{align}
  \dot K_t^{(i)}
  &=
  \dot C_{xz,t}^{(i)}S_t^{-1}
  -K_t\dot S_t^{(i)}S_t^{-1},\\
  \dot m_t^{(i)}
  &=
  \dot m_t^{-(i)}
  +\dot K_t^{(i)}v_t
  +K_t\dot v_t^{(i)},\\
  \dot P_t^{(i)}
  &=
  \dot P_t^{-(i)}
  -\dot K_t^{(i)}S_tK_t^\top
  -K_t\dot S_t^{(i)}K_t^\top
  -K_tS_t\dot K_t^{(i)\top}.
  \label{eq:bf-hd-fsgq-update-score}
\end{align}
Equations \eqref{eq:bf-hd-fsgq-pred-point-score}--\eqref{eq:bf-hd-fsgq-update-score}
are enough to implement a value-and-gradient pass for
\(\widehat\ell_T^{\rm FSGQ}\), provided the same fixed sparse-grid cloud,
factor branch, covariance stabilization policy, and model derivative routines
are used in both the value and gradient paths.

\begin{algorithm}
\caption{FixedSparseGridValueAndScore}
\label{alg:bf-hd-fsgq-score}
\begin{algorithmic}[1]
\Require Observations \(y_{1:T}\), parameters \(\theta\), initial
\(m_0,P_0,\dot m_0^{(i)},\dot P_0^{(i)}\), fixed cloud
\(\{(\xi^{(r)},w_r)\}\), model derivatives
\Ensure \(\widehat\ell_T^{\rm FSGQ}\) and \(\nabla_\theta\widehat\ell_T^{\rm FSGQ}\)
\State \(\widehat\ell\gets0\), score components \(g_i\gets0\)
\For{\(t=1,\ldots,T\)}
  \State Compute prediction value and sensitivities by
  \eqref{eq:bf-hd-fsgq-pred-point-score}--\eqref{eq:bf-hd-fsgq-pred-cov-score}
  \State Compute observation value and sensitivities by
  \eqref{eq:bf-hd-fsgq-obs-point-score}--\eqref{eq:bf-hd-fsgq-S-score}
  \State If \(P_t^-\) or \(S_t\) fails the declared PSD/PD diagnostic, return veto
  \State Add \(\widehat\ell_t^{\rm FSGQ}\) to \(\widehat\ell\)
  \State For each \(i\), add \eqref{eq:bf-hd-fsgq-score} to \(g_i\)
  \State Compute posterior sensitivities by \eqref{eq:bf-hd-fsgq-cross-score}
  and \eqref{eq:bf-hd-fsgq-update-score}
\EndFor
\end{algorithmic}
\end{algorithm}

\paragraph{When these approximations can feed HMC.}
The question for HMC is not whether a method name sounds differentiable.  The
question is whether the implementation reports a scalar
\(\widehat\ell_T(\theta)\) and the derivative of that same scalar.  The table
below states this in ordinary mathematical terms.  None of the entries is a
convergence guarantee; they only say what object an HMC target would be
differentiating.

\begin{longtable}{@{}p{0.25\linewidth}p{0.30\linewidth}p{0.34\linewidth}@{}}
\toprule
Method family & Scalar seen by HMC & What must be fixed or checked \\
\midrule
\endhead
Extended Kalman filter under declared approximate scalar \(\widehat\ell_t\) &
The EKF Gaussian innovation approximation. &
The linearization rule, covariance factor branch, and derivative path must be
the same in value and gradient calculations. \\
Iterated extended Kalman filter under declared approximate scalar
\(\widehat\ell_t\) &
No single scalar unless the iteration branch is frozen or smoothed. &
Iteration count and convergence basin can change with \(\theta\); without a
declared branch, the derivative is not the derivative of one smooth function. \\
UKF, CKF, high-degree CKF, fixed tensor-product GHQ &
The fixed-rule Gaussian innovation scalar
\eqref{eq:bf-hd-gq-loglik}. &
Offsets, weights, covariance factors, and stabilization rules must be fixed
smooth parts of the scalar.  Large tensor-product GHQ usually fails the cost
test before it becomes a serious HMC candidate. \\
Fixed nonadaptive SGQF &
The fixed sparse-grid scalar
\(\widehat\ell_T^{\rm FSGQ}\). &
The index set, nodes, weights, duplicate-node merge, covariance policy, and
factor derivative must be fixed and used identically in value and gradient. \\
Frozen ASGHF branch &
The fixed SGQF scalar built from the selected grid. &
The adaptive search must run before HMC.  Once the grid is frozen, the filter
is mathematically the fixed SGQF above. \\
Live adaptive ASGHF &
No smooth scalar is supplied by the algorithm as written. &
The point set changes by discrete decisions.  HMC should not be handed this
branch-changing procedure as if it were one differentiable likelihood. \\
Large tensor-product GHQ &
A clear scalar in principle. &
The exponential point count makes it a small-block reference and debugging
tool, not the high-dimensional route. \\
\bottomrule
\end{longtable}

\paragraph{Implementability checklist.}
The table is a post-derivation checklist, not the explanation.

\begin{longtable}{@{}p{0.22\linewidth}p{0.18\linewidth}p{0.18\linewidth}p{0.18\linewidth}p{0.16\linewidth}@{}}
\toprule
Object & Nodes/weights & Filter recursion & Gradient recursion & Use in inference \\
\midrule
\endhead
Tensor-product GHQF & Eq. \eqref{eq:bf-hd-ghq-tensor-rule} &
Alg. \ref{alg:bf-hd-ghqf-step} & Generic fixed-rule score only &
Small-block reference because \(s^b\) points do not scale \\
Fixed SGQF & Alg. \ref{alg:bf-hd-construct-fixed-sgq} &
Alg. \ref{alg:bf-hd-fixed-sgqf-step} &
Alg. \ref{alg:bf-hd-fsgq-score} &
Selected fixed approximate likelihood target \\
ASGHF & Alg. \ref{alg:bf-hd-select-asghf} then freeze &
Frozen SGQF recursion & Fixed-SGQF gradient after freeze &
Grid design before inference; fixed SGQF after freeze \\
\bottomrule
\end{longtable}

\section{Limitations of These Methods}
\label{sec:bf-hd-gq-defects}

The limitations are not bureaucratic cautions; they are mathematical reasons
the methods can fail.

\begin{enumerate}[label=(\arabic*)]
  \item Gaussian projection preserves a mean/covariance object, not the full
  posterior law.  The scalar quadratic cell can be bimodal while every method
  here stores one Gaussian.
  \item EKF and second-order Taylor filters depend on local derivatives.  They
  fail when the expansion point is not representative of the posterior mass or
  when Hessian contractions are too dense to compute reliably.
  \item The Julier--Uhlmann UT matches the input mean and covariance by
  construction, but negative or badly scaled weights can damage positive
  semidefinite covariance semantics.
  \item CKF is a clean third-degree Gaussian integral rule.  It misses fourth
  moments such as the variance of \(X^2\) in the scalar cell.
  \item Jia--Xin--Cheng high-degree CKF improves polynomial moment exactness,
  but fifth-degree and higher rules add point cost and can introduce negative
  weights.  The broader Stroud/Genz/Cools cubature foundations are not used as
  independent support in this chapter.
  \item Tensor-product Gauss--Hermite filtering is easy to understand and hard
  to scale: \(s^b\) points is a hard veto outside tiny blocks.  Its main role
  here is reference construction and small-block parity testing.
  \item Jia--Xin--Cheng SGQF reduces tensor-product cost for fixed level, but it
  omits interactions outside the selected sparse-grid structure and can produce
  signed weights.  The PSD status of \(P_t^-\) and \(S_t\) is therefore an
  algorithmic veto, not an implementation detail.
  \item Singh et al.'s adaptive sparse-grid filter is useful as a computational
  allocation device, but it must be separated into grid selection and
  frozen-grid filtering before HMC can use the resulting approximate scalar.
  \item The approximate likelihood gradient in
  \eqref{eq:bf-hd-fsgq-score} is the derivative of the fixed sparse-grid
  approximate Gaussian innovation scalar.  It is not a derivative through live
  adaptation, hidden covariance clipping, or a changed node set.
\end{enumerate}

\paragraph{How synthesis uses this chapter.}
The synthesis decision is a mathematical filtering calculation, not a label
lookup.  Start with a block \(B\) of dimension \(b\) and a Gaussian prediction
\(\calN(m_B^-,P_B^-)\).  Choose a deterministic rule \(Q_B\) with standardized
points and weights \(\{(\xi^{(r)},w_r)\}_{r=1}^{M_B}\).  That rule is allowed
to enter the candidate architecture only if it can compute the five Gaussian
projection objects
\[
  m_B^-,\qquad
  P_B^-,\qquad
  \bar z_B,\qquad
  S_B,\qquad
  C_{xz,B}
\]
by explicit weighted sums of the form derived above, with \(P_B^-\succeq0\)
and \(S_B\succ0\) after the declared stabilization policy.  The local
likelihood contribution is then
\[
  \widehat\ell_{B,t}
  =
  -\frac12\{\log\det S_B
  +(y_{B,t}-\bar z_B)^\top S_B^{-1}(y_{B,t}-\bar z_B)
  +n_B\log(2\pi)\}.
\]
For a collection of blocks \(\mathcal B_t\), the candidate scalar used
downstream is the declared sum
\[
  \widehat\ell_t^{\rm cand}(\theta)
  =
  \sum_{B\in\mathcal B_t}\widehat\ell_{B,t}(\theta)
  +\widehat\ell_{t,\rm cross}(\theta),
\]
where the cross term is either zero by an explicit conditional-independence or
block-diagonal approximation, or is supplied by a stated low-rank/correction
model.  This equation is where synthesis must be honest: ignoring cross-block
dependence is an approximation to the likelihood, not a computational detail.

The promotion rule is therefore concrete.  Use tensor-product GHQ only when
\(M_B=s^b\) is small enough for reference evaluation.  Use CKF or high-degree
CKF when the active block is small and the needed moments are low-order smooth
polynomial-like effects.  Use fixed SGQF when \(M_{b,L}\) fits the point
budget, signed weights do not break the covariance diagnostics, and the
fixed-grid score recursion in
\eqref{eq:bf-hd-fsgq-pred-point-score}--\eqref{eq:bf-hd-fsgq-update-score}
is the derivative actually supplied to the inference engine.  Use ASGHF only to
select a grid before this calculation; after selection, replace it by the
frozen fixed-grid rule and re-run the same value-gradient checks.
