\chapter[Particle, Transport, And Tensor Filters]{Particle, Transport, Tensor-Train, and Tensor-Network Filters}
\label{ch:bf-highdim-particle-transport-tensor}

\section{Three Ways To Escape A Gaussian, And How Each Fails}
\label{sec:bf-hd-ptt-purpose}

Particle filters, transport maps, and tensor decompositions are three different
answers to the same discomfort: a single Gaussian can be the wrong object.
Particles represent a probability law by weighted samples.  Transports move a
reference law, forecast ensemble, or proposal toward posterior mass.  Tensor
trains and tensor networks compress density, operator, covariance, or map
objects.  These are not interchangeable objects, and a mixed numerical panel
should not be asked to infer the difference from acronyms.

The exact target remains the filtering law and normalizer from
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}.  The approximation is the
finite empirical measure, fitted map/coupling, or low-rank representation.  The
industrial role is conditional: these methods are useful only when correction,
support, rank, mass, positivity, covariance, likelihood, and score diagnostics
are visible.

\paragraph{Running quadratic-observation cell.}
Use again
\[
  x_t=\rho x_{t-1}+\eta_t,\qquad y_t=x_t^2+\epsilon_t .
\]
A particle filter tries to place weighted samples near both posterior lobes.
A transport tries to move forecast or reference mass toward those lobes while
preserving a density ratio, Jacobian, or source-supported deterministic
analysis contract.  A tensor representation tries to store the density,
operator, covariance, or transport map compactly.  The same scalar cell asks
three distinct questions:
\[
  \text{Did samples reach both lobes?}\quad
  \text{Did the map preserve correction and support?}\quad
  \text{Did compression preserve probability or covariance semantics?}
\]

\begin{longtable}{@{}p{0.18\linewidth}p{0.33\linewidth}p{0.35\linewidth}@{}}
\toprule
Representation & Object & First semantic veto \\
\midrule
\endhead
Empirical particle measure & \(\sum_i\bar w_i\delta_{x_i}\) & ESS, maximum
weight, unique ancestors, and coverage of relevant modes. \\
Proposal density & sampled law \(q(x)\) with computable density & target
support not covered or density-ratio correction not finite. \\
Transport map & map/coupling from reference or forecast law & invalid support,
inverse, monotonicity, Jacobian, or deterministic analysis contract. \\
TT density/operator & compressed grid/function/operator cores & mass,
positivity, boundary residual, rank spike. \\
TN covariance/factor & compressed covariance or square-root factor & PSD or
factor reconstruction failure. \\
\bottomrule
\end{longtable}

\section{Empirical Measures And Importance Weights}
\label{sec:bf-hd-pf}

The bootstrap filter of \citet{gordon1993novel} and the SIR mechanics
summarized by \citet{arulampalam2002tutorial} approximate the filtering law by
the empirical measure
\begin{equation}
  \widehat\pi_t^N(dx)=\sum_{i=1}^N \bar w_t^{(i)}\delta_{x_t^{(i)}}(dx),
  \qquad
  \bar w_t^{(i)}=\frac{w_t^{(i)}}{\sum_{j=1}^N w_t^{(j)}}.
  \label{eq:bf-hd-pf-measure}
\end{equation}
This is a probability measure: it assigns mass \(\bar w_t^{(i)}\) to each
simulated point \(x_t^{(i)}\).  It is not a density, not a covariance matrix,
and not a map.

Suppose a proposal samples
\[
  x_t^{(i)}\sim q_t(\cdot\mid x_{t-1}^{a(i)},y_t).
\]
Then the incremental importance weight is
\begin{equation}
  \alpha_t^{(i)}
  =
  \frac{
  g_\theta(y_t\mid x_t^{(i)})
  f_\theta(x_t^{(i)}\mid x_{t-1}^{a(i)})
  }{
  q_t(x_t^{(i)}\mid x_{t-1}^{a(i)},y_t)
  }.
  \label{eq:bf-hd-pf-weight}
\end{equation}
The bootstrap proposal sets \(q_t=f_\theta\), so
\(\alpha_t^{(i)}=g_\theta(y_t\mid x_t^{(i)})\).  A guided proposal changes
where particles are placed, but \eqref{eq:bf-hd-pf-weight} is the correction
that keeps the target interpretation.

\begin{algorithm}
\caption{Sequential importance resampling filter}
\label{alg:bf-hd-sir-filter}
\begin{algorithmic}[1]
\Require Particle count \(N\), transition \(f_\theta\), likelihood
  \(g_\theta\), proposal \(q_t\), resampling threshold \(\tau\).
\State Initialize particles and weights.
\For{\(t=1,\ldots,T\)}
  \State Resample ancestors if the declared resampling rule is triggered.
  \State Draw \(x_t^{(i)}\sim q_t(\cdot\mid x_{t-1}^{a(i)},y_t)\).
  \State Compute \(\alpha_t^{(i)}\) by \eqref{eq:bf-hd-pf-weight}.
  \State Normalize weights and record ESS, maximum weight, and unique ancestors.
\EndFor
\end{algorithmic}
\end{algorithm}

The effective sample size
\begin{equation}
  \mathrm{ESS}_t=\frac{1}{\sum_i(\bar w_t^{(i)})^2}
  \label{eq:bf-hd-ess}
\end{equation}
measures concentration of normalized weights.  It does not prove posterior
accuracy.  In the running cell, particles may all land near \(+\sqrt y\) and
still have high ESS within that lobe; the missing \(-\sqrt y\) mode is then a
coverage failure, not a normalization failure.

\section{Weight Collapse And The HMC Boundary For Particles}
\label{sec:bf-hd-pf-collapse}

\begin{proposition}[Weight-collapse warning]
\label{prop:bf-hd-pf-collapse}
In the iid/high-dimensional settings analyzed by \citet{bengtsson2008curse}
and in the log-likelihood variance framing of \citet{snyder2008obstacles},
prior-proposal particle weights can concentrate so strongly that a small
number of particles dominate the normalized measure unless ensemble size grows
substantially with the effective observation dimension.
\end{proposition}

\begin{proof}[Mechanism derivation]
Write \(W_i=\exp(L_i)\), where \(L_i\) is a log likelihood contribution.  In a
block-observation model with approximately weakly dependent components,
\[
  L_i=\sum_{k=1}^{d_{\rm eff}}\ell_{ik},\qquad
  \Var(L_i)\approx d_{\rm eff}\sigma_\ell^2=:v_d.
\]
The following lognormal calculation is not a substitute for the cited
theorems, but it makes their warning visible:
\[
  \E[W_i]=\exp(\mu_d+v_d/2),\qquad
  \E[W_i^2]=\exp(2\mu_d+2v_d),
\]
so
\[
  \frac{\E[W_i^2]}{\E[W_i]^2}=\exp(v_d).
\]
Self-normalized importance sampling must fight this second-moment ratio.  If
\(v_d\) grows with effective dimension, fixed \(N\) cannot keep weights
balanced.
\end{proof}

Particle-filter gradients are a separate issue.  A standard resampling
particle filter does not give an ordinary smooth gradient for HMC.  With fixed
random numbers, no resampling, and fixed ancestry, one may differentiate a
branch-local algorithmic scalar.  With multinomial/systematic/residual
resampling, ancestor identities can change discontinuously with parameters.
Unbiased likelihood estimates support pseudo-marginal logic under the
assumptions of particle-MCMC theory, but that is not ordinary HMC on a smooth
potential.  Differentiable resampling belongs to the DPF implementation lane
and is not imported here as mathematical evidence.

HMC labels: bootstrap/SIR with ordinary resampling is
\texttt{HMC\_NOT\_ADMISSIBLE\_UNTIL\_SMOOTHED}; no-resampling fixed-random-number
diagnostics are \texttt{HMC\_BRANCH\_LOCAL\_ONLY}; particle likelihood estimators are
\texttt{DIAGNOSTIC\_ONLY} for this monograph unless a separate pseudo-marginal or
differentiable-particle contract is explicitly invoked.

\paragraph{Reader checkpoint.}
Object preserved: an empirical measure and possibly a likelihood estimate.
Approximation enters at: proposal choice, finite \(N\), resampling, and
finite-sample normalization.  Exported to synthesis: collapse diagnostics veto
runtime comparisons before speed or ESS-per-second is meaningful.

\section{Guided Proposals And Density-Ratio Correction}
\label{sec:bf-hd-guided-proposals}

A guided proposal tries to put samples where the current observation makes them
useful.  The price is that the proposal density must be known or corrected.  If
\(\gamma_t(x_t)\) is the unnormalized target density for the current step and
\(q_t(x_t)\) is the proposal density, importance sampling uses
\[
  w(x_t)=\frac{\gamma_t(x_t)}{q_t(x_t)}.
\]
The numerator might be
\[
  \gamma_t(x_t)
  =
  g_\theta(y_t\mid x_t)
  \int f_\theta(x_t\mid x_{t-1})\,\pi_{t-1}(dx_{t-1})
\]
or a particle approximation of this quantity.  The proposal \(q_t\) is not
the target.  It becomes useful only through the correction.

\begin{proposition}[Corrected proposal identity]
\label{prop:bf-hd-corrected-proposal}
If \(q(x)>0\) wherever the unnormalized target \(\gamma(x)>0\), then
\[
  \frac{\int \varphi(x)\gamma(x)\,dx}{\int\gamma(x)\,dx}
  =
  \frac{\E_q[\varphi(X)\gamma(X)/q(X)]}
       {\E_q[\gamma(X)/q(X)]},
\]
whenever the displayed expectations exist.
\end{proposition}

\begin{proof}
Substitute \(q(x)\{\gamma(x)/q(x)\}\) for \(\gamma(x)\) in both numerator and
denominator.  The support condition prevents division by zero on target mass.
\end{proof}

In the running cell, a proposal that samples only positive \(x\) values after a
large positive observation may look efficient if the realized data favor the
positive lobe.  It cannot represent the symmetric negative lobe unless the
target has genuinely ruled it out or another component covers it.  This is why
support is a mathematical condition, not a tuning preference.

\section{Transport Maps As Change-Of-Variables Objects}
\label{sec:bf-hd-transport-change}

Let \(z\sim r\) and let \(x=T_\phi(z,\theta)\) be a differentiable invertible
map.  The induced proposal density is
\begin{equation}
  q_\theta(x)
  =
  r(z)\left|\det D_zT_\phi(z,\theta)\right|^{-1},
  \qquad z=T_\phi^{-1}(x,\theta).
  \label{eq:bf-hd-map-density}
\end{equation}
The log correction against an unnormalized target \(\gamma_\theta\) is
\begin{equation}
  \log w(x,\theta)=\log\gamma_\theta(x)-\log q_\theta(x).
  \label{eq:bf-hd-transport-log-weight}
\end{equation}
This is the same change-of-variables logic used in transport-map MCMC and
normalizing flows.  The inspected P1R anchor for \citet{parno2018transport} is
the lower-triangular map proposal and adaptive proposal construction in its
Sections 2--3 and Algorithm 1.  The normalizing-flow survey
\citet{papamakarios2021flows} is context-only here; it is not used for a
filtering theorem.

\begin{proposition}[Transport proposal correction]
\label{prop:bf-hd-transport-correction}
If \(T_\phi\) is differentiable and invertible on the reference support and
\(q_\theta>0\) wherever \(\gamma_\theta>0\), then importance weights
\(\gamma_\theta(x)/q_\theta(x)\) correct samples from the transported proposal
for expectations under \(\gamma_\theta/\int\gamma_\theta\).
\end{proposition}

\begin{proof}
Equation \eqref{eq:bf-hd-map-density} is the change-of-variables formula.
Applying Proposition~\ref{prop:bf-hd-corrected-proposal} with this \(q_\theta\)
gives the corrected expectation identity.
\end{proof}

For transformed HMC in a latent coordinate \(z\), the potential must include
the Jacobian:
\begin{equation}
  U_z(z)
  =
  -\log \pi_q(T_\phi(z))
  -\log|\det D_zT_\phi(z)|.
  \label{eq:bf-hd-transport-hmc-potential}
\end{equation}
The gradient is the derivative of this scalar, including the derivative of the
log determinant and the pullback through \(T_\phi\).  A map-training loss is
not an HMC target.  HMC label: transport-preconditioned HMC is
\texttt{HMC\_ADMISSIBLE\_APPROXIMATE\_TARGET} only when the corrected scalar and
Jacobian are finite and differentiable on the sampler path; otherwise support
or Jacobian failure is a veto.

\paragraph{Reader checkpoint.}
Object preserved: corrected target expectations or a Jacobian-corrected
transformed scalar.  Approximation enters at: map fit, finite ensemble or
training data, support coverage, and determinant stability.  Exported to
synthesis: transport can reduce geometry only after support, correction, and
Jacobian checks pass.

\section{Triangular And Ensemble Transport Filters}
\label{sec:bf-hd-ensemble-transport}

\citet{rosenblatt1952remarks} gives a triangular conditional-CDF
transformation from an absolutely continuous multivariate distribution to
independent uniforms.  Modern filtering papers often describe monotone
triangular maps as Knothe--Rosenblatt maps; because the Knothe original remains
outside the checked support here, this chapter uses Rosenblatt and modern
transport-filtering sources rather than relying on an uninspected original.

The checked filtering anchor is \citet{spantini2022coupling}, which formulates
stochastic and deterministic map filters for nonlinear ensemble filtering and
develops localization for high dimension; the inspected P1R anchors are
Algorithms 1--3, Remarks 1--7, and localization Appendix B.  \citet{ramgraber2023ensemble}
is cited only for its smoothing-side triangular transport recursions in
Sections 2--3, Proposition 1, and Appendix C algorithms; it is not used as a
foundational filtering theorem.

\begin{algorithm}
\caption{Triangular-map ensemble filtering step}
\label{alg:bf-hd-map-filter}
\begin{algorithmic}[1]
\Require Forecast ensemble, observation \(y_t\), triangular map class,
  localization graph or block policy.
\State Fit a map or coupling using the checked source objective.
\State Apply the stochastic or deterministic analysis update.
\State If localized, solve local map problems and recombine by the declared
  graph policy.
\State Record monotonicity/Jacobian failures, map residual, localization
  sensitivity, ensemble spread, and reference metrics where available.
\end{algorithmic}
\end{algorithm}

A triangular map is attractive for the running cell because it can in principle
send a simple reference law into a two-lobed or skewed distribution.  It can
also fail in a way that is hard to see if the text only says "transport": the
chosen triangular ordering may make the conditional map hard to fit, the
ensemble may be too small to estimate the map, and localization may remove an
important dependence.  The industrial question is therefore not whether a map
can be written down; it is whether the fitted map has a source-supported
analysis contract and passes support/Jacobian/localization checks.

\section{Tensor-Train Density And PDE Filters}
\label{sec:bf-hd-tt-pde}

A tensor train is a compressed representation of a multidimensional array or
discretized function.  For a grid tensor \(A(i_1,\ldots,i_d)\), the TT format
of \citet{oseledets2011tt} writes
\begin{equation}
  A(i_1,\ldots,i_d)=G_1(i_1)G_2(i_2)\cdots G_d(i_d),
  \label{eq:bf-hd-tt-format}
\end{equation}
where the \(G_k(i_k)\) are small matrices with connecting ranks.  If the mode
size is \(M\) and ranks are bounded by \(r\), storage is \(O(dMr^2)\) instead
of \(O(M^d)\).  The condition is rank stability: if ranks grow with the
effective dimension, the compression disappears.  The TT-SVD and rounding
results in \citet{oseledets2011tt} support this storage and approximation
framework.  \citet{oseledets2010ttcross} is the checked anchor for TT-cross
interpolation/integration of black-box tensors under low-rank assumptions.

For filtering, the tensor may represent a density, a Fokker--Planck or DMZ
operator, a likelihood multiplier, or a conditional transport map.  Those are
different semantics.  A small TT residual for array entries does not by itself
prove positivity, mass conservation, boundary correctness, or likelihood
accuracy.

\citet{li2019ttfiltering} is used as a direct source for TT/QTT nonlinear
filtering through the pathwise robust DMZ/FKE route; the checked anchors are
Propositions 4.1 and 4.3, Algorithms 1--2, Lemmas 5.3--5.6, and Theorem 5.7.
They support source-local offline operator compression and online assimilation
under the paper's assumptions, not general rank control.  \citet{fox2021grid}
supports the Bayes-optimal continuous-discrete grid-filtering framing through
Fokker--Planck/continuity equations (1)--(7), finite-volume propagation, Bayes
update, and functional TT Section 4.  \citet{meng2026correlated} is a recent
arXiv source for correlated-noise DMZ/Zakai SPDE and TT approximation
conditions, especially model equations (1)--(2), Algorithm 1, and Theorems
4.3--4.6 and 5.7; it is not used here as mature validation.

\begin{algorithm}
\caption{TT/PR-DMZ filtering pilot}
\label{alg:bf-hd-tt-pde}
\begin{algorithmic}[1]
\Require Checked PDE/filtering assumptions, tensor grid or functional basis,
  rank cap, rounding tolerance, low-dimensional reference.
\State Form the PR-DMZ/FKE/DMZ operator specified by the cited source.
\State Represent initial density and operators in TT/QTT/FTT format.
\For{each observation interval}
  \State Apply the cited offline/online or time-stepping update.
  \State Round/truncate ranks with declared tolerance and branch label.
  \State Renormalize only if the density semantics require it.
  \State Record ranks, mass error, positivity error, boundary residual,
    discretization error, and reference moment/density error.
\EndFor
\State Keep as diagnostic until likelihood and posterior checks pass.
\end{algorithmic}
\end{algorithm}

For the running cell, a TT representation is unnecessary in one dimension; it
is pedagogical.  It shows that the object being represented is a density over a
grid or basis.  In a high-dimensional macro-finance block, TT becomes plausible
only if economic block structure keeps the unfolding ranks small and if
probability semantics survive rounding.

HMC label: the TT/PDE material is not operational HMC support in this lane.
It remains \texttt{DIAGNOSTIC\_ONLY} until a concrete implementation defines a
scalar
\[
  \widehat\ell_T(\theta)=\sum_t\log\widehat Z_t(\theta)
\]
with a differentiable branch and a gradient that is the derivative of that
same scalar.  Frozen ranks and pivots are at most
\texttt{HMC\_BRANCH\_LOCAL\_ONLY}; rank adaptation, TT-cross pivot changes,
clipping, hard truncation, and failed normalization are
\texttt{HMC\_NOT\_ADMISSIBLE\_UNTIL\_SMOOTHED} for ordinary HMC.  TT compression
residuals do not by themselves control likelihood-gradient error.

\section{TT Sequential Learning And Conditional KR Transports}
\label{sec:bf-hd-zhao-cui}

\citet{zhao2024ttsequential} is the central discrete-time TT source for this
block.  It treats sequential state and parameter learning in state-space models
by recursively approximating posterior densities with tensor trains.  Its
checked technical role here is direct: SSM equations (1)--(9), Algorithms
1--5, Proposition 2, Proposition 4, Theorems 7--8, Corollary 12, and the
conditional Knothe--Rosenblatt maps built from the TT approximation.

The bridge is as follows.  A nonnegative TT approximation supplies conditional
density factors or conditional cumulative distribution functions.  A
triangular conditional map can then be constructed from one-dimensional
conditional CDFs, which is the KR transport idea in operational form.  For
BayesFilter this is attractive because it connects low-rank density
approximation to transport preconditioning.  It remains conditional because
normalization, monotonicity, rank error, and Jacobian stability are part of the
method, not afterthoughts.

\citet{cui2021deep} is used through the inspected anchors recorded in P1R:
functional TT Section 2.3, Propositions 1--7, Theorems 1, 3, 4, and Algorithms
1--4 for deep inverse Rosenblatt transports.  Its role here is to support the
existence of a TT/KR transport construction path, not to prove that BayesFilter
has a validated transport-HMC backend.

\begin{longtable}{@{}p{0.22\linewidth}p{0.33\linewidth}p{0.33\linewidth}@{}}
\toprule
TT/KR object & What must be checked & HMC implication \\
\midrule
\endhead
Squared-TT density & mass, nonnegativity construction, rank error & can define
a scalar only after \(\widehat Z_t\) is declared. \\
Conditional CDF/map & monotonicity, inverse, boundary behavior & transport is
valid only with finite Jacobian and support coverage. \\
Frozen branch & fixed ranks, pivots, active grids & branch-local derivative
only. \\
Adaptive branch & changing ranks, pivots, truncation & not ordinary HMC until
smoothed or moved outside the sampler. \\
\bottomrule
\end{longtable}

\section{Tensor-Network Kalman And Square-Root Caution}
\label{sec:bf-hd-tn-kalman}

\citet{batselier2016tnkf} rewrites Kalman recursions for exponentially large
lifted Volterra-system state vectors using tensor-network cores.  This is a
linear-Gaussian recursion over a lifted representation, not a general
nonlinear filter over the original state.  \citet{menzen2024tnsrkf} is a
recent preprint warning that tensor-network covariance rounding can destroy
positive definiteness and lead to divergence, motivating a tensor-network
square-root Kalman filter.  The robust conclusion is algebraic: covariance
compression needs a PSD or factor gate.

\begin{proposition}[Covariance compression needs a factor gate]
\label{prop:bf-hd-factor-gate}
Any tensor-network compression of a covariance-like object used by a Gaussian
update must expose either a positive-semidefinite reconstruction check or a
factor/square-root representation whose product is positive semidefinite by
construction.
\end{proposition}

\begin{proof}
The Gaussian update uses \(P\) and \(S\) as covariance matrices.  If rounding
returns an indefinite approximation, the innovation solve and posterior
covariance no longer correspond to a Gaussian law.  A square-root factor
\(P=LL^\top\) preserves positive semidefiniteness algebraically up to factor
rounding and reconstruction checks.
\end{proof}

The defect appears in a two-by-two example:
\[
  P=\begin{pmatrix}1&1\\1&1\end{pmatrix},\qquad
  \widehat P_\epsilon=
  \begin{pmatrix}1&1+\epsilon\\1+\epsilon&1\end{pmatrix}.
\]
The eigenvectors \((1,1)^\top\) and \((1,-1)^\top\) have eigenvalues
\(2+\epsilon\) and \(-\epsilon\).  Thus an arbitrarily small unconstrained
rounding perturbation can leave the PSD cone.  A tensor-network covariance
filter that does not expose a PSD or factor gate is not merely approximate; it
may stop representing a covariance.

HMC label: TN covariance filters are generally \texttt{DIAGNOSTIC\_ONLY} in this lane.
If used inside a Gaussian approximate likelihood, they inherit the Gaussian
innovation score from Chapter~\ref{ch:bf-highdim-gaussian-quadrature} only when
the covariance factor branch and scalar are fixed and differentiable.

\section{Complexity And Diagnostics}
\label{sec:bf-hd-ptt-complexity}

\begin{longtable}{@{}p{0.18\linewidth}p{0.24\linewidth}p{0.31\linewidth}p{0.17\linewidth}@{}}
\toprule
Family & Scaling object & Diagnostics & Industrial role \\
\midrule
\endhead
Bootstrap/SIR PF & \(O(N)\) model calls per step, \(O(Nn)\) memory & ESS, max
weight, unique ancestors, likelihood variance, mode coverage & baseline and
collapse warning. \\
Guided/flow PF & proposal/map cost plus corrected weights & finite
density-ratio, support mismatch, corrected ESS & structured proposal research. \\
Triangular map filter & ensemble size times map-estimation cost; localization
dependent & monotonicity, map residual, localization sensitivity & promising
when block/local structure is real. \\
TT/QTT PDE filter & \(O(dMr^2)\) storage plus rank-dependent operator cost &
rank growth, mass, positivity, boundary, discretization & direct density
research path. \\
TT/KR transport & TT density plus conditional-CDF/map cost & normalization,
monotonicity, logdet, propagation error & TT--transport bridge. \\
TN Kalman/square-root & tensor-network rank dependent & PSD/factor residual,
rounding drift & lifted Gaussian compression. \\
\bottomrule
\end{longtable}

\section{Defects Passed To Synthesis}
\label{sec:bf-hd-ptt-defects}

Particles, transports, and tensor methods pass five defects to the synthesis
chapter.

\begin{enumerate}[label=(\arabic*)]
  \item Particle methods expose collapse through log-weight variance, ESS,
  maximum weight, unique ancestors, and mode-coverage diagnostics.  Runtime
  comparisons are meaningless after collapse.
  \item Guided and transported proposals must retain the density-ratio
  correction or an explicitly source-supported deterministic analysis
  contract.  Better geometry without correction is bias.
  \item Transport maps require support, monotonicity or invertibility, and
  finite Jacobian diagnostics before smoothing or MCMC benefits can be
  interpreted.
  \item TT density filters require rank, mass, positivity, boundary, and
  likelihood/score checks.  A small storage footprint is not a probability
  statement.
  \item Tensor-network covariance filters require PSD or square-root/factor
  gates.  Entrywise or corewise rounding is not closed under the PSD cone.
\end{enumerate}

\paragraph{Exported to synthesis.}
\begingroup
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{longtable}{@{}p{0.10\linewidth}p{0.13\linewidth}p{0.12\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.08\linewidth}p{0.11\linewidth}p{0.12\linewidth}@{}}
\toprule
Object & Exact target & Approximation site & Failure mode & Diagnostic/veto &
Cost variable & Promotion gate & Anchor \\
\midrule
\endhead
Particle measure & filtering law and normalizer estimate & finite \(N\),
proposal, resampling & collapse or lost ancestors & ESS, max weight, unique
ancestors & \(Nd\) and model calls & collapse absent or corrected &
\citet{gordon1993novel,arulampalam2002tutorial,bengtsson2008curse,snyder2008obstacles} \\
Corrected proposal & target expectations under correction & proposal/map fit &
support or density-ratio failure & finite correction, ESS, support check & map
or proposal cost & correction auditable &
Prop.~\ref{prop:bf-hd-corrected-proposal} \\
Transport map/filter & Jacobian-corrected law or analysis contract & fitted map
and localization & wrong coupling or lost dependence & map residual, logdet,
localization sensitivity & ensemble/map dimension & source contract satisfied &
\citet{rosenblatt1952remarks,spantini2022coupling} \\
TT density/PDE & filtering density or operator & discretization, rank,
rounding & rank spike, negative density, mass drift & rank, mass, positivity,
boundary & TT rank \(r\) & semantic checks stable &
\citet{oseledets2011tt,li2019ttfiltering,fox2021grid,zhao2024ttsequential} \\
TN covariance & PSD covariance/factor & tensor-network rounding & indefinite
covariance & PSD/factor residual & TN rank & square-root/factor gate &
Prop.~\ref{prop:bf-hd-factor-gate} \\
\bottomrule
\end{longtable}
\endgroup

\section{Methodological Boundary And Sources}
\label{sec:bf-hd-ptt-boundary}

The substantive claims in this chapter are tied to particle, transport, TT, and
tensor-network sources where their checked constructions are used.  Savostyanov
maxvol theory and Knothe's original rearrangement source were not inspected
directly here; the chapter therefore uses Oseledets--Tyrtyshnikov for TT-cross
and Rosenblatt plus modern transport-filtering papers for triangular maps.
DPF implementation evidence and student baselines are not used as mathematical
authority.  No source here is used to claim production readiness, NAWM
readiness, posterior accuracy, tensor-method validation, transport-method
validation, HMC convergence, GPU/XLA readiness, or default readiness.

Exported forward: Chapter~\ref{ch:bf-highdim-hmc-research} may import only the
declared scalar/HMC consequences of the particle, transport, TT, and TN objects
named here, and Chapter~\ref{ch:bf-highdim-candidate-synthesis} may import only
the defects, diagnostics, promotion gates, and source-risk boundaries stated
here.  Neither later chapter should treat TT/KR transport existence,
branch-local differentiability, or covariance-compression feasibility as stronger
than these checked sources support.
