\chapter[HMC Research Program For Nonlinear SSMs]{HMC as a Research Program for Nonlinear State-Space Models}
\label{ch:bf-highdim-hmc-research}

\section{The Target Comes Before The Sampler}
\label{sec:bf-hd-hmc-target}

Hamiltonian Monte Carlo is only as meaningful as the scalar potential it is
given.  For an unconstrained coordinate $q\in\R^d$ and transform
$\theta=\tau(q)$, the potential associated with the exact likelihood is
\begin{equation}
  U(q)
  =
  -\ell_T(\tau(q))-\log \pi(\tau(q))
  -\log\left|\det D\tau(q)\right|.
  \label{eq:bf-hd-hmc-potential}
\end{equation}
If the nonlinear filter supplies $\widehat\ell_T$, HMC samples the posterior
defined by $\widehat\ell_T$ unless an exact correction is supplied.  The
gradient must be the gradient of the same scalar:
\begin{equation}
  \nabla_q U(q)
  =
  -D\tau(q)^\top
  \left[
  \nabla_\theta\ell_T(\theta)+\nabla_\theta\log\pi(\theta)
  \right]_{\theta=\tau(q)}
  -\nabla_q\log|\det D\tau(q)|.
  \label{eq:bf-hd-hmc-gradient}
\end{equation}

The central object is therefore a scalar function \(U(q)\), not a filtering
distribution.  The input is a likelihood or approximate likelihood from a
filter, a prior, and a parameter transform.  The output is a Markov transition
that is meaningful only for the density proportional to \(\exp\{-U(q)\}\).  The
exact target uses \(\ell_T\) from
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}; the approximate target uses
the declared \(\widehat\ell_T\).  The diagnostic role is to check finite
values, same-scalar gradients, support, Jacobians, divergences, and energy
behavior before any efficiency comparison.

\paragraph{Running quadratic-observation cell.}
In the scalar cell \(x_t=\rho x_{t-1}+\eta_t,\ y_t=x_t^2+\epsilon_t\), HMC
does not rescue an unnamed filtering approximation.  For parameter inference it
uses the scalar likelihood
\[
  \ell_T(\theta)=\sum_{t=1}^T\log Z_t(\theta),
\]
or a declared approximation \(\widehat\ell_T(\theta)\).  The two-lobed state
update matters because it makes \(Z_t\) and its derivative hard to compute;
HMC begins only after the scalar and its gradient have been named.

\small
\begin{longtable}{@{}p{0.24\linewidth}p{0.31\linewidth}p{0.33\linewidth}@{}}
\toprule
Gradient object & Scalar it differentiates & What a check can and cannot say \\
\midrule
\endhead
Exact sensitivity gradient & \(\ell_T=\sum_t\log Z_t\) & Supports the exact
parameter target under the assumptions of
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}. \\
Autodiff through approximate filter & \(\widehat\ell_T\) actually executed by
the algorithm & Supports the declared approximate target, not the exact one. \\
Finite-difference parity & reported \(U(q)\) at nearby points & Vetoes
same-scalar mismatch; does not prove convergence or posterior accuracy. \\
Transport/transform gradient & Jacobian-corrected transformed scalar & Valid
only with support and determinant terms included. \\
\bottomrule
\end{longtable}
\normalsize

\section{Analytical HMC Gradient From The Filtering Likelihood}
\label{sec:bf-hd-analytical-hmc-gradient}

The analytical HMC gradient is a chain-rule consequence of the scalar target.
Let
\[
  F(\theta)=\ell_T(\theta)+\log\pi(\theta),
  \qquad \theta=\tau(q),
\]
where \(\tau:\R^d\to\Theta\) is differentiable on the interior of the
unconstrained parameter space and \(D\tau(q)\) is its Jacobian.  From
\eqref{eq:bf-hd-hmc-potential},
\[
  U(q)=-F(\tau(q))-\log|\det D\tau(q)|.
\]
For any coordinate \(q_j\), the ordinary chain rule gives
\[
  \frac{\partial U}{\partial q_j}
  =
  -\sum_k
  \frac{\partial \tau_k}{\partial q_j}
  \frac{\partial F}{\partial \theta_k}(\tau(q))
  -
  \frac{\partial}{\partial q_j}\log|\det D\tau(q)|.
\]
Stacking the coordinate derivatives gives
\begin{equation}
  \nabla_q U(q)
  =
  -D\tau(q)^\top
  \left[
  \nabla_\theta\ell_T(\theta)+\nabla_\theta\log\pi(\theta)
  \right]_{\theta=\tau(q)}
  -
  \nabla_q\log|\det D\tau(q)|.
  \label{eq:bf-hd-analytical-hmc-gradient}
\end{equation}
The likelihood gradient in \eqref{eq:bf-hd-analytical-hmc-gradient} is the
filtering sensitivity recursion
\eqref{eq:bf-hd-prediction-sensitivity}--\eqref{eq:bf-hd-likelihood-gradient}.
Thus the HMC gradient is analytical only after the filter has supplied the
analytical or explicitly approximated derivative of the same scalar likelihood
reported to the sampler.

\begin{proposition}[HMC gradient contract for a state-space likelihood]
\label{prop:bf-hd-hmc-gradient-contract}
For the dominated discrete-time state-space model in
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}, an HMC implementation for
parameters \(\theta=\tau(q)\) targets the exact posterior only if its potential
uses \eqref{eq:bf-hd-hmc-potential} and its gradient is
\eqref{eq:bf-hd-analytical-hmc-gradient} with
\(\nabla_\theta\ell_T\) computed from the exact likelihood normalizers.
If the filter supplies only an approximate scalar \(\widehat\ell_T\), then
ordinary HMC can at most target the declared approximate posterior defined by
that scalar.  Recovering the exact posterior from an approximate or randomized
likelihood would require a separately defined exact-target extended-state or
Metropolis correction scheme; such a scheme is outside the scope of this
chapter.
\end{proposition}

Plain-English claim: HMC can only be trusted for the posterior whose scalar
potential it actually evaluates.  Object being approximated or preserved:
the posterior density in the unconstrained coordinate \(q\).  Assumptions:
the transform is differentiable with valid support/Jacobian, the likelihood
normalizers are finite and positive, and the likelihood gradient is the one
derived from those normalizers.  Approximation step: replacing
\(\ell_T\) by \(\widehat\ell_T\) changes the target unless a separately defined
exact-target correction scheme is introduced.  Failure if assumptions break: HMC
may still run and report ESS, but the chain no longer targets the stated
posterior.  Industrial interpretation: same-scalar parity is a promotion gate
before speed or geometry improvements matter.

\begin{proof}
The posterior density in \(\theta\) coordinates is proportional to
\(\exp\{\ell_T(\theta)\}\pi(\theta)\).  The change of variables
\(\theta=\tau(q)\) gives the \(q\)-density proportional to
\(\exp\{\ell_T(\tau(q))\}\pi(\tau(q))|\det D\tau(q)|\), whose negative log
density is \eqref{eq:bf-hd-hmc-potential} up to an additive constant.  The
gradient is the chain rule in \eqref{eq:bf-hd-analytical-hmc-gradient}.  The
identity \(\nabla_\theta\ell_T=\sum_t\nabla_\theta Z_t/Z_t\) and the
sensitivity recursions are derived in
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}.  If a different scalar or a
different gradient is used, the Hamiltonian is no longer the exact posterior
Hamiltonian.
\end{proof}

\begin{remark}[Scalar Gaussian likelihood gradient]
\label{ex:bf-hd-scalar-gaussian-gradient}
Let \(x_t\sim\calN(\theta,\sigma_x^2)\) independently and
\(y_t\mid x_t\sim\calN(x_t,\sigma_y^2)\), with known variances.  Integrating
out \(x_t\) gives \(y_t\sim\calN(\theta,S)\), \(S=\sigma_x^2+\sigma_y^2\), so
\[
  \ell_T(\theta)
  =
  -\frac12\sum_{t=1}^T
  \left\{\log(2\pi S)+\frac{(y_t-\theta)^2}{S}\right\}.
\]
The analytical likelihood gradient is
\[
  \nabla_\theta\ell_T(\theta)
  =
  \sum_{t=1}^T\frac{y_t-\theta}{S}.
\]
With a Gaussian prior \(\theta\sim\calN(\mu_0,\sigma_0^2)\) and identity
transform \(\tau(q)=q\), the HMC potential is
\[
  U(q)=-\ell_T(q)+\frac12\frac{(q-\mu_0)^2}{\sigma_0^2}+C,
  \qquad
  \nabla_q U(q)
  =
  -\sum_{t=1}^T\frac{y_t-q}{S}
  +\frac{q-\mu_0}{\sigma_0^2}.
  \label{eq:bf-hd-scalar-gaussian-hmc-gradient}
\]
This toy example is not a nonlinear-filter validation.  Its role is to show
the scalar contract: the value and gradient both come from the same integrated
likelihood and prior.
\end{remark}

\section{Approximate-Filter Gradients}
\label{sec:bf-hd-approx-filter-gradients}

If a filter reports
\[
  \widehat\ell_T(\theta)=\sum_{t=1}^T\log\widehat Z_t(\theta),
\]
then an HMC chain using
\[
  \widehat U(q)
  =
  -\widehat\ell_T(\tau(q))-\log\pi(\tau(q))
  -\log|\det D\tau(q)|
  \label{eq:bf-hd-approx-hmc-potential}
\]
targets the approximate posterior defined by \(\widehat\ell_T\), not the exact
posterior in general.  Its same-scalar gradient is
\[
  \nabla_q \widehat U(q)
  =
  -D\tau(q)^\top
  \left[
  \nabla_\theta\widehat\ell_T(\theta)+
  \nabla_\theta\log\pi(\theta)
  \right]_{\theta=\tau(q)}
  -
  \nabla_q\log|\det D\tau(q)|.
  \label{eq:bf-hd-approx-hmc-gradient}
\]
Automatic differentiation through an approximate filter can be valuable, but it
differentiates the algorithmic scalar it actually executed.  If tensor
rounding, resampling, clipping, adaptive sparse-grid stopping, rank caps,
localization gates, or invalid-state penalties enter the calculation, the
reported gradient is defensible only after the reported scalar has been named.
Finite differences check scalar parity; they do not prove exactness.  An
approximate-filter gradient cannot be described as an exact posterior gradient
or exact state-space likelihood gradient unless a separately specified
exact-target correction scheme has been defined and justified.

Chapter~\ref{ch:bf-highdim-gaussian-quadrature} now gives the concrete
Gaussian/quadrature case: the scalar is the approximate Gaussian innovation
likelihood in \eqref{eq:p31-fsgq-loglik}, and the score is
\eqref{eq:p31-score}.  That formula covers fixed EKF/UKF/CKF,
high-degree cubature, tensor-product Gauss--Hermite, and fixed sparse-grid
branches only for the approximate scalar they report.  Chapter~\ref{ch:bf-highdim-particle-transport-tensor}
then records why ordinary resampling particles, adaptive TT ranks, TT-cross
pivot changes, and adaptive sparse-grid point changes are not smooth HMC
gradients unless the target is smoothed or the branch is frozen and labeled.

\paragraph{Reader checkpoint.}
What should now be clear: HMC is downstream of the filtering likelihood, not a
replacement for it.  Object preserved: the scalar potential and the density it
defines.  Approximation enters at: \(\widehat Z_t\), autodiff through the
filter, transforms, and nondifferentiable gates.  Exported to synthesis:
performance comparisons are admissible only after value-gradient parity,
finite support/Jacobian checks, and baseline diagnostics pass.

\begin{proposition}[Same-scalar condition]
\label{prop:bf-hd-same-scalar}
A Metropolis-corrected HMC transition targets the density proportional to
$\exp\{-U(q)\}$ only if the Hamiltonian evaluated in the accept/reject step is
the same Hamiltonian whose gradient is used by the numerical trajectory, up to
the declared integrator approximation.
\end{proposition}

\begin{proof}
The HMC argument in \citet{neal2011mcmc} relies on a reversible,
volume-preserving numerical map followed by a Metropolis correction using the
Hamiltonian for which that map is a discretized flow.  If the gradient comes
from a different scalar, the proposal is not a reversible discretization of the
accepted Hamiltonian, so the usual detailed-balance argument no longer applies.
\end{proof}

The diagnostic version is a scalar parity test.  For a coordinate direction
\(e_j\), define
\[
  \Delta_j(q;h)=
  \frac{U_{\rm reported}(q+h e_j)-U_{\rm reported}(q-h e_j)}{2h}
  -[\nabla U_{\rm used}(q)]_j .
\]
If \(\max_j|\Delta_j(q;h)|\) is not small on typical states over a stable range
of \(h\), the implementation is not using the derivative of the scalar it
reports.  Passing this test does not prove HMC convergence; failing it vetoes
the target contract.

\section{Plain HMC, NUTS, And Diagnostics}
\label{sec:bf-hd-hmc-diagnostics}

\citet{neal2011mcmc} is the checked source for the Hamiltonian mechanics,
leapfrog, and Metropolis correction.  \citet{hoffman2014nuts} supplies the NUTS
reference algorithm and adaptation logic.  \citet{betancourt2017conceptual}
supplies the diagnostic geometry vocabulary: divergences, energy behavior, and
bad curvature are not cosmetic warnings.

\begin{algorithm}
\caption{Fixed-trajectory HMC diagnostic kernel}
\label{alg:bf-hd-fixed-hmc}
\begin{algorithmic}[1]
\Require Potential $U$, gradient $\nabla U$, mass matrix $M$, step size
  $\epsilon$, leapfrog count $L$, current $q$.
\State Check that $U(q)$ and $\nabla U(q)$ are finite.
\State Draw $p\sim\calN(0,M)$.
\State Apply $L$ leapfrog steps using the declared invalid-region policy.
\State Recompute the same scalar Hamiltonian at start and proposal.
\State Accept/reject by the Metropolis probability.
\State Record energy error, divergence flag, acceptance, finite status, and
  invalid-region hits.
\end{algorithmic}
\end{algorithm}

NUTS is a valuable reference and diagnostic sampler, but this monograph does
not promote TFP NUTS or any dynamic-control-flow sampler as a BayesFilter
production default for nonlinear SSMs.  That is an engineering boundary, not a
claim that NUTS is mathematically inferior.  The later HMC chapters,
Chapter~\ref{ch:bf-hmc-for-state-space} and
Chapter~\ref{ch:bf-transport-surrogates}, carry the broader sampler and
transport-policy discussion; the role here is only to state which high-dimensional
filter targets are admissible inputs to those later decisions.

\section{RMHMC And Metric Burden}
\label{sec:bf-hd-rmhmc}

\citet{girolami2011riemann} replaces a fixed mass matrix with a
position-dependent metric.  In a nonlinear SSM this is appealing because the
posterior curvature can rotate across the parameter space.  It is also costly:
one needs a positive-definite metric, derivatives of that metric, generalized
leapfrog solves for a nonseparable Hamiltonian, and stable filtering Hessian or
Fisher information approximations.  RMHMC is therefore a geometry-aware
competitor and a conceptual warning: a dense mass matrix can rescale axes, but
it cannot remove arbitrary nonlinear geometry.

\section{Transport-Map Accelerated MCMC}
\label{sec:bf-hd-transport-mcmc}

\citet{parno2018transport} constructs adaptive lower-triangular transport maps
to precondition Metropolis proposals while preserving an exact target through
Metropolis correction.  If $q=T_\phi(z)$, the transformed density in $z$ is
\begin{equation}
  \log \pi_z(z)=\log \pi_q(T_\phi(z))+\log|\det DT_\phi(z)|.
  \label{eq:bf-hd-transport-target}
\end{equation}
This is the same object that appears in normalizing-flow and NeuTra-style HMC.
The map can reduce posterior dependence, but if its tails or support are wrong
the transformed geometry can be worse than the original.

\section{NeuTra And Learned Transports}
\label{sec:bf-hd-neutra}

\citet{hoffman2019neutra} uses neural transports to "neutralize" bad geometry
before HMC.  The relevant claim for this monograph is modest: an invertible
learned map can be a preconditioner for HMC if the transformed target includes
the Jacobian term and if the sampler diagnostics improve.  It is not a
correctness substitute.  A trained density model or variational loss can suggest
a map; it cannot replace the Metropolis target contract.

\begin{proposition}[Transport preconditioning preserves target only with the
Jacobian]
\label{prop:bf-hd-jacobian-target}
Let $T$ be a differentiable bijection from $z$ to $q$ coordinates.  HMC in $z$
targets the original $q$-density $\pi_q$ only when the $z$-potential includes
$-\log\pi_q(T(z))-\log|\det DT(z)|$.
\end{proposition}

\begin{proof}
For any measurable set $A$ in $q$ space, the probability under the pushforward
of $\pi_z$ is $\int_{T^{-1}(A)}\pi_z(z)\,dz$.  Choosing
$\pi_z(z)=\pi_q(T(z))|\det DT(z)|$ and applying the change of variables
$q=T(z)$ gives $\int_A\pi_q(q)\,dq$.
\end{proof}

The sign of the Jacobian term is a common source of silent target changes.  If
the sampler evolves in \(z\) but reports inference in \(q=T(z)\), then
\[
  \pi_z(z)\,dz=\pi_q(q)\,dq
  =\pi_q(T(z))|\det DT(z)|\,dz.
\]
The potential is therefore
\[
  U_z(z)=-\log\pi_q(T(z))-\log|\det DT(z)|.
\]
Using the opposite sign samples a different density unless the determinant is
constant.

\section{TT/KR Transports As A Bridge}
\label{sec:bf-hd-tt-kr-hmc}

The deep inverse Rosenblatt transport construction of \citet{cui2021deep} and
the conditional KR maps in \citet{zhao2024ttsequential} suggest a bridge:
tensor-train density approximations can induce triangular transports, and those
transports can precondition sampling or sequential learning.  This is directly
aligned with the failure mode of plain mass-matrix HMC on nonlinear geometry.
It is also fragile.  The TT density must remain nonnegative or be squared by
construction, ranks must stay controlled, the map must be monotone, and the
transformed HMC target must use the Jacobian-corrected scalar.

\section{Promotion Ladder}
\label{sec:bf-hd-hmc-ladder}

\begin{algorithm}
\caption{Transport/HMC promotion ladder for nonlinear SSMs}
\label{alg:bf-hd-hmc-ladder}
\begin{algorithmic}[1]
\Require Filter target, baseline HMC kernel, candidate transport or metric,
  controlled reference problem.
\State Verify finite value and gradient for the same scalar target.
\State Run fixed HMC and record divergences, energy error, acceptance,
  $\widehat R$, ESS, and posterior/reference checks where meaningful.
\State Fit or construct transport/metric without changing the intended target.
\State Run transformed HMC with the Jacobian-corrected potential.
\State Reject on any persistent divergence, same-scalar mismatch, nonfinite
  Jacobian, or failed posterior/reference diagnostic.
\State Compare ESS per gradient or wall time only after validity gates pass.
\end{algorithmic}
\end{algorithm}

\section{Defects Passed To Synthesis}
\label{sec:bf-hd-hmc-defects}

The HMC research program passes five defects to the synthesis chapter.

\begin{enumerate}[label=(\arabic*)]
  \item The filter must define the scalar likelihood or approximate likelihood
  before any sampler can be interpreted.
  \item The gradient used by HMC must be the gradient of the same scalar used in
  the Metropolis correction and posterior interpretation.
  \item Transformations, including NeuTra and TT/KR-style maps, require the
  Jacobian-corrected target and valid support.
  \item Divergences, energy failures, invalid states, and posterior/reference
  mismatches are veto diagnostics.  ESS per gradient and wall time enter only
  after these vetoes pass.
  \item Geometry acceleration changes efficiency, not the target contract.  A
  transport or metric can make a valid target easier to explore; it cannot make
  an invalid or approximate target exact.
\end{enumerate}

The synthesis chapter therefore places HMC downstream of filtering,
normalization, scalar parity, and controlled-reference checks.

\paragraph{Exported to synthesis.}
\begingroup
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{longtable}{@{}p{0.10\linewidth}p{0.13\linewidth}p{0.12\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.08\linewidth}p{0.11\linewidth}p{0.12\linewidth}@{}}
\toprule
Object & Exact target & Approximation site & Failure mode & Diagnostic/veto &
Cost variable & Promotion gate & Citation/derivation anchor \\
\midrule
\endhead
Potential \(U(q)\) & transformed exact posterior & approximate likelihood,
prior/transform error & wrong target & finite scalar, support, Jacobian &
likelihood cost & scalar named & \eqref{eq:bf-hd-hmc-potential} \\
Gradient \(\nabla U\) & derivative of the same scalar & autodiff path,
sensitivity approximation & value-gradient mismatch & finite-difference
parity, nonfinite gradient & gradient-through-filter cost & same-scalar pass &
Proposition~\ref{prop:bf-hd-hmc-gradient-contract} \\
Transport HMC & Jacobian-corrected transformed target & map fit and determinant
& changed posterior geometry/target & determinant, inverse, support checks &
map/logdet cost & corrected scalar pass &
\citet{parno2018transport,hoffman2019neutra} \\
Sampler diagnostics & Markov transition for declared target & integrator,
adaptation, metric & divergences or invalid regions & energy, divergences,
\(\widehat R\), reference checks & gradients per effective draw & validity
before speed & \citet{neal2011mcmc,hoffman2014nuts,betancourt2017conceptual} \\
\bottomrule
\end{longtable}
\endgroup

\section{Implementation Boundary}
\label{sec:bf-hd-hmc-boundary}

BayesFilter has no nonlinear HMC convergence result in this lane, no NeuTra
backend, no RMHMC backend, no tensor/KR transport-preconditioned HMC backend,
and no NAWM posterior recovery benchmark.  The chapter defines a research
ladder.  It does not promote HMC, NUTS, RMHMC, NeuTra, XLA, or GPU execution as
production-ready for nonlinear state-space models.

This chapter imports the exact-target and likelihood-normalizer contract from
Chapter~\ref{ch:bf-highdim-nonlinear-foundations}, the approximate Gaussian
innovation scalar and score only from
Chapter~\ref{ch:bf-highdim-gaussian-quadrature}, and the particle/transport/TT/TN
object boundaries only from Chapter~\ref{ch:bf-highdim-particle-transport-tensor}.
It does not re-license those chapters' approximation families as exact targets,
validated transport backends, or production-ready HMC components.  Exported
forward, Chapter~\ref{ch:bf-highdim-candidate-synthesis} may consume only the
same-scalar, Jacobian, divergence, and promotion-ladder consequences stated
here, not a stronger convergence or deployment claim.
